### 2025-01-31 products/manuscript/Manuscript.qmd
Project Part 1

---
title: "Analysis of published transciptomic data studies to identify common genes and potential markers of bovine respiratory disease (BRD) and respiratory infection."
author: "Shaun van den Hurk"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

**Authors**

* First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678) 
* Second Author$^{3, *}$
* Third Autor$^{3, \dagger}$
* Last Author$^{1, 2, \land}$

**Author affiliations**

1. College of Public Health, University of Georgia, Athens, GA, USA.
2. Another Affiliation, Athens, GA, USA.
3. Yet another Affiliation.

$*$ These authors contributed equally to this work.

$\land$ Corresponding author: some@email.com

$\dagger$ Disclaimer: The opinions expressed in this article are the
author's own and don't reflect their employer.

{{< pagebreak >}}

# Summary/Abstract
_Write a summary of your project._
Respiratory disease is an important condition that affects the health, welfare and production of cattle. Respiratory disease in cattle is often attributed to the Bovine Respiratory Disease Complex (BRDC) and can be caused by a number of different pathogens. Currently there are over 9 different viruses and bacteria that are implicated in the BRDC, with the list growing, and potentiated by factors such as the environment and management factors.

The proposed project is a type of meta-analysis of different published datasets of transcriptomic data as a means to identify some potential biomarkers (such as miRNA) that might be differentially expressed with respiratory infection in bovines. The focus will be on the gene pathways that appear to be differentially expressed. The aim is to identify if there are any associations between infection (broadly and with specific pathogens) and these transcripts, which then could be used as a predictor of disease.

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
The data to be used is from published studies and online data found on the NCBI Gene Expression Omnibus (GEO) database. A search was performed on this database to look for data related to bovine respiratory disease as well as transcriptome, and/or protein and/or genes. There were between 10 than 22 results search results (depending on the search) on the database, and from these projects with published data, relevant projects are being identified. The studies are only included if they have retrievable data uploaded, either on the GEO database or the corresponding published article (including supplemental data).

Currently, five potential suitable sources have been identified, one such source is listed on GEO as "GSE199108", with the published article citation "O'Donoghue, S., Earley, B., Johnston, D., Mccabe, M.S., KIM, J., Taylor, J.F., Duffy, C., Lemon, K., Mcmenamy, M., Cosby, S.L. and Morris, D.W., Whole blood transcriptome analysis in dairy calves experimentally challenged with Bovine Herpesvirus 1 (BoHV-1) and comparison to a Bovine Respiratory Syncytial Virus (BRSV) challenge. Frontiers in Genetics, 14, p.171.".
From the mentioned resources, the data with the gene counts and other raw data will be retrieved and these Excel and Text files will provide the raw data. 
The researchers from these studies published the raw data and the scientific journal article, but not their analysis or their workflow for this. Additionally, each of these studies was performed with direct sampling, and I am not familiar with a study that has combined all of the data in a meta-analysis with collation of the data for a deeper analysis of these transcriptome profiles in cattle.

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

- Are there differences in the expression of genes and proteins between healthy and infected cattle that can be identified through transciptomic analysis? Can these genes be identied in this way.

- Are there any genes or biomarkers that are associated with infection with particular pathogens?

- Are there any common genes or biomarkers identified that are associated with infection with multiple different pathogens?

Thus the outcomes of interest are primarily the levels of expression of these targets and the disease status of the animals sampled.

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

The raw data has samples from healthy and experimentally infected animals which will be compared to get an indication of differential expression of certain genes between these groups. This will serve as the basis for the analysis as the focus will be on the subset of differentially expressed genes.
The raw data from the different sources will be cleaned up in R and analysed before they are combined in a single analysis to look for common features and to compare the expression profiles. Heatmaps will be helpful to help identify and visualise the differential expression of genes, and an approach might be to sample the top 10 genes with high and low levels of expression. 
Statistical analysis will focus largely on the associations between the factors, for example a t-test, and regression models might be considered too. We will look for associations between gene expression and infection status (heath vs infected).

It is expected that there will be some obvious differences in these levels of expression that will be asociated with the disease status and that many of these will be common between the different respiratory pathogens.

If potential genes or biomarkers are identified they will be run through online sources and databases (including tools such as the PubChem online tool) to identify the common names (or symbol) and proposed functions of these genes.

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2025-02-22 code/processing-code/processingfile-v1.qmd
Part 2 of project

---
title: "Processing CDC hypertension and CDV Mortality disease data from adults in the USA"
author: "Shaun van den Hurk"
date: "2025-02-20"
output: html_document
---

This file is a continuation of the data processing and wrangling steps. 
The first document highlighted the steps from importing the Excel document, the intial exploration and how the data was cleaned and reduced in R to remove the redundant columns (columns with single values).
Some of the other columns were renamed to have the name of the column reflect what these single value columns had contained.
The dataframe was then saved as an RDS file.

The initial Excel sheet was removed from the repository because it was almost 300MB and this was not suitable for uploading to the GitHUB repository.
Therefore, the code in that first document will not run because the file was removed. But the file remains to show the steps that were executed. If desired, the source Excel document could be downloaded from the CDC and steps run.

This document will load the RDS file and continue with some of the data processing, namely providing a couple of figures and summary tables of the dataset.
We likely do not need to do much further cleaning as that was done in the other steps.

# Setup

Load needed packages. make sure they are installed.

```{r}
library(readxl) #for loading Excel files (in case we want to include an additional dataset)
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(knitr) #for nice tables and data integration
```

# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

```{r}
# path to data
# note the use of the here() package and not absolute paths
cleaned_hypertension_adult_mortality_location <- here::here("data","raw-data","cleaned_hypertens_adult_mortality.rds")
cleaned_hypertens_mortailty_adult <- readRDS(cleaned_hypertension_adult_mortality_location)
```

# Check data

We check the data again.

```{r}
dplyr::glimpse(cleaned_hypertens_mortailty_adult)
summary(cleaned_hypertens_mortailty_adult)

```

# Cleaning
From our summary we can see that there are still a lot of NA values in our dataset. 
We want to look at these more closely to see if there is any valuable information in these rows or if we should just remove them.

```{r}
#Use filter function to search for rows with NA values in the mortalities column.
NA_Data <- cleaned_hypertens_mortailty_adult |> filter(is.na(mortalities))
#Checking these NA values
summary(NA_Data)
print(NA_Data)
View(NA_Data)

```

There do not seem to be any data values attached to these NA values. So we will remove them and create a new dataset without them.

```{r}
#Once again using the filter function to leave us the dataframe without NA values
hypertens_adult_filtered <- cleaned_hypertens_mortailty_adult |> filter(!is.na(mortalities))
summary(hypertens_adult_filtered)
skimr::skim(hypertens_adult_filtered)
print(hypertens_adult_filtered)
```

We want a copy of this skimr data for a table for the mauscript. We will make edits of this later in the process.

```{r}
skim_summary_hypertens_filtered<-skimr::skim(hypertens_adult_filtered)|> as.data.frame()
kable(skim_summary_hypertens_filtered)
saveRDS(skim_summary_hypertens_filtered, file = here::here("data","processed-data", "skim_summary_hypertens_filtered.rds"))
print(skim_summary_hypertens_filtered)

```

Our filtered data is now down to 572572 observations (from the initial 1.1 million). We have also removed many of the unnecessary columns (down to 14 from the initial 24).

Our summaries and the skimr function help to provide a good overview for further evaluation of the data. 
We can see that our dataset contains a combination of numeric and character data. 
The numeric data includes the actual incidence rates of the hypertension-related cardiovascular disease death rates, as well as the confidence intervals that were used, and geographic coordinates.
Our character data includes the race, sex, age group, location data and the year.

This character data will be used as predictors of the mortality data as we break down these different variables and look for potential patterns and associations. The primary outcome of interest is the mortality rates but the important potential predictors are the location, year, race/ethnicity, and sex. AGe groups only provide two broad categories, namely 35 to 65 years old and 65 years and older, which is a bit limited but will allow for comparisons between these two groups.

Our skimr data shows that the mean of the hypertension cardiovascular-diseases mortalities is around 203, with a standard deviation of 223. 
We want to get an idea on the distribution of this data and will use a histogram for this.

```{r histogram mortalities}
#Generate a histogram of the mortality rates
mortalities_histogram <- hist(hypertens_adult_filtered$mortalities)
saveRDS(mortalities_histogram, file = here::here("data","processed-data", "mortalities_histogram.rds"))
plot(mortalities_histogram <- readRDS(here::here("data","processed-data", "mortalities_histogram.rds"))
# Print the mortalities histogram
print(mortalities_histogram))
```

We are happy with this for now.

# Save data 
The cleaned data will be saved as a new RDS file which will go under the processed data.

```{r}
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(hypertens_adult_filtered, file = save_data_location)
```

### 2025-02-22 products/manuscript/Manuscript.qmd
Part 2 of project

@@
 ---
-title: "Analysis of published transciptomic data studies to identify common genes and potential markers of bovine respiratory disease (BRD) and respiratory infection."
+title: "Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States."
 author: "Shaun van den Hurk"
@@
 ---
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
 
@@
 
-1. College of Public Health, University of Georgia, Athens, GA, USA.
-2. Another Affiliation, Athens, GA, USA.
-3. Yet another Affiliation.
+1. College of Veterinary Medicine, University of Georgia, Athens, GA, USA.
 
@@
 # Summary/Abstract
-_Write a summary of your project._
-Respiratory disease is an important condition that affects the health, welfare and production of cattle.
-Respiratory disease in cattle is often attributed to the Bovine Respiratory Disease Complex (BRDC) and can be caused by a number of different pathogens.
-Currently there are over 9 different viruses and bacteria that are implicated in the BRDC, with the list growing, and potentiated by factors such as the environment and management factors.
-
-The proposed project is a type of meta-analysis of different published datasets of transcriptomic data as a means to identify some potential biomarkers (such as miRNA) that might be differentially expressed with respiratory infection in bovines.
-The focus will be on the gene pathways that appear to be differentially expressed.
-The aim is to identify if there are any associations between infection (broadly and with specific pathogens) and these transcripts, which then could be used as a predictor of disease.
+Lifestyle diseases such as cardiovascular disease make up the highest causes of death in developed nations such as the United States.
+This project focuses on hypertension-related cardiovascular disease and hopes to help provide a greater understandign of the dynamics and patterns that may exist with hypertension and cardiovascualr diseased related mortality in the United States.
+The project seeks to evaluate data from the CDC on the rates of hypertension-related cardiovascular disease mortality and the associated demographic factors for this collected data, including county, age group, race/ethnicity, and sex.
+The data represents information collected from 2000 to 2019.
+This will be used to see if there are any potential patterns or correlation between the rate of mortality and these other demographic factors within the data.
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
+
+Lifestyle diseases such as cardiovascular disease make up the highest causes of death in developed nations such as the United States.
+This data analysis hopes to evaluate the rates of hypertension-related cardiovascular disease mortality in adults in the United States and trends that might be associated with demographic and geographic factors.
 
 ## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
-The data to be used is from published studies and online data found on the NCBI Gene Expression Omnibus (GEO) database.
-A search was performed on this database to look for data related to bovine respiratory disease as well as transcriptome, and/or protein and/or genes.
-There were between 10 than 22 results search results (depending on the search) on the database, and from these projects with published data, relevant projects are being identified.
-The studies are only included if they have retrievable data uploaded, either on the GEO database or the corresponding published article (including supplemental data).
 
-Currently, five potential suitable sources have been identified, one such source is listed on GEO as "GSE199108", with the published article citation "O'Donoghue, S., Earley, B., Johnston, D., Mccabe, M.S., KIM, J., Taylor, J.F., Duffy, C., Lemon, K., Mcmenamy, M., Cosby, S.L.
-and Morris, D.W., Whole blood transcriptome analysis in dairy calves experimentally challenged with Bovine Herpesvirus 1 (BoHV-1) and comparison to a Bovine Respiratory Syncytial Virus (BRSV) challenge.
-Frontiers in Genetics, 14, p.171.".
-From the mentioned resources, the data with the gene counts and other raw data will be retrieved and these Excel and Text files will provide the raw data. 
-The researchers from these studies published the raw data and the scientific journal article, but not their analysis or their workflow for this.
-Additionally, each of these studies was performed with direct sampling, and I am not familiar with a study that has combined all of the data in a meta-analysis with collation of the data for a deeper analysis of these transcriptome profiles in cattle.
+The dataset was collected from the CDC website and the information was sourced from the National Vital Statistics System.
+The dataset contains rates and trends in the hypertension-related cardiovascular disease (CVD) death rates.
+The study period is from 2000 to 2019, with two main intervals displayed in the data 2000-2009, and 2010-2019.
+The data is further distributed by county, age group (36-64 years and 65 years and older), race/ethnicity, and sex.
 
 ## Questions/Hypotheses to be addressed
-_State the research questions you plan to answer with this analysis._
 
-- Are there differences in the expression of genes and proteins between healthy and infected cattle that can be identified through transciptomic analysis? Can these genes be identied in this way.
+- Hypertension-related cardiovascular disease mortalities are higher in select demographics (age groups, race/ethnicities and sex), and thus these demographics show correlated with mortalities rates.
 
-- Are there any genes or biomarkers that are associated with infection with particular pathogens?
-
-- Are there any common genes or biomarkers identified that are associated with infection with multiple different pathogens?
-
-Thus the outcomes of interest are primarily the levels of expression of these targets and the disease status of the animals sampled.
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
+- There are clear regional/geographic distributions associated with hypertension-related cardiovascular disease mortalities.
 
@@
 
-_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._
+## Data import and cleaning
 
-The raw data has samples from healthy and experimentally infected animals which will be compared to get an indication of differential expression of certain genes between these groups.
-This will serve as the basis for the analysis as the focus will be on the subset of differentially expressed genes.
-The raw data from the different sources will be cleaned up in R and analysed before they are combined in a single analysis to look for common features and to compare the expression profiles.
-Heatmaps will be helpful to help identify and visualise the differential expression of genes, and an approach might be to sample the top 10 genes with high and low levels of expression.
-Statistical analysis will focus largely on the associations between the factors, for example a t-test, and regression models might be considered too.
-We will look for associations between gene expression and infection status (heath vs infected).
+This data was downloaded from the CDC online database at the following url: https://data.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Hypertension-related-Cardiovas/uc9k-vc2j/about_data.
+A csv file was downloaded which was almost 300MB in size.
+The initial document/dataset contained 1103872 observations across 24 variables.
+There were some initial processing steps performed to evaluate and delete some of the unnecessary columns and to save this data to a .rds document to reduce the size.
+This was all outlined in the "processingcode" document, which will not run automatically because the Excel (csv) file will not be available due to size concerns.
 
-It is expected that there will be some obvious differences in these levels of expression that will be asociated with the disease status and that many of these will be common between the different respiratory pathogens.
+The intial exploration and cleaning and removal of the redundant columns (columns with repeated single values without any additional information).
+Some of the other columns were renamed to have the name of the column reflect what these single value columns had contained instead.
+The columns were checked to see what they contained and to see how many unique entries there were and based upon this columns were deleted and others renamed.
 
-If potential genes or biomarkers are identified they will be run through online sources and databases (including tools such as the PubChem online tool) to identify the common names (or symbol) and proposed functions of these genes.
+Deleted columns were: DataSource, Topic, Class, StratificationCategory1/2/3, Data_Value_Footnote, Data_Value_Footnote_Symbol, Geographic_Level, TopicID
 
-## Schematic of workflow
+Renamed columns: 
+Stratification1 column to age_group, Stratification2 column to race_ethnicity, Stratification3 column to sex, Data_Value to mortalities.
 
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
+After the initial cleaning the data was evaluated again and some further small cleaning and processing steps were performed (code is visible in the processingfile-v1.qmd file) to reduce the number of rows by removing NA values.
+The resulting dataframe contains 572572 observations and 14 variables.
+
+Our summaries and the skimr function help to provide a good overview for further evaluation of the data. 
+We can see that our dataset contains a combination of numeric and character data. 
+The numeric data includes the actual incidence rates of the hypertension-related cardiovascular disease death rates, as well as the confidence intervals that were used, and geographic coordinates.
+Our character data includes the race, sex, age group, location data and the year.
+
+This character data will be used as predictors of the mortality data as we break down these different variables and look for potential patterns and associations.
+The primary outcome of interest is the mortality rates but the important potential predictors are the location, year, race/ethnicity, and sex.
+AGe groups only provide two broad categories, namely 35 to 65 years old and 65 years and older, which is a bit limited but will allow for comparisons between these two groups.
+
+A summary table was produced using the skimr function which is printed below.
 
 ```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
+skim_summary_hypertens_filtered <- readRDS(here::here("data","processed-data", "skim_summary_hypertens_filtered.rds"))
+# Print the skim summary
+print(skim_summary_hypertens_filtered)
+
 ```
 
-## Data aquisition
-_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._
+Our skimr data shows that the mean of the hypertension cardiovascular-diseases mortalities is around 203, with a standard deviation of 223.
 
-## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
+A histogram was made of the mortalities columns since this is the main (only) column of interest with numerical data.
+
+```{r}
+mortalities_histogram <- readRDS(here::here("data","processed-data", "mortalities_histogram.rds"))
+# Print the mortalities histogram
+plot(mortalities_histogram)
+```
 
 ## Statistical analysis
-_Explain anything related to your statistical analyses._
 
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
-
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
-```
-
 ## Basic statistical analysis
-
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
-```
 
@@
 
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
+_
 
@@
 ## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
 
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
 
 ## Conclusions
-_What are the main take-home messages?_
-
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
 

### 2025-02-22 code/processing-code/processingcode.R

@@ -1,12 +1,5 @@
 ###############################
-# processing script
-#
-#this script loads the raw data, processes and cleans it 
-#and saves it as Rds file in the processed-data folder
-#
-# Note the ## ---- name ---- notation
-# This is done so one can pull in the chunks of code into the Quarto document
-# see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
+# Initial data processing script
 ###############################
 
 
@@ -17,122 +10,120 @@ library(dplyr) #for data processing/cleaning
 library(tidyr) #for data processing/cleaning
 library(skimr) #for nice visualization of data 
 library(here) #to set paths
+library(readr) #open excel csv files
 
 
 ## ---- loaddata --------
 #path to data
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw-data","exampledata.xlsx")
+hypertension_cdv_adult_mortality_location <- here::here("data","raw-data","hypertension_related_CDV_disease_mortality_US_adults_35_demographics_2000-2019.csv")
+
+
 #load data. 
-#note that for functions that come from specific packages (instead of base R)
-# I often specify both package and function like so
-#package::function() that's not required one could just call the function
-#specifying the package makes it clearer where the function "lives",
-#but it adds typing. You can do it either way.
-rawdata <- readxl::read_excel(data_location)
-# We might also want to load the codebook to look at it
-codebook <- readxl::read_excel(data_location, sheet ="Codebook")
+hypertension_cdv_adult_mortality_data <- readr::read_csv(hypertension_cdv_adult_mortality_location)
 
 
-## ---- exploredata --------
+## ---- explore hypertension data --------
 #take a look at the data
-dplyr::glimpse(rawdata)
+dplyr::glimpse(hypertension_cdv_adult_mortality_data)
 #another way to look at the data
-summary(rawdata)
+summary(hypertension_cdv_adult_mortality_data)
 #yet another way to get an idea of the data
-head(rawdata)
+head(hypertension_cdv_adult_mortality_data)
 #this is a nice way to look at data
-skimr::skim(rawdata)
+skimr::skim(hypertension_cdv_adult_mortality_data)
 #look in the Codebook for a variable explanation
-print(codebook)
+print(hypertension_cdv_adult_mortality_data)
+
+View(hypertension_cdv_adult_mortality_data)
+
+
+
+
+#It is clear that this is a massive dataset with over 1.1 million observations. 
[…]

### 2025-03-17 code/processing-code/processingfile-v1.qmd
Part 3 updates

@@
 ```{r}
+#install.packages("sf")
+
 library(readxl) #for loading Excel files (in case we want to include an additional dataset)
@@
 library(knitr) #for nice tables and data integration
+library(corrplot) #for correlation matrix visualization
+library(sf) #for geospatial analysis
+library(ggpubr) #for advanced visualization tools
+library(ggplot2) #for plots
+
 ```
@@
 
+As an additional check we will look for any duplicates in the data:
+```{r}
+sum(duplicated(hypertens_adult_filtered)) # Count duplicates
+```
+We are happy that there are no duplicates in our dataset.
+
+We see from the summary that time is a character but we want numeric values and not character.
+We will confirm and convert if it is not numeric.
+(Lubridate function could have been used if the dates included month and days and time instead of only years.
+But numeric is simplest in this case.)
+```{r}
+str(hypertens_adult_filtered$year) # Check type
+#We see that time is a character (we saw this from the summary)
+hypertens_adult_filtered <- hypertens_adult_filtered %>% 
+  mutate(Year = as.numeric(Year))
+
+```
+
 We want a copy of this skimr data for a table for the mauscript. We will make edits of this later in the process.
@@
 saveRDS(mortalities_histogram, file = here::here("data","processed-data", "mortalities_histogram.rds"))
-plot(mortalities_histogram <- readRDS(here::here("data","processed-data", "mortalities_histogram.rds"))
+mortalities_histogram <- readRDS(here::here("data","processed-data", "mortalities_histogram.rds"))
 # Print the mortalities histogram
-print(mortalities_histogram))
+print(mortalities_histogram)
 ```
 
-We are happy with this for now.
+Our histogram depicts the hypertension-related cardiovasculat disease mortalities in the dataset, showing the number of mortalities and the frequency (count) of sprecific mortality values.
+The mortality counts are clustered towards the lower end of the scale, with a strong right-skew (long tail to the right/higher values).
+
+However, I acknowledge an issue in the current structure of the data with different subsets of data together, which makes these summaries limited for understanding trends.
+
+We will start by removing the total percent change values from the dataset and create a new dataset.
+```{r}
+hypertens_adult_filtered_no_percent <- hypertens_adult_filtered %>%
+  filter(Data_Value_Type != "Total percent change")
+```
+
+We will stratify the data based upon: County-level overall trends, stratify by race and sex (together and separately). All of these are based on the dataset with the total percent change removed.
+We will also create separate dataset with the percent change data from each county.
+We will not stratify by age group because of how the data is organised, this will not summarise much information or will need further summaries.
+
+Stratify by country-level overall:
+```{r}
+#Using the filter function to help us with stratification
+county_overall <- hypertens_adult_filtered_no_percent %>%
+  filter(sex == "Overall", race_ethnicity == "Overall") %>% # Keep only total county-level data
+  select(LocationID, LocationDesc, Year, mortalities, X_long, Y_lat, LocationAbbr, Data_Value_Unit, Data_Value_Type, Confidence_limit_Low, Confidence_limit_High, age_group)
+
+```
+
+Stratify by race and sex per county (removing totals and where both values are overall):
+```{r}
+county_race_sex <- hypertens_adult_filtered_no_percent %>%
+  filter(!(sex == "Overall" & race_ethnicity == "Overall")) %>% # Remove rows where both sex and race are 'Overall'
+select(LocationID, Year, LocationAbbr, LocationDesc, mortalities, Data_Value_Unit, Data_Value_Type, Confidence_limit_Low, Confidence_limit_High, age_group, race_ethnicity, sex, X_long, Y_lat)
+```
+
+To allow us to assess each of these alone we will make sex and race have their own stratifications too.
+
+Stratify by race per county:
+```{r}
+county_race <- hypertens_adult_filtered_no_percent %>%
+  filter(race_ethnicity != "Overall") %>% # Remove total values for race
+  select(LocationID, Year, LocationAbbr, LocationDesc, mortalities, Data_Value_Unit, Data_Value_Type, Confidence_limit_Low, Confidence_limit_High, age_group, race_ethnicity, X_long, Y_lat)
+```
+
+Stratify by sex per county:
+```{r}
+county_sex <- hypertens_adult_filtered_no_percent %>%
+  filter(sex != "Overall") %>% # Remove total values for sex
+  select(LocationID, Year, LocationAbbr, LocationDesc, mortalities, Data_Value_Unit, Data_Value_Type, Confidence_limit_Low, Confidence_limit_High, age_group, sex, X_long, Y_lat)
+```
+
+We create a new dataset where we stratify by percent change per county. (this may also serve as reference since we are removing this from the other stratified datasets):
+```{r}
+percent_change_data <- hypertens_adult_filtered %>%
+  filter(Data_Value_Type == "Total percent change") %>% # Keep only percent change data
+  select(LocationID, Year, LocationAbbr, LocationDesc, mortalities, Data_Value_Unit, Confidence_limit_Low, Confidence_limit_High, age_group, race_ethnicity, sex, X_long, Y_lat)
+```
+
+We are happy that we have improved the cleaning steps of our data and will revisit some summaries after saving the data.
 
@@
 The cleaned data will be saved as a new RDS file which will go under the processed data. 
+The different datasets will be saved individually and as one collective file.
 
 ```{r}
-save_data_location <- here::here("data","processed-data","processeddata.rds")
-saveRDS(hypertens_adult_filtered, file = save_data_location)
+save_processed_data_location <- here::here("data", "processed-data")
+
+saveRDS(hypertens_adult_filtered, file = file.path(save_processed_data_location, "processeddata.rds"))
 ```
+
+```{r}
+# Save individual datasets
+saveRDS(hypertens_adult_filtered_no_percent, file = file.path(save_processed_data_location, "processeddata_no_percent.rds"))
+saveRDS(county_overall, file = file.path(save_processed_data_location, "county_overall.rds"))
+saveRDS(county_race_sex, file = file.path(save_processed_data_location, "county_race_sex.rds"))
+saveRDS(county_race, file = file.path(save_processed_data_location, "county_race.rds"))
+saveRDS(county_sex, file = file.path(save_processed_data_location, "county_sex.rds"))
+saveRDS(percent_change_data, file = file.path(save_processed_data_location, "percent_change_data.rds"))
+```
+
+```{r}
+# Save all datasets in a single RDS file
+final_processed_data <- list(
+  processed_data_no_percent = hypertens_adult_filtered_no_percent,
+  county_overall = county_overall,
+  county_race_sex = county_race_sex,
+  county_race = county_race,
+  county_sex = county_sex,
+  percent_change_data = percent_change_data
+)
+saveRDS(final_processed_data, file = file.path(save_processed_data_location, "final_processed_data.rds"))
+```
+
+Running some EDA and checking summaries of data:
+
+We are going to make use of loops and creating a list of the different datasets that we created to reduce separate code to run the descriptive statisitics for each dataset.
+We are not include the percent change dataset in these evaluations.
+
+Creating this datset list:
+```{r}
+list_datasets <- list(
+  "Full Dataset" = hypertens_adult_filtered_no_percent,
+  "County Race-Sex" = county_race_sex,
+  "County Race" = county_race,
+  "County Sex" = county_sex,
+  "County Overall" = county_overall
+)
+
+```
+
+We will run the descriptive statistics for each dataset (as a single loop)
+```{r}
+# Loop through each dataset and print basic statistics
+summary_stats <- bind_rows(lapply(names(list_datasets), function(name) {
+  stats <- list_datasets[[name]] %>%
+    summarise(
+      mean = mean(mortalities, na.rm = TRUE),
+      median = median(mortalities, na.rm = TRUE),
+      sd = sd(mortalities, na.rm = TRUE),
+      min = min(mortalities, na.rm = TRUE),
+      max = max(mortalities, na.rm = TRUE)
+    )
+  stats$Dataset <- name  # Add dataset name for identification
+  return(stats)
+}))
+
+# Print summary statistics
+tibble::as_tibble(summary_stats) %>% 
+  relocate(Dataset) %>% 
+  print()
+```
+
+# Identify potential missing values in each dataset
+```{r}
+cat("\n### Missing Data Summary:\n")
+lapply(names(list_datasets), function(name) {
+  missing_count <- sum(is.na(list_datasets[[name]]))
+  cat(name, "has", missing_count, "missing values.\n")
+})
+```
+
+We want to visualise the different datasets:
+
+We will revisit the histogram with the percent change removed.
+```{r}
+# Visualize the distribution of mortality rates with a histogram
+revised_mortalities_histogram <- ggplot(hypertens_adult_filtered_no_percent, aes(x = mortalities)) +
+  geom_histogram(binwidth = 20, fill = "blue", alpha = 0.7) +
+  scale_x_continuous(breaks = seq(0, max(hypertens_adult_filtered_no_percent$mortalities, na.rm = TRUE), by = 200)) +
+  theme_minimal() +
+  labs(title = "Distribution of Mortality Rates", x = "Mortality Rates", y = "Count")
+
+saveRDS(revised_mortalities_histogram, file = here::here("data","processed-data", "revised_mortalities_histogram.rds"))
+revised_mortalities_histogram <- readRDS(here::here("data","processed-data", "revised_mortalities_histogram.rds"))
+# Print the mortalities histogram
+print(revised_mortalities_histogram)
+```
+
+This is the total dataset and so the data is similar to the previous histogram.
+There is a strong right skew and long tail.
+Peak mortalities are below 100, and the total values likely account for the second peak.
+
+We will use a boxlot to get an overview of the other data:
+
+```{r}
+# Compare mortality rates across race-sex and sex stratifications
+
+boxplot_race_sex <- ggplot(county_race_sex, aes(x = race_ethnicity, y = mortalities, fill = sex)) +
+  geom_boxplot() +
+  theme_minimal() +
+  labs(title = "Mortality Rates by Race and Sex", x = "Race/Ethnicity", y = "Mortality Rate")
+
+saveRDS(boxplot_race_sex, file = here::here("data","processed-data", "boxplot_race_sex.rds"))
+
+boxplot_sex <- ggplot(county_sex, aes(x = sex, y = mortalities)) +
+  geom_boxplot() +
+  theme_minimal() +
+  labs(title = "Mortality Rates by Sex", x = "Sex", y = "Mortality Rate")
+
+saveRDS(boxplot_sex, file = here::here("data","processed-data", "boxplot_sex.rds"))
+
+```
+
+The box plots provide a good overview of the data nd indicate that there seem to be distinct trends and differences between races.
+It also shows that there are differences between sexes but these are less marked and distinct than between races.
+We will ook at this in the statistical analysis.
+
+We want an idea on any sptial trends in the data. We are using the sf function to plot the mortality data by the latitude and longitude.
+```{r}
+# Convert data to spatial format and create a map of mortality rates
+# Ensure longitude and latitude values are numeric
+hypertens_adult_filtered_no_percent <- hypertens_adult_filtered_no_percent %>%
+  mutate(X_long = as.numeric(X_long), Y_lat = as.numeric(Y_lat))
+# Convert to spatial object
+sf_data <- st_as_sf(hypertens_adult_filtered_no_percent, coords = c("X_long", "Y_lat"), crs = 4326)
+
+# Generate a spatial plot of mortality rates
+ggplot(sf_data) +
+  geom_sf(aes(color = mortalities)) +
+  scale_color_viridis_c() +
+  theme_minimal() +
+  labs(title = "Geospatial Distribution of Mortality Rates")
+
+```
+
+This is a very nice overview and can help us to evauate the trends further or to work on a more intricate plot. We must remember that we have not separated groups or totals yet.
+
+We want an overview of any temporal (time based) trends in the data.
+We will provide a proof of concept using a single county due to the overlap of data that will occur otherwise.
+The statistical analysis will include further refinements.
+We are only working with the values where sex and race were both overall to reduce overlap.
+```{r}
+# Line plot of mortality rates over time
+# Select a single county to illustrate the trend
+selected_county <- "13037"  # County Location ID for Calhoun, GA
+county_subset <- county_overall %>% filter(LocationID == selected_county)
+
+temporal_trends_single_county <- ggplot(county_subset, aes(x = Year, y = mortalities, group = LocationID)) +
+  geom_line(alpha = 0.6, color = "blue") +
+  geom_smooth(se = FALSE, color = "red") +
+  theme_minimal() +
+  labs(title = paste("Mortality Rate Trends in Calhoun"), 
+       x = "Year", y = "Mortality Rate")
+
+print(temporal_trends_single_county)
+```
+
+## Key Findings and Next Steps
+
+- The distribution of mortality rates is right-skewed, suggesting potential transformations before modeling.
+- Mortality rates vary markedly by race and sex, justifying stratified analysis.
+- A proof-of-concept temporal trend was created for a single county (Calhoun, GA) using Location ID, ensuring unique selection.
+- Temporal trend indicate a general increase the selected county, which should be modeled.
+- All processed data has been saved into a single RDS file (`final_processed_data.rds`) for use in statistical analysis.

### 2025-03-17 products/manuscript/Manuscript.qmd
Part 3 updates

@@
 # load a few R packages
+#install.packages("magick")
 library(here)
 library(knitr)
+library(magick)
 ```
@@
 
+## Statistical analysis
+
+{{< pagebreak >}}
+
+# Results
+
+## Exploratory/Descriptive analysis
+
 A summary table was produced using the skimr function which is printed below.
@@
 
-Our skimr data shows that the mean of the hypertension cardiovascular-diseases mortalities is around 203, with a standard deviation of 223.
+Our summary data shows that the mean of the hypertension cardiovascular-diseases mortalities is around 203, with a standard deviation of 223.
 
@@
 ```{r}
-mortalities_histogram <- readRDS(here::here("data","processed-data", "mortalities_histogram.rds"))
+revised_mortalities_histogram <- readRDS(here::here("data","processed-data", "revised_mortalities_histogram.rds"))
 # Print the mortalities histogram
-plot(mortalities_histogram)
+plot(revised_mortalities_histogram)
 ```
 
-## Statistical analysis
+Our histogram depicts the hypertension-related cardiovasculat disease mortalities in the dataset, showing the number of mortalities and the frequency (count) of sprecific mortality values.
+The mortality counts are clustered towards the lower end of the scale, with a strong right-skew (long tail to the right/higher values).
+Peak mortalities are below 100, and the total values likely account for the second peak.
 
-{{< pagebreak >}}
+Boxplots were created to evaluate the mortalities by race and by sex.
 
-# Results
+```{r}
+mortality_race_boxplot <- image_read(here::here("products", "manuscript", "supplement", "mortality_by_race.png"))
+print(mortality_race_boxplot)
 
-## Exploratory/Descriptive analysis
+mortality_sex_boxplot <- image_read(here::here("products", "manuscript", "supplement", "mortality_by_sex.png"))
+print(mortality_sex_boxplot)
+```
+
+## Key Findings and Next Steps
+
+- The distribution of mortality rates is right-skewed, suggesting potential transformations before modeling.
+- Mortality rates vary markedly by race and sex, justifying stratified analysis.
+- A proof-of-concept temporal trend was created for a single county (Calhoun, GA) using Location ID, ensuring unique selection.
+- Temporal trend indicate a general increase the selected county, which should be modeled.
+- All processed data has been saved into a single RDS file (`final_processed_data.rds`) for use in statistical analysis.
 
 ## Basic statistical analysis
+
+Spatial plots of mortalities per geographic regions were generated for each race and compared on one plot.
+Two generalized linear models were evaluated, the first examined how time, race/ethnicity, and sex (plot below) influence mortality rates, and the second tested whether the effect of time varied by race/ethnicity.
+
+Overall, black individuals had the highest predcted mortality rates from the generalized linear models, followed by white and native american populations.
+Black individuals also had the highest hypertension-related mortalities on the boxplot follwoed by white individuals.
+Females appeared to have lower mortality rates than males across all races.
+
+Race, sex and time all seem to be predictors of mortality. We see differences in trends over time for different racial groups.
+
+Our spatial clustering has revealed geographic clustering of high-mortality regions, with high densities in the Eastern regions of the country.
+There also appears to be a higher density of mortalities from people in the race/ethnic group Black in certain regions.
+Asian and Pacific Islander groups had relatively lower mortality rates and geographic clustering.
+
+```{r}
+#Print the glm plot
+glm_interaction_plot <- image_read(here::here("products", "manuscript", "supplement", "glm_interaction_plot.png"))
+print(glm_interaction_plot)
+```
+
+```{r}
+#Print the combined spatial plot
+combined_spatial_plot <- image_read(here::here("products", "manuscript", "supplement", "combined_spatial_plot.png"))
+print(combined_spatial_plot)
+```
 

### 2025-03-17 code/analysis-code/statistical-analysis.R

@@ -5,52 +5,218 @@
 #and saves the results to the results folder
 
 #load needed packages. make sure they are installed.
+#install.packages("gt")
 library(ggplot2) #for plotting
 library(broom) #for cleaning up output from lm()
 library(here) #for data loading/saving
+library(tidyverse)
+library(tidymodels)
+library(gt)
+library(stringr)
+library(patchwork)
+
 
 #path to data
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","processeddata.rds")
+data_location <- here::here("data","processed-data","final_processed_data.rds")
 
 #load data. 
-mydata <- readRDS(data_location)
+final_processed_data <- readRDS(data_location)
 
+# Extract key datasets
+processed_data_no_percent <- final_processed_data[["processed_data_no_percent"]]
+county_race_sex <- final_processed_data[["county_race_sex"]]
+county_race <- final_processed_data[["county_race"]]
+county_sex <- final_processed_data[["county_sex"]]
+county_overall <- final_processed_data[["county_overall"]]
 
 ######################################
 #Data fitting/statistical analysis
 ######################################
 
 ############################
+
+# Exploring relationships between mortality rates and individual predictors
+
+# Relationship between mortality and race
+race_plot <- ggplot(county_race, aes(x = race_ethnicity, y = mortalities)) +
+  geom_boxplot() +
+  theme_minimal() +
+  labs(title = "Mortality Rates by Race/Ethnicity", x = "Race/Ethnicity", y = "Mortality Rate") +
+  theme(axis.text.x = element_text(size = 10)) +
+  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10))
+
+print(race_plot)
+ggsave(here::here("products", "manuscript", "supplement", "mortality_by_race.png"), plot = race_plot)
+
+# Relationship between mortality and sex
+sex_plot <- ggplot(county_sex, aes(x = sex, y = mortalities)) +
+  geom_boxplot() +
+  theme_minimal() +
+  labs(title = "Mortality Rates by Sex", x = "Sex", y = "Mortality Rate")
+
+print(sex_plot)
+ggsave(here::here("products", "manuscript", "supplement", "mortality_by_sex.png"), plot = sex_plot)
+
+
 #### First model fit
[…]

### 2025-04-11 code/analysis-code/statistical-analysis-v2-slim-models.qmd
Part 4 additions

---
title: "Processing CDC hypertension and CDV Mortality disease data from adults in the USA exploratory analysis script"
format: html
editor: visual
---

##Setup

```{r setup}
#load needed packages. make sure they are installed.
#install.packages("gt")
#install.packages("emmeans")
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidyverse)
library(tidymodels)
library(gt)
library(stringr)
library(patchwork)
library(emmeans)
library(sf) #for geospatial analysis
library(spdep)
library(spatialreg)
library(vip) # For variable importance plots
library(dplyr)
library(tidytext)
library(forcats)

```

```{r}
#path to data
data_location <- here::here("data","processed-data","eda_outputs.rds")

#load data. 
final_processed_data <- readRDS(data_location)

# Extract key datasets
processed_data_no_percent <- final_processed_data[["processed_data_no_percent"]]
county_race_sex <- final_processed_data[["county_race_sex"]]
county_race <- final_processed_data[["county_race"]]
county_sex <- final_processed_data[["county_sex"]]
county_overall <- final_processed_data[["county_overall"]]

```

#Data fitting/statistical analysis

##Aim 1: Temporal Trends by Age Group

We evaluated national trends in mortality rates over time using multiple modeling approaches. These included a simple linear regression, generalized linear models (GLMs) with race, sex, and age group predictors, and interaction models to assess differential trends. All models assumed a Gaussian distribution and were fit using complete national data.

####1.1 Linear Model - Overall Mortality over time

Fit a basic linear regression to estimate overall trend in mortality over time

```{r}
# Checking the relationship between mortality and year
lm_model_year <- lm(mortalities ~ Year, data = county_overall)
lm_results_year <- broom::tidy(lm_model_year)

# Display and save the regression results
print(lm_results_year)
saveRDS(lm_results_year, file = here::here("products", "manuscript", "supplement", "lm_results_year.rds"))

# Plot linear regression results
lm_plot_year <- ggplot(county_overall, aes(x = Year, y = mortalities)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  theme_minimal() +
  labs(title = "Linear Regression: Mortality Rates Over Time", x = "Year", y = "Mortality Rate")

print(lm_plot_year)
ggsave(here::here("products", "manuscript", "supplement", "lm_plot_year.png"), plot = lm_plot_year)

```

This model provides a first look at overall temporal trends in mortality rates. The flat trend line suggests that overall mortality rates changed minimally from 2000 to 2019, though high variability exists.

#####1.2 GLM: Race and Sex as predictors
GLM with race and sex as additive predictors to assess overall differences in mortality rates
```{r}
# Examining multiple predictors
glm_model <- glm(mortalities ~ Year + race_ethnicity + sex, data = county_race_sex, family = gaussian())
glm_results <- broom::tidy(glm_model)

# Display and save the GLM results
print(glm_results)
saveRDS(glm_results, file = here::here("products", "manuscript", "supplement", "glm_results.rds"))

# Plot GLM predictions
  county_race_sex$predicted_mortalities <- predict(glm_model, type = "response")
  glm_race_sex_plot <- ggplot(county_race_sex, aes(x = Year, y = predicted_mortalities, color = race_ethnicity)) +
    geom_line(size = 1) +
    theme_minimal() +
    labs(title = "GLM Predictions: Mortality Rates by Year and Race/Ethnicity", x = "Year", y = "Predicted Mortality Rate", color = "Race/Ethnicity") +
    scale_x_continuous(breaks = seq(min(county_race_sex$Year, na.rm = TRUE), max(county_race_sex$Year, na.rm = TRUE), by = 5)) +
    theme(legend.position = "right", legend.text = element_text(size = 10), legend.key.size = unit(0.4, "cm"), legend.spacing.y = unit(0.2, "cm"), axis.text.x = element_text(angle = 45, hjust = 1))

  print(glm_race_sex_plot)
  ggsave(here::here("products", "manuscript", "supplement", "glm_race_sex_plot.png"), plot = glm_race_sex_plot)

```

The GLM model incorporating race/ethnicity and sex reveals substantial disparities in mortality rates between groups. Black populations consistently show the highest predicted rates across the period.

####1.3 GLM: Year by Race interaction model:
Fit GLM including interaction between year and race to examine race-specific trends over time

```{r}

#explore whether interactions exist between year and race/sex:

glm_model_interaction <- glm(mortalities ~ Year * race_ethnicity + sex, data = county_race_sex, family = gaussian())

glm_results_interaction <- broom::tidy(glm_model_interaction)
print(glm_results_interaction)

saveRDS(glm_results_interaction, file = here::here("products", "manuscript", "supplement", "glm_results_interaction.rds"))

# Plot the interaction model results
county_race_sex$predicted_mortalities_interaction <- predict(glm_model_interaction, type = "response")
glm_interaction_plot <- ggplot(county_race_sex, aes(x = Year, y = predicted_mortalities_interaction, color = race_ethnicity)) +
  geom_line(size = 1) +
  theme_minimal() +
  labs(title = "GLM Interaction Model: Mortality Trends by Year and Race", x = "Year", y = "Predicted Mortality Rate", color = "Race/Ethnicity") +
  scale_x_continuous(breaks = seq(min(county_race_sex$Year, na.rm = TRUE), max(county_race_sex$Year, na.rm = TRUE), by = 5)) +
  theme(legend.position = "right", legend.text = element_text(size = 10), legend.key.size = unit(0.4, "cm"), legend.spacing.y = unit(0.2, "cm"), axis.text.x = element_text(angle = 45, hjust = 1))

print(glm_interaction_plot)
ggsave(here::here("products", "manuscript", "supplement", "glm_interaction_plot.png"), plot = glm_interaction_plot)

saveRDS(county_race_sex, file = here::here("results", "output", "county_race_sex_with_preds.rds"))
```
By including interaction terms, we can evaluate whether different racial/ethnic groups experienced distinct temporal trends. Here, trends diverge sharply: Black mortality rates decline, while rates rise for American Indian/Alaska Native populations.

####1.4: GLM: Year by Age group

GLM with interaction between year and age group to examine differential temporal trends
```{r}
# GLM with interaction: does time trend differ by age group?
glm_age_group <- glm(mortalities ~ Year * age_group, 
                     data = processed_data_no_percent,
                     family = gaussian())

# Tidy results for review
glm_age_group_results <- broom::tidy(glm_age_group)
print(glm_age_group_results)

# Save results
saveRDS(glm_age_group_results, file = here::here("products", "manuscript", "supplement", "glm_age_group_results.rds"))

# Add predicted values
processed_data_no_percent$predicted_glm_age <- predict(glm_age_group, type = "response")

# Plot actual vs predicted for each age group
glm_age_plot <- ggplot(processed_data_no_percent, 
                       aes(x = Year, y = mortalities, color = age_group)) +
  geom_point(alpha = 0.4) +
  geom_line(aes(y = predicted_glm_age), size = 1.2) +
  theme_minimal() +
  labs(title = "Predicted Mortality Trends by Age Group",
       subtitle = "From GLM: mortalities ~ Year * age_group",
       x = "Year", y = "Mortality Rate") +
  scale_color_brewer(palette = "Dark2")

print(glm_age_plot)

# Save plot
ggsave(here::here("products", "manuscript", "supplement", "glm_age_plot.png"), 
       plot = glm_age_plot, width = 9, height = 6)

```
This model identifies age-specific trends. Mortality among individuals aged 65+ remains substantially higher and shows modest increases, while younger age groups have lower, flatter trends.

####1.5: Slope comparison by Age group

Comparing slopes using the emmeans function.
Use estimated marginal trends (emtrends) to compare yearly slope by age group.

```{r}
# Get slope (trend) estimates for Year, stratified by age group
em_trends_age <- emtrends(glm_age_group, ~ age_group, var = "Year")

# View summary
summary(em_trends_age)
em_trends_age

# compare trends between groups (pairwise contrast)
slope_contrast <- contrast(em_trends_age, method = "pairwise")
summary(slope_contrast)

# Save
saveRDS(em_trends_age, file = here::here("products", "manuscript", "supplement", "emtrends_age_group.rds"))
saveRDS(slope_contrast, file = here::here("products", "manuscript", "supplement", "slope_contrast_age_group.rds"))

# Convert emtrends object to data frame
em_trend_df <- as.data.frame(summary(em_trends_age))

# Create horizontal bar plot with fixed x-axis breaks
slope_plot_agegroup <- ggplot(em_trend_df, aes(x = Year.trend, y = age_group)) +
  geom_col(fill = "#2C77B0", width = 0.5) +
  geom_errorbarh(aes(xmin = Year.trend - 1.96 * SE,
                     xmax = Year.trend + 1.96 * SE),
                 height = 0.15, color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_x_continuous(
    breaks = seq(0, 1.5, by = 0.25),
    limits = c(0, 1.5)
  ) +
  labs(
    title = "Estimated Annual Mortality Rate Change by Age Group",
    subtitle = "Point estimates and 95% confidence intervals",
    x = "Estimated Slope (Yearly Change)",
    y = "Age Group"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14, margin = margin(b = 6)),
    plot.subtitle = element_text(size = 11, margin = margin(b = 10)),
    axis.text = element_text(size = 12),
    panel.grid.major.y = element_blank())

# Display plot
slope_plot_agegroup

# Save figure to manuscript folder
ggsave(
  filename = here::here("results", "figures", "slope_plot_age_group.png"),
  plot = slope_plot_agegroup,
  width = 9, height = 6, dpi = 300)

# Save RDS for reproducibility
saveRDS(slope_plot_agegroup, file = here::here("results", "figures", "slope_plot_age_group.rds"))

```

Using estimated marginal trends, we formally test whether the yearly rate of change differs between age groups. Older adults (65+) show a significantly steeper upward trend compared to younger adults.

## Aim 2: Spatial Analysis of Hypertension Mortality Rates
## Evaluating spatial distribution and clustering patterns

Evaluating spatial patterns for mortality data for all races together and individually.

###2.1: Static maps by Race:

Converting county data to spatial object using longitude/latitude.
Creating static spatial map for all race groups.
```{r}
#Convert dataset to an SF spatial object
county_race_sex_sf <- county_race_sex %>%
  st_as_sf(coords = c("X_long", "Y_lat"), crs = 4326, remove = FALSE)

# Spatial plot for all races
spatial_plot_all <- ggplot() +
  geom_sf(data = county_race_sex_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Spatial Distribution of Mortality Rates (All Races)", x = "Longitude", y = "Latitude", color = "Mortality Rate")

print(spatial_plot_all)
ggsave(here::here("products", "manuscript", "supplement", "spatial_plot_all.png"), plot = spatial_plot_all)

# Spatial plot for Black race
black_race_sf <- county_race_sex_sf %>% filter(race_ethnicity == "Black")
black_spatial_plot <- ggplot() +
  geom_sf(data = black_race_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Spatial Distribution of Mortality Rates (Black Race)", x = "Longitude", y = "Latitude", color = "Mortality Rate")

print(black_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "black_spatial_plot.png"), plot = black_spatial_plot)

# Spatial plot for White race
white_race_sf <- county_race_sex_sf %>% filter(race_ethnicity == "White")
white_spatial_plot <- ggplot() +
  geom_sf(data = white_race_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Spatial Distribution of Mortality Rates (White Race)", x = "Longitude", y = "Latitude", color = "Mortality Rate")

print(white_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "white_spatial_plot.png"), plot = white_spatial_plot)

# Spatial plot for Hispanic race
hispanic_sf <- county_race_sex_sf %>% filter(race_ethnicity == "Hispanic")
hispanic_spatial_plot <- ggplot() +
  geom_sf(data = hispanic_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Spatial Distribution of Mortality Rates (Hispanic)", x = "Longitude", y = "Latitude", color = "Mortality Rate")

print(hispanic_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "hispanic_spatial_plot.png"), plot = hispanic_spatial_plot)

# Spatial plot for American Indian and Alaska Native race
native_sf <- county_race_sex_sf %>% filter(race_ethnicity == "American Indian and Alaska Native")
native_spatial_plot <- ggplot() +
  geom_sf(data = native_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Spatial Distribution of Mortality Rates\n(American Indian and Alaska Native)", x = "Longitude", y = "Latitude", color = "Mortality Rate")

print(native_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "native_spatial_plot.png"), plot = native_spatial_plot)

# Spatial plot for Asian and Pacific Islander race
asian_sf <- county_race_sex_sf %>% filter(race_ethnicity == "Asian and Pacific Islander")
asian_spatial_plot <- ggplot() +
  geom_sf(data = asian_sf, aes(color = mortalities), size = 0.7) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Spatial Distribution of Mortality Rates\n(Asian and Pacific Islander)", x = "Longitude", y = "Latitude", color = "Mortality Rate")

print(asian_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "asian_spatial_plot.png"), plot = asian_spatial_plot)

```

###2.2 Combined spatial plot

Combining all spatial race-specific maps into a multi-panel figure.
```{r}

# Combine all of the spatial charts using patchwork
combined_spatial_plot <- (spatial_plot_all | black_spatial_plot | white_spatial_plot) /
  (hispanic_spatial_plot | native_spatial_plot |
  asian_spatial_plot)

# Print and save the combined spatial plot
print(combined_spatial_plot)
ggsave(here::here("products", "manuscript", "supplement", "combined_spatial_plot.png"), 
       plot = combined_spatial_plot, width = 14, height = 10)

```

Evaluating spatial clustering of Mortality Rates

### 2.3: Global Moran’s I – All Races Combined

We will run a Global Morans analysis to evalaute spatial clustering using K-nearest neighbour clustering. We are looking at the full dataset here which is very computationally intensive. Thus we are setting the code to load the output, but if its not present (not saved) then it will run.
We have saved the output in the large

Running Global Moran's I test to evaluate overall spatial autocorrelation.
```{r spatial-weights global-morans-i, message=FALSE}
# Define file paths for output objrects
knn_path <- here::here("results", "output", "knn_neighbors.rds") # Larger, reusable object
lw_path <- here::here("results", "output", "spatial_weights_list.rds") # Larger, reusable object
moran_path <- here::here("results", "output", "global_morans_i.rds") # Smaller, result-level output

#Reuse or compute spatial weights
#computing is very computationally intensive because we are usign the full dataset

#Check if neighbor and weight objects are already saved
if (file.exists(knn_path) && file.exists(lw_path)) {
   # Load saved spatial weights
  message("Reusing saved spatial weights and neighbors.")
  knn_nb <- readRDS(knn_path)
  lw <- readRDS(lw_path)
} else {
  message("Generating spatial weights from coordinates.")
coords <- st_coordinates(county_race_sex_sf)
knn_nb <- knn2nb(knearneigh(coords, k = 5)) # Create neighbors list - 5 nearest neighbours
lw <- nb2listw(knn_nb, style = "W") # Convert to spatial weights list

  # Save for reuse
  saveRDS(knn_nb, file = knn_path)
  saveRDS(lw, file = lw_path)}

# Global Moran’s I analysis
moran_result <- moran.test(county_race_sex_sf$mortalities, lw)
print(moran_result)

# Save result
saveRDS(moran_result, file = moran_path)

```

Global Moran's I: I = 0.0435, p < 0.001, this indicates weak but statistically significant spatial autocorrelation.
There is weak-to-moderate positive spatial autocorrelation in mortality rates across U.S. counties — counties with high (or low) mortality tend to be near others with similarly high (or low) values. Although the magnitude is small (I = 0.0435), the pattern is considered highly statistically significant based on the p-value.

###2.4: Local Moran’s I (LISA)

Running Local Moran's I (LISA) to detect local spatial clusters.
```{r lisa-analysis, message=FALSE}

# Run Local Moran’s I (LISA)
local_moran <- localmoran(county_race_sex_sf$mortalities, lw)

# Build a clean spatial object with LISA values
county_lisa <- county_race_sex_sf %>%
  st_drop_geometry() %>%
  bind_cols(geometry = st_geometry(county_race_sex_sf)) %>%
  mutate(local_I = local_moran[, 1],     # Local I statistic
         local_p = local_moran[, 5]) %>% # p-value
  st_as_sf()

# Save outputs
saveRDS(local_moran, file = here::here("results", "output", "local_morans_matrix.rds"))
saveRDS(county_lisa, file = here::here("results", "output", "county_race_sex_sf_with_lisa.rds"))
saveRDS(moran_result, here::here("results", "output", "us_global_morans_i.rds"))

```

Creating map of statistically significant LISA clusters (p < 0.05).
```{r}

# Create a data frame with centroids for plotting and significance flag
county_lisa_coords <- county_lisa %>%
  mutate(long = st_coordinates(.)[, 1],
    lat = st_coordinates(.)[, 2],
    significant_cluster = ifelse(local_p < 0.05, "p < 0.05", "Not Significant")) %>%
  st_drop_geometry()

# Plot using points instead of polygons
lisa_point_plot <- ggplot(county_lisa_coords) +
  geom_point(aes(x = long, y = lat, color = significant_cluster), alpha = 0.6, size = 0.7) +
  scale_color_manual(values = c("gray80", "red"),
                     labels = c("Not Significant", "p < 0.05")) +
  labs(title = "Counties with Statistically Significant Clustering",
       color = "LISA Significance") +
  coord_equal() +
  theme_minimal(base_size = 12)

# Save plot
ggsave(filename = here::here("products", "manuscript", "supplement", "lisa_significance_map_centroids.png"),
  plot = lisa_point_plot,
  width = 10,
  height = 6)

print(lisa_point_plot)

```
Classifying counties based on LISA cluster type using z-scores.

```{r classify-lisa-clusters, message=FALSE}
# Compute spatial lag of mortality
spatial_lag <- lag.listw(lw, county_lisa$mortalities)

# Standardize original values and spatial lags (z-scores)
mortality_z <- scale(county_lisa$mortalities)[, 1]
lag_z       <- scale(spatial_lag)[, 1]

# Add z-scores and spatial lag to dataset
county_lisa <- county_lisa %>%
  mutate(
    mortality_z = mortality_z,
    lag_z = lag_z,
    lisa_cluster_type = case_when(
      local_p >= 0.05              ~ "Not Significant",
      mortality_z > 0 & lag_z > 0  ~ "High-High",
      mortality_z < 0 & lag_z < 0  ~ "Low-Low",
      mortality_z > 0 & lag_z < 0  ~ "High-Low (Outlier)",
      mortality_z < 0 & lag_z > 0  ~ "Low-High (Outlier)",
      TRUE                         ~ NA_character_))

```

Generating map of LISA cluster types (High-High, Low-Low, Outliers).

```{r lisa-cluster-type-map, message=FALSE}
# Create centroid-based coordinates for plotting
county_lisa_coords <- county_lisa %>%
  mutate(long = st_coordinates(.)[, 1],
         lat = st_coordinates(.)[, 2]) %>%
  st_drop_geometry()

# Define cluster color palette
cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray85"
)

# Plot cluster types
lisa_cluster_plot <- ggplot(county_lisa_coords) +
  geom_point(aes(x = long, y = lat, color = lisa_cluster_type), alpha = 0.7, size = 0.75) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "Local Spatial Clusters of Mortality (LISA)",
    subtitle = "High–High = Hotspot, Low–Low = Coldspot, Others = Outliers",
    color = "Cluster Type"
  ) +
  coord_equal() +
  theme_minimal(base_size = 12)

# Print and save
print(lisa_cluster_plot)

ggsave(
  here::here("products", "manuscript", "supplement", "lisa_cluster_type_map.png"),
  plot = lisa_cluster_plot,
  width = 10, height = 6
)

```
Local Moran’s I was computed on the full national dataset using 5-nearest-neighbor spatial weights. Each county was represented by a centroid, and clustering patterns reflect spatial autocorrelation of mortality rates across all years and subgroups.

####Focusing at a State level
Due to the size of the dataset it is challenging to work with the full dataset.
For convenience and because the SouthEast is historically known to have a high prevalence of cardiovasccular mortalities we will focus on Georgia for this next step.
Our previous analysis support that there is a general distribution of cases here, even if there are not distinct hotspots.

Analyzing spatial clusters in Georgia using LISA.
```{r}
# Georgia: Create or load spatial weights
ga_knn_path <- here::here("results", "output", "ga_knn_neighbors.rds")
ga_lw_path  <- here::here("results", "output", "ga_spatial_weights.rds")

county_ga_sf <- county_race_sex_sf %>%
  filter(LocationAbbr == "GA")

if (file.exists(ga_knn_path) && file.exists(ga_lw_path)) {
  message("Reusing saved GA spatial weights and neighbors.")
  ga_knn_nb <- readRDS(ga_knn_path)
  ga_lw <- readRDS(ga_lw_path)
} else {
  message("Generating GA spatial weights.")
  ga_coords <- st_coordinates(county_ga_sf)
  ga_knn_nb <- knn2nb(knearneigh(ga_coords, k = 5))
  ga_lw     <- nb2listw(ga_knn_nb, style = "W")
  saveRDS(ga_knn_nb, file = ga_knn_path)
  saveRDS(ga_lw, file = ga_lw_path)
}

```

```{r ga-lisa-analysis, message=FALSE}

# Global Moran's I
ga_moran_result <- moran.test(county_ga_sf$mortalities, ga_lw)
print(ga_moran_result)
saveRDS(ga_moran_result, here::here("results", "output", "ga_global_morans_i.rds"))

# Local Moran's I (LISA)
ga_local_moran <- localmoran(county_ga_sf$mortalities, ga_lw)

# Create cluster type classification
ga_spatial_lag <- lag.listw(ga_lw, county_ga_sf$mortalities)
ga_z <- scale(county_ga_sf$mortalities)[,1]
ga_lag_z <- scale(ga_spatial_lag)[,1]
ga_p <- ga_local_moran[,5]

county_ga_sf <- county_ga_sf %>%
  mutate(
    local_I = ga_local_moran[, 1],
    local_p = ga_p,
    mortality_z = ga_z,
    lag_z = ga_lag_z,
    cluster_type = case_when(
      local_p >= 0.05 ~ "Not Significant",
      mortality_z > 0 & lag_z > 0 ~ "High-High",
      mortality_z < 0 & lag_z < 0 ~ "Low-Low",
      mortality_z > 0 & lag_z < 0 ~ "High-Low (Outlier)",
      mortality_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
      TRUE ~ NA_character_))

# Plot cluster types using centroids
ga_coords_df <- county_ga_sf %>%
  mutate(long = st_coordinates(.)[,1],
         lat = st_coordinates(.)[,2]) %>%
  st_drop_geometry()

cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray80"
)

ga_lisa_plot <- ggplot(ga_coords_df) +
  geom_point(aes(x = long, y = lat, color = cluster_type), alpha = 0.8, size = 2) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "LISA Cluster Types: Georgia",
    subtitle = "High–High = Hotspot; Low–Low = Coldspot; Others = Outliers",
    color = "Cluster Type"
  ) +
  coord_equal() +
  theme_minimal(base_size = 13)

# Display and save
print(ga_lisa_plot)

ggsave(
  here::here("products", "manuscript", "supplement", "ga_lisa_cluster_plot.png"),
  plot = ga_lisa_plot,
  width = 8,
  height = 6)
```

A spatial weights matrix was created for Georgia counties using nearest-neighbor distances. LISA statistics were computed to explore clustering, but no statistically significant local hotspots were observed.

Analyzing spatial clusters in Mississippi using LISA.

```{r ms-lisa-analysis, message=FALSE}

# Define file paths for MS spatial objects
ms_knn_path <- here::here("results", "output", "ms_knn_neighbors.rds")
ms_lw_path  <- here::here("results", "output", "ms_spatial_weights.rds")
ms_moran_path <- here::here("results", "output", "ms_global_morans_i.rds")

# Subset to Mississippi (MS)
county_ms_sf <- county_race_sex_sf %>%
  filter(LocationAbbr == "MS")

# Reuse or generate new spatial weights
if (file.exists(ms_knn_path) && file.exists(ms_lw_path)) {
  message("Reusing saved MS spatial weights and neighbors.")
  ms_knn_nb <- readRDS(ms_knn_path)
  ms_lw     <- readRDS(ms_lw_path)
} else {
  message("Generating MS spatial weights.")
  ms_coords <- st_coordinates(county_ms_sf)
  ms_knn_nb <- knn2nb(knearneigh(ms_coords, k = 5))
  ms_lw     <- nb2listw(ms_knn_nb, style = "W")

  # Save for reuse
  saveRDS(ms_knn_nb, file = ms_knn_path)
  saveRDS(ms_lw, file = ms_lw_path)
}

# Global Moran’s I for MS
ms_moran_result <- moran.test(county_ms_sf$mortalities, ms_lw)
print(ms_moran_result)
saveRDS(ms_moran_result, file = ms_moran_path)

# Local Moran’s I (LISA)
ms_local_moran <- localmoran(county_ms_sf$mortalities, ms_lw)

# Classify cluster types
ms_spatial_lag <- lag.listw(ms_lw, county_ms_sf$mortalities)
ms_z <- scale(county_ms_sf$mortalities)[,1]
ms_lag_z <- scale(ms_spatial_lag)[,1]
ms_p <- ms_local_moran[,5]

county_ms_sf <- county_ms_sf %>%
  mutate(
    local_I = ms_local_moran[, 1],
    local_p = ms_p,
    mortality_z = ms_z,
    lag_z = ms_lag_z,
    cluster_type = case_when(
      local_p >= 0.05 ~ "Not Significant",
      mortality_z > 0 & lag_z > 0 ~ "High-High",
      mortality_z < 0 & lag_z < 0 ~ "Low-Low",
      mortality_z > 0 & lag_z < 0 ~ "High-Low (Outlier)",
      mortality_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
      TRUE ~ NA_character_
    )
  )

# Plot with centroids
ms_coords_df <- county_ms_sf %>%
  mutate(long = st_coordinates(.)[,1],
         lat = st_coordinates(.)[,2]) %>%
  st_drop_geometry()

cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray80"
)

ms_lisa_plot <- ggplot(ms_coords_df) +
  geom_point(aes(x = long, y = lat, color = cluster_type), alpha = 0.8, size = 2.5) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "LISA Cluster Types: Mississippi",
    subtitle = "High–High = Hotspot; Low–Low = Coldspot; Others = Outliers",
    color = "Cluster Type"
  ) +
  coord_equal() +
  theme_minimal(base_size = 13)

print(ms_lisa_plot)

ggsave(
  filename = here::here("products", "manuscript", "supplement", "ms_lisa_cluster_plot.png"),
  plot = ms_lisa_plot,
  width = 8,
  height = 6
)

```

LISA analysis for Mississippi counties detected spatial relationships using 5-nearest-neighbor weights. Similar to other states, no prominent local clusters were found despite elevated mortality rates in some areas.

Analyzing spatial clusters in Texas using LISA.

```{r tx-lisa-analysis, message=FALSE}

# Define file paths for Texas-specific spatial objects
tx_knn_path <- here::here("results", "output", "tx_knn_neighbors.rds")
tx_lw_path  <- here::here("results", "output", "tx_spatial_weights.rds")
tx_moran_path <- here::here("results", "output", "tx_global_morans_i.rds")

# Subset to Texas counties
county_tx_sf <- county_race_sex_sf %>%
  filter(LocationAbbr == "TX")

# Reuse or generate neighbors/weights
if (file.exists(tx_knn_path) && file.exists(tx_lw_path)) {
  message("Reusing saved TX spatial weights and neighbors.")
  tx_knn_nb <- readRDS(tx_knn_path)
  tx_lw     <- readRDS(tx_lw_path)
} else {
  message("Generating TX spatial weights.")
  tx_coords <- st_coordinates(county_tx_sf)
  tx_knn_nb <- knn2nb(knearneigh(tx_coords, k = 5))
  tx_lw     <- nb2listw(tx_knn_nb, style = "W")

  # Save for reuse
  saveRDS(tx_knn_nb, file = tx_knn_path)
  saveRDS(tx_lw, file = tx_lw_path)
}

```

```{r tx-lisa-analysis, message=FALSE}

# Global Moran’s I for Texas
tx_moran_result <- moran.test(county_tx_sf$mortalities, tx_lw)
print(tx_moran_result)
saveRDS(tx_moran_result, file = tx_moran_path)

# Local Moran’s I (LISA)
tx_local_moran <- localmoran(county_tx_sf$mortalities, tx_lw)

# Spatial lag and z-scores
tx_spatial_lag <- lag.listw(tx_lw, county_tx_sf$mortalities)
tx_z <- scale(county_tx_sf$mortalities)[,1]
tx_lag_z <- scale(tx_spatial_lag)[,1]
tx_p <- tx_local_moran[,5]

# Add LISA values and classification
county_tx_sf <- county_tx_sf %>%
  mutate(
    local_I = tx_local_moran[, 1],
    local_p = tx_p,
    mortality_z = tx_z,
    lag_z = tx_lag_z,
    cluster_type = case_when(
      local_p >= 0.05 ~ "Not Significant",
      mortality_z > 0 & lag_z > 0 ~ "High-High",
      mortality_z < 0 & lag_z < 0 ~ "Low-Low",
      mortality_z > 0 & lag_z < 0 ~ "High-Low (Outlier)",
      mortality_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
      TRUE ~ NA_character_
    )
  )

# Convert to points (centroids) for visualization
tx_coords_df <- county_tx_sf %>%
  mutate(long = st_coordinates(.)[,1],
         lat = st_coordinates(.)[,2]) %>%
  st_drop_geometry()

# Cluster colors
cluster_colors <- c(
  "High-High" = "red3",
  "Low-Low" = "blue3",
  "High-Low (Outlier)" = "orange",
  "Low-High (Outlier)" = "purple",
  "Not Significant" = "gray80"
)

# Plot
tx_lisa_plot <- ggplot(tx_coords_df) +
  geom_point(aes(x = long, y = lat, color = cluster_type), alpha = 0.8, size = 2) +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "LISA Cluster Types: Texas",
    subtitle = "High–High = Hotspot; Low–Low = Coldspot; Others = Outliers",
    color = "Cluster Type"
  ) +
  coord_equal() +
  theme_minimal(base_size = 13)

print(tx_lisa_plot)

# Save plot
ggsave(
  filename = here::here("products", "manuscript", "supplement", "tx_lisa_cluster_plot.png"),
  plot = tx_lisa_plot,
  width = 9, height = 6
)
```

LISA analysis was performed on counties in Texas using centroid-based spatial weights. While global spatial autocorrelation was detected, no statistically significant local clusters were identified.

Analyzing spatial clusters in Kentucky using LISA.

```{r}

# Paths for saving spatial weights
ky_knn_path <- here::here("results", "output", "ky_knn_neighbors.rds")
ky_lw_path  <- here::here("results", "output", "ky_spatial_weights.rds")

# Subset to Kentucky counties
county_ky_sf <- county_race_sex_sf %>%
  filter(LocationAbbr == "KY")

# Reuse or generate spatial weights for Kentucky
if (file.exists(ky_knn_path) && file.exists(ky_lw_path)) {
  message("Reusing saved KY spatial weights and neighbors.")
  ky_knn_nb <- readRDS(ky_knn_path)
  ky_lw     <- readRDS(ky_lw_path)
} else {
  message("Generating KY spatial weights.")
  ky_coords <- st_coordinates(county_ky_sf)
  ky_knn_nb <- knn2nb(knearneigh(ky_coords, k = 5))
  ky_lw     <- nb2listw(ky_knn_nb, style = "W")

  # Save for reuse
  saveRDS(ky_knn_nb, file = ky_knn_path)
  saveRDS(ky_lw, file = ky_lw_path)
}

# Compute local Moran’s I
local_moran_ky <- localmoran(county_ky_sf$mortalities, ky_lw)

# Add results to object
county_ky_sf <- county_ky_sf %>%
  mutate(
    local_I = local_moran_ky[, 1],
    local_p = local_moran_ky[, 5],
    mortality_z = scale(mortalities)[, 1],
    lag_z = lag.listw(ky_lw, mortality_z),
    lisa_cluster = case_when(
      local_p > 0.05 ~ "Not Significant",
      mortality_z > 0 & lag_z > 0 ~ "High-High",
      mortality_z < 0 & lag_z < 0 ~ "Low-Low",
      mortality_z > 0 & lag_z < 0 ~ "High-Low (Outlier)",
      mortality_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
      TRUE ~ NA_character_
    )
  )

# Get centroid coordinates
county_ky_coords <- county_ky_sf %>%
  mutate(long = st_coordinates(.)[, 1],
         lat = st_coordinates(.)[, 2]) %>%
  st_drop_geometry()

# Plot
lisa_ky_plot <- ggplot(county_ky_coords) +
  geom_point(aes(x = long, y = lat, color = lisa_cluster), alpha = 0.7, size = 0.8) +
  scale_color_manual(
    values = c("High-High" = "red", "Low-Low" = "blue",
               "High-Low (Outlier)" = "orange", "Low-High (Outlier)" = "purple", "Not Significant" = "gray80"),
    na.value = "black"
  ) +
  labs(
    title = "LISA Cluster Types: Kentucky",
    subtitle = "High–High = Hotspot; Low–Low = Coldspot; Others = Outliers",
    color = "Cluster Type"
  ) +
  coord_equal() +
  theme_minimal(base_size = 12)

# Save
ggsave(
  here::here("products", "manuscript", "supplement", "lisa_cluster_kentucky.png"),
  plot = lisa_ky_plot, width = 9, height = 6)

# Compute Global Moran’s I for Kentucky
ky_moran_result <- moran.test(county_ky_sf$mortalities, ky_lw)

# Save the result
saveRDS(ky_moran_result, here::here("results", "output", "ky_global_morans_i.rds"))

print(lisa_ky_plot)

```
the LISA plot shows spatial patterns across all available data, not just one year

Compile Global Moran’s I test results from the USA and selected high-mortality states (GA, MS, TX, KY).
This table summarizes the spatial autocorrelation of mortality rates.

2.5 Global Moran's I Summary Table

```{r}
# Already saved or placed earlier i
extract_moran_values <- function(obj, geo) {
  tibble(
    Geography = geo,
    `Moran's I` = round(obj$estimate[["Moran I statistic"]], 4),
    Expectation = round(obj$estimate[["Expectation"]], 4),
    Variance = round(obj$estimate[["Variance"]], 6),
    `Z Statistic` = round(obj$statistic, 4),
    `p-value` = signif(obj$p.value, 4))}

# Load all Moran's I results
usa_moran <- readRDS(here::here("results", "output", "global_morans_i.rds"))
ga_moran  <- readRDS(here::here("results", "output", "ga_global_morans_i.rds"))
ms_moran  <- readRDS(here::here("results", "output", "ms_global_morans_i.rds"))
tx_moran  <- readRDS(here::here("results", "output", "tx_global_morans_i.rds"))
ky_moran  <- readRDS(here::here("results", "output", "ky_global_morans_i.rds"))

# Build summary table
moran_summary <- bind_rows(
  extract_moran_values(usa_moran, "USA (All Counties)"),
  extract_moran_values(ga_moran,  "Georgia"),
  extract_moran_values(ms_moran,  "Mississippi"),
  extract_moran_values(tx_moran,  "Texas"),
  extract_moran_values(ky_moran,  "Kentucky"))

# Format with gt
moran_gt_tbl <- moran_summary %>%
  gt() %>%
  tab_header(
    title = "Global Moran's I Statistics by Geography",
    subtitle = "Spatial Autocorrelation of Mortality Rates"
  ) %>%
  fmt_number(columns = 2:5, decimals = 4) %>%  # Keep decimals for other metrics
  fmt_scientific(columns = vars(`p-value`), decimals = 2) %>%  # Use sci notation for p-values
  cols_label(
    `Moran's I` = "Moran's I",
    `Z Statistic` = "Z Statistic",
    `p-value` = "p-value"
  ) %>%
  tab_options(
    table.font.size = px(12),
    heading.title.font.size = px(14),
    heading.subtitle.font.size = px(12),
    column_labels.font.weight = "bold"
  )

# Save table
saveRDS(moran_gt_tbl, here::here("results", "tables", "moran_global_summary.rds"))

# View
moran_gt_tbl
```

A summary table of Global Moran’s I statistics for the full U.S. dataset and selected states (GA, MS, TX, KY) is shown below, indicating statistically significant spatial autocorrelation in all geographies examined (p < 0.001).

Local Indicators of Spatial Association (LISA) – Statistically Significant Clustering of Mortality Rates in Kentucky.
This map highlights counties with statistically significant spatial clustering of hypertension-related cardiovascular disease (CVD) mortality rates, as detected by Local Moran’s I. Each point represents a county centroid. Counties shown in red are spatial clusters where both the county and its neighbors have high mortality values (high-high) or display significant spatial autocorrelation (e.g., high-low or low-high). Spatial relationships were calculated using 5-nearest-neighbor spatial weights based on county centroids. This analysis includes data from all available years and demographic subgroups; therefore, significance reflects elevated spatial correlation within any combination of groupings, not just overall rates.

Local spatial analysis (LISA) identified a small number of statistically significant clusters in Kentucky, including high-mortality counties adjacent to other high-mortality areas. These patterns support the presence of localized mortality hotspots and may inform targeted investigation or intervention efforts.

##Aim 3: Demographic Disparities

### 3.1: Estimated sloped by Race and Sex

Slope Comparison by Race and Sex using emmeans function for all ages.

Extract estimated annual mortality trends by race/ethnicity and sex across all adults (35+). These values represent the modeled rate of change in mortality per year.

```{r}
# Extract estimated yearly trends (slopes) by race_ethnicity and sex
em_trends_race_sex <- emtrends(glm_model_interaction, ~ race_ethnicity + sex, var = "Year")

# Summarize estimated slopes
# View slope estimates (with SE and CI) for each group
summary(em_trends_race_sex)
print(em_trends_race_sex)

# Compare slopes between race/sex groups
slope_contrast_race_sex <- contrast(em_trends_race_sex, method = "pairwise")

# View significance of differences in slopes
summary(slope_contrast_race_sex)
print(slope_contrast_race_sex)

# Save slope estimates and pairwise contrasts
saveRDS(em_trends_race_sex, file = here::here("results", "output", "emtrends_race_sex.rds"))
saveRDS(slope_contrast_race_sex, file = here::here("results", "output", "slope_contrast_race_sex.rds"))

# Plot slope estimates
slope_plot_race_sex <- plot(em_trends_race_sex) +
  ggtitle("Estimated Annual Mortality Change by Race and Sex") +
  ylab("Yearly Slope Estimate") +
  xlab("Race and Sex Group") +
  theme_minimal(base_size = 12)

# Save and display the plot
ggsave(here::here("results", "output", "slope_plot_race_sex.png"),
       plot = slope_plot_race_sex, width = 10, height = 6)
print(slope_plot_race_sex)
```

Estimated yearly mortality rate trends were computed for each race and sex group. Positive slope estimates indicate increasing mortality over time, while negative values suggest improvements. Pairwise comparisons were used to assess whether these differences were statistically significant across groups. This was for all ages aged 35+ (both groups combined).

###3.2 Demogrpahic disparities by Age group

We now wish to use a model to extract yearly mortality trends by combinations of race/ethnicity and sex while adjusting for age group to see the impact of demosgraphic disparities between age groups. To assess disparities across age, we fit a GLM including three-way interactions between year, race/ethnicity, and sex, adjusting for age group. This allows age-stratified analysis of temporal trends.

Fit the Interaction Model
```{r}
# GLM with 3-way interaction to explore mortality trends by Year * race_ethnicity * sex, stratified by age group

## ---- load_glm_race_sex_age_model, message=TRUE, warning=FALSE -------------------

glm_path <- here::here("results", "output", "glm_race_sex_age_model.rds")

if (file.exists(glm_path)) {
  message("Loading saved GLM model with 3-way interaction...")
  glm_race_sex_age <- readRDS(glm_path)
} else {
  message("Fitting GLM model with 3-way interaction...")
  glm_race_sex_age <- glm(mortalities ~ Year * race_ethnicity * sex + age_group,
                          data = county_race_sex, family = gaussian())
  saveRDS(glm_race_sex_age, glm_path)
}

#glm_race_sex_age <- glm(mortalities ~ Year * race_ethnicity * sex + age_group, data = processed_data_no_percent, family = gaussian())

# Save model
#saveRDS(glm_race_sex_age, here::here("results", "output", "glm_race_sex_age_model.rds"))

```

We fit a GLM with 3-way interactions between year, race/ethnicity, and sex, while adjusting for age group. This allows us to examine mortality trends by race and sex, separately for younger and older adults.

Get Estimated Slopes Stratified by Age Group
```{r}
# Extract yearly slopes within each race-sex group, stratified by age_group
em_trends_race_sex_age <- emtrends(glm_race_sex_age,
                                   ~ race_ethnicity + sex | age_group,
                                   var = "Year")

# Save slopes
saveRDS(em_trends_race_sex_age, here::here("results", "output", "emtrends_race_sex_by_age.rds"))

# Print to console
summary(em_trends_race_sex)
```

This extracts slope estimates (rate of annual change in mortality) for each demographic subgroup, separately for younger and older adults. Estimated yearly changes in mortality rates were extracted for each race/sex group, stratified by age. This approach highlights disparities not only across race and sex but also how these trends differ across age strata.

###3.3 Pairwise Comparisons Within Each Age Group

Compare estimated slopes between all demographic subgroups within each age group using pairwise contrasts.

```{r}
# Compare slopes between race/sex groups within each age group
slope_contrast_race_sex_age <- contrast(em_trends_race_sex_age, method = "pairwise")

# Save
saveRDS(slope_contrast_race_sex_age, here::here("results", "output", "slope_contrast_race_sex_by_age.rds"))

```

Pairwise comparisons were conducted to formally test differences in temporal mortality trends between groups within each age group. These contrasts help assess whether slope differences between groups are statistically significant within each age group.

###3.4: Visualize Slope Contrasts
Top slope differences (absolute magnitude) were visualized to highlight which subgroup comparisons showed the greatest disparities. Significant differences (p < 0.05) are shown in red.
```{r}
# Summarize the contrast object
contrast_summary <- as.data.frame(summary(slope_contrast_race_sex_age)) %>%
  filter(!is.na(estimate)) %>%  # remove any NAs if they exist
  mutate(
    significant = p.value < 0.05,
    contrast = factor(contrast))

contrast_top <- contrast_summary %>%
  arrange(desc(abs(estimate))) %>%
  group_by(age_group) %>%
  slice_head(n = 20) %>%   # Top 20 slope differences per age group
  ungroup()

# Plot
contrast_plot <- ggplot(contrast_top, aes(x = estimate, y = reorder(contrast, estimate), color = significant)) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = estimate - 1.96 * SE, xmax = estimate + 1.96 * SE), height = 0.3) +
  facet_wrap(~ age_group, scales = "free_y") +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "gray40")) +
  labs(
    title = "Top Pairwise Differences in Annual Mortality Trends by Race and Sex",
    x = "Estimated Difference in Yearly Mortality Slope",
    y = "Group Comparison",
    color = "Significant (p < 0.05)") +
  theme_minimal(base_size = 11)

# Save and display
ggsave(
  here::here("results", "output", "pairwise_slope_contrasts_filtered.png"),
  plot = contrast_plot, width = 11, height = 8)
print(contrast_plot)

```

To improve clarity, cleaned labels were applied to simplify group names and facet plots were split by age group. Bar height reflects annual change in mortality; error bars represent 95% confidence intervals.

```{r}
# Convert estimated emtrends object to a data frame for plotting
# Create a group label that includes race, sex, and age group
slope_df <- as.data.frame(em_trends_race_sex_age) %>%
  mutate(group = paste(race_ethnicity, sex),
         group = paste(group, age_group, sep = "___"))

# Helper function to clean group labels (removes age group text)
clean_labels <- function(df) {
  df %>%
    mutate(
      group = str_remove(group, "___Ages 35-64 years"),
      group = str_remove(group, "___Ages 65\\+ years"),
      group = str_replace(group, "Asian and Pacific Islander", "Asian & Pacific\nIslander"),
      group = str_replace(group, "American Indian and Alaska Native", "American Indian\n& Alaska Native"),
      group = str_replace(group, "Overall Female", "Female"),
      group = str_replace(group, "Overall Male", "Male"),
      group = str_replace(group, "Overall Overall", "All Groups"),
      group = str_replace(group, "White Overall", "White"),
      group = str_replace(group, "Black Overall", "Black"),
      group = str_replace(group, "Hispanic Overall", "Hispanic"),
      group = str_replace(group, "Asian & Pacific\nIslander Overall", "Asian & Pacific\nIslander"),
      group = str_replace(group, "American Indian\n& Alaska Native Overall", "American Indian\n& Alaska Native"),
      group = str_trim(group))}

# Apply and split
slope_df_35_64 <- slope_df %>% filter(age_group == "Ages 35-64 years") %>% clean_labels()
slope_df_65_plus <- slope_df %>% filter(age_group == "Ages 65+ years") %>% clean_labels()

# Common ggplot function
make_slope_plot <- function(data, title, fill_color) {
  data <- data %>% 
    filter(!is.na(group), !is.na(Year.trend)) %>%   # remove rows with missing group or slope
    mutate(group = as.factor(group))                # make sure group is a factor

  ggplot(data, aes(x = Year.trend, y = fct_reorder(group, Year.trend))) +
    geom_col(fill = fill_color) +
    geom_errorbarh(aes(xmin = Year.trend - 1.96 * SE,
                       xmax = Year.trend + 1.96 * SE),
                   height = 0.3, color = "black", alpha = 0.7) +
    labs(
      title = title,
      x = "Estimated Annual Change in Mortality Rate",
      y = "Race and Ethnicity"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(hjust = 0.5),
      axis.text.y = element_text(size = 11),
      axis.text.x = element_text(size = 11))}

# Generate plots
slope_plot_35_64 <- make_slope_plot(slope_df_35_64, "Estimated Annual Mortality Change (Ages 35–64)", "#F8766D")
slope_plot_65_plus <- make_slope_plot(slope_df_65_plus, "Estimated Annual Mortality Change (Ages 65+)", "#00BFC4")

# Save plots
ggsave(here("results", "figures", "slope_plot_35_64.png"),
       slope_plot_35_64, width = 11, height = 7, dpi = 300)

ggsave(here("results", "figures", "slope_plot_65_plus.png"),
       slope_plot_65_plus, width = 11, height = 7, dpi = 300)

# Save RDS objects (optional)
saveRDS(slope_plot_35_64, here("results", "figures", "slope_plot_35_64.rds"))
saveRDS(slope_plot_65_plus, here("results", "figures", "slope_plot_65_plus.rds"))

slope_plot_35_64
slope_plot_65_plus
```

Estimated annual change in hypertension-related CVD mortality per 100,000 adults, stratified by race/ethnicity and age group (2000–2019). Bars represent modeled slope estimates, and error bars represent 95% confidence intervals. “Overall” groups include both sexes; sex-specific estimates were excluded where modeling was not feasible.

We assessed yearly changes in hypertension-related cardiovascular disease (CVD) mortality across race/ethnicity and sex groups, stratified by age (35–64 and 65+ years). Modeled slope estimates revealed significant disparities in temporal mortality trends. Black individuals consistently showed the largest reductions in mortality over time, while males and American Indian/Alaska Native populations experienced the steepest increases. These patterns were consistent across both age groups. Pairwise comparisons highlighted statistically significant differences between subgroups, especially between Black adults and other demographic groups. Overall, these findings indicate that improvements in CVD mortality have not been shared equally across populations and age groups.

##Aim 4: Predict County-level Mortality rates

We will make use of some models to try to predict mortality rates based on the different potential predictors (location, race, sex, age-group and year)

####Data Preparation for modeling

```{r}
# Set seed for reproducibility
set.seed(123)

# Use the cleaned modeling dataset
model_data <- county_race_sex %>%
  drop_na(mortalities, Year, age_group, race_ethnicity, sex)  # Remove rows with missing key variables

# Initial train/test split (80/20) - test split stratified by outcome
data_split <- initial_split(model_data, prop = 0.8, strata = mortalities)
train_data <- training(data_split)
test_data  <- testing(data_split)

# 10-fold cross-validation folds (stratified by outcome)
cv_folds <- vfold_cv(train_data, v = 10, strata = mortalities)

```

Generating the processing recipe and model workflow

```{r}
# Preprocessing recipe
mortality_recipe <- recipe(mortalities ~ Year + age_group + race_ethnicity + sex, data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%  # Convert categorical predictors to dummies
  step_zv(all_predictors()) %>%             # Remove zero-variance predictors (if any)
  step_normalize(all_numeric_predictors())  # Normalize numeric predictors (being cautious)

# Define random forest model
rf_model <- rand_forest(
  mtry = tune(),     # number of variables to try at each split 
  trees = 500,       # total number of trees
  min_n = tune()     # minimum data points in a node 
) %>%
  set_engine("ranger", importance = "impurity") %>%  # impurity-based variable importance
  set_mode("regression")

# Combine into a workflow
rf_workflow <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(mortality_recipe)

```

Hyperparamter tuning and fitting a random forest model:
We will tune a random forest with 10-Fold cross-validation:

# Define path to save tuning results

```{r}
# Define file paths for saving/loading tuning results
tuned_path     <- here::here("results", "large-files", "rf_tuned.rds")
metrics_path   <- here::here("results", "output", "rf_tuned_metrics.rds")
best_param_path <- here::here("results", "output", "rf_best_params.rds")

# Check if tuning results already exist
if (file.exists(tuned_path) && file.exists(metrics_path) && file.exists(best_param_path)) {
  message( "Tuning results already saved. Loading RF tuning results...")

  # Load saved objects
  rf_tuned <- readRDS(tuned_path)
  rf_tuned_metrics <- readRDS(metrics_path)
  best_rf_params <- readRDS(best_param_path)

} else {
  message(" Running Random Forest tuning...")
# Set up a small tuning grid for efficiency
rf_grid <- grid_regular(
  mtry(range = c(1, 4)),     # Explore 1 to 4 predictors
  min_n(range = c(5, 25)),   # Minimum observations per leaf
  levels = 3)                # Total levels in grid for each param

# Tune the random forest model
set.seed(123)
rf_tuned <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(save_pred = TRUE))

# Collect metrics from the tuning process
rf_tuned_metrics <- collect_metrics(rf_tuned)
best_rf_params <- select_best(rf_tuned, metric = "rmse")

# Save results to avoid re-running this chunk
  saveRDS(rf_tuned, tuned_path)
  saveRDS(rf_tuned_metrics, metrics_path)
  saveRDS(best_rf_params, best_param_path)

  message("Tuning complete and results saved.")}

```
```{r}
## ---- fit_or_load_rf_model, message=TRUE, warning=FALSE -------------------

use_slim_model <- FALSE # Toggle this to FALSE to regenerate slim model if needed

slim_rf_path <- here::here("results", "output", "rf_model_minimal.rds")
full_rf_path <- here::here("results", "large-files", "rf_tuned.rds")
full_rf_metrics_path <- here::here("results", "output", "rf_tuned_metrics.rds")
best_rf_params_path <- here::here("results", "output", "rf_best_params.rds")

if (use_slim_model) {
  if (!file.exists(slim_rf_path)) {
    stop("Slimmed model not found. Please run with use_slim_model <- FALSE first to generate it.")
  }

  message("Loading slimmed RF model...")
  rf_slim <- readRDS(slim_rf_path)
  rf_final_model <- rf_slim$model
  rf_metrics <- rf_slim$metrics
  rf_vip <- rf_slim$vip
  best_rf_params <- readRDS(best_rf_params_path)

} else {
  if (!file.exists(full_rf_path)) {
    stop("Full Random Forest model not found. Please rerun the tuning block.")
  }

  message("Loading full RF model and generating slim version...")
  rf_tuned <- readRDS(full_rf_path)
  rf_metrics <- readRDS(full_rf_metrics_path)
  rf_final_model <- extract_workflow(rf_tuned)
  rf_vip <- extract_fit_parsnip(rf_final_model) %>% vip::vi()
  best_rf_params <- readRDS(best_rf_params_path)

  if (!file.exists(slim_rf_path)) {
    saveRDS(
      list(model = rf_final_model, metrics = rf_metrics, vip = rf_vip),
      file = slim_rf_path
    )
    message("Slimmed model saved.")}}

```
Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.

The slim model can be toggled on or off here by changing from TRUE to FALSE if wanting to use the full version. The full RDS file is saved in the large-files folder which will not automatically upload to github.

```{r}
#Slim version already loads rf_metrics, so no need to reload
rf_metrics %>% filter(.metric == "rmse") %>% arrange(mean)

```

from this we see that the best model based on the lowest RMSE is the first one listed (Preprocessor1_Model6) with all 4 predictors used (mtry = 4), 15 observations per leaf in the tree ( min_n = 15), and RMSE = 115.22.

```{r}

# Plot RMSE across combinations
rf_metrics %>% filter(.metric == "rmse") %>%
  ggplot(aes(x = factor(mtry), y = mean, color = factor(min_n))) +
  geom_point(size = 3) +
  geom_line(aes(group = min_n)) +
  labs(title = "Random Forest Tuning Results (RMSE)",
       x = "mtry (Number of Predictors)",
       y = "Mean RMSE",
       color = "min_n") +
  theme_minimal()

```

```{r}
# File paths for saving
fit_path       <- here::here("results", "output", "final_rf_fit.rds")
workflow_path  <- here::here("results", "output", "final_rf_workflow.rds")
pred_path      <- here::here("results", "output", "rf_test_predictions.rds")
metrics_path   <- here::here("results", "output", "rf_test_metrics.rds")

# Skip if already saved
if (file.exists(fit_path) && file.exists(pred_path)) {
  message("Final model and predictions already saved. Skipping re-run...")
  final_rf_fit <- readRDS(fit_path)
  final_rf_workflow <- readRDS(workflow_path)
  rf_predictions <- readRDS(pred_path)
  rf_metrics <- readRDS(metrics_path)

} else {
  message(" Finalizing and fitting Random Forest model...")

  # Extract best parameters
  best_params <- best_rf_params

  # Finalize the workflow
  final_rf_workflow <- finalize_workflow(rf_workflow, best_params)

  # Fit final model on training set
  set.seed(123)
  final_rf_fit <- fit(final_rf_workflow, data = train_data)

  # Predict on test set
  rf_predictions <- predict(final_rf_fit, new_data = test_data) %>%
    bind_cols(test_data %>% select(mortalities))

  # Evaluate performance
  rf_metrics <- rf_predictions %>%
    metrics(truth = mortalities, estimate = .pred)

  # Save all results
  saveRDS(final_rf_fit, file = fit_path)
  saveRDS(final_rf_workflow, file = workflow_path)
  saveRDS(rf_predictions, file = pred_path)
  saveRDS(rf_metrics, file = metrics_path)

  message("Final model fit and results saved.")
}

# Review metrics
print(rf_metrics)

```

The model is performing well, RMSE and MAE both reflect reasonable prediction error magnitudes, with reasonable error and strong explanatory power.

```{r}

# Load saved RF metrics
rf_metrics <- readRDS(here::here("results", "output", "rf_test_metrics.rds"))

# Convert to wide format for easy table summary
rf_summary_table <- rf_metrics %>%
  select(.metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate) %>%
  mutate(
    Model = "Random Forest",
    Notes = "Tuned with 4 predictors (mtry = 4), min_n = 15"
  ) %>%
  select(Model, RMSE = rmse, R2 = rsq, MAE = mae, Notes)

# Save the summary table
saveRDS(rf_summary_table, here::here("results", "tables", "model_performance_summary.rds"))

rf_summary_table
```

```{r}
## Random Forest: Predicted vs Observed Plot

# Load predictions (if not already in memory)
rf_predictions <- readRDS(here::here("results", "output", "rf_test_predictions.rds"))

# Create the predicted vs observed plot
rf_pred_plot <- ggplot(rf_predictions, aes(x = mortalities, y = .pred)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Random Forest: Predicted vs Observed Mortality",
    x = "Observed Mortality",
    y = "Predicted Mortality"
  ) +
  theme_minimal()

# Save the plot to results/figures
ggsave(
  filename = here::here("results", "figures", "rf_pred_vs_obs.png"),
  plot = rf_pred_plot,
  width = 7, height = 5, dpi = 300)

rf_pred_plot
```

```{r}
## Random Forest: Variable Importance Plot

# Load fitted model (if not already in memory)
final_rf_fit <- readRDS(here::here("results", "output", "final_rf_fit.rds"))

# Create VIP plot
rf_vip_plot <- final_rf_fit %>%
  extract_fit_parsnip() %>%
  vip::vip(num_features = 10) +
  labs(title = "Top 10 Predictors (Random Forest)") +
  theme_minimal()

# Save the plot to results/figures
ggsave(
  filename = here::here("results", "figures", "rf_variable_importance.png"),
  plot = rf_vip_plot,
  width = 7)

rf_vip_plot
```

A Random Forest model was tuned and evaluated to predict county-level hypertension-related cardiovascular mortality. On the held-out test set, the model achieved an RMSE of 115.3, MAE of 66.9, and R² of 0.748, indicating that it explained approximately 75% of the variance in mortality rates across U.S. counties. These results reflect strong predictive performance and consistency with cross-validation metrics.

This model will serve as a benchmark against which we compare other algorithms, including regularized regression (e.g., LASSO).

####LASSO model

Define the LASSO model specification
```{r}
# LASSO uses glmnet engine with penalty tuning and fixed mixture = 1 (pure LASSO)
lasso_spec <- linear_reg(penalty = tune(), mixture = 1) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

#Create the LASSO workflow
# Workflow with recipe
lasso_workflow <- workflow() %>%
  add_model(lasso_spec) %>%
  add_recipe(mortality_recipe)

```

Tune the penalty using cross-validation

```{r}
## Tune LASSO penalty parameter

# File paths
lasso_tuned_path <- here::here("results", "large-files", "lasso_tuned.rds")
lasso_metrics_path <- here::here("results", "output", "lasso_tuned_metrics.rds")
lasso_best_param_path <- here::here("results", "output", "lasso_best_params.rds")

# Load or tune
if (file.exists(lasso_tuned_path) && file.exists(lasso_metrics_path) && file.exists(lasso_best_param_path)) {
  message("Loading saved LASSO tuning results...")

  lasso_tuned <- readRDS(lasso_tuned_path)
  lasso_tuned_metrics <- readRDS(lasso_metrics_path)
  best_lasso_params <- readRDS(lasso_best_param_path)
} else {
  message("Running LASSO tuning...")

  lasso_grid <- grid_regular(penalty(range = c(-4, -1)), levels = 10)  # log10 scale

  set.seed(123)
  lasso_tuned <- tune_grid(
    lasso_workflow,
    resamples = cv_folds,
    grid = lasso_grid,
    metrics = metric_set(rmse, rsq, mae),
    control = control_grid(save_pred = TRUE))

  lasso_tuned_metrics <- collect_metrics(lasso_tuned)
  best_lasso_params <- select_best(lasso_tuned, metric = "rmse")

  # Save results
  saveRDS(lasso_tuned, lasso_tuned_path)
  saveRDS(lasso_tuned_metrics, lasso_metrics_path)
  saveRDS(best_lasso_params, lasso_best_param_path)

  message("LASSO Tuning complete and results saved.")
}

# Preview metrics
print(lasso_metrics)
```

Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.
```{r}
## ---- slim_lasso_model_save, message=TRUE, warning=FALSE -------------------

slim_lasso_path <- here::here("results", "output", "lasso_model_minimal.rds")

if (!file.exists(slim_lasso_path)) {
  message("Creating slimmed LASSO model object...")

  lasso_final <- extract_workflow(lasso_tuned)
  lasso_metrics <- lasso_tuned_metrics

  # Extract coefficients
  lasso_fit <- extract_fit_parsnip(lasso_final)
  lasso_coeffs <- as.matrix(lasso_fit$fit$beta)

  saveRDS(
    list(model = lasso_final, metrics = lasso_metrics, coefficients = lasso_coeffs),
    file = slim_lasso_path
  )

  message("Saved slimmed LASSO model as lasso_model_minimal.rds")
} else {
  message("Slimmed LASSO model already exists. Skipping save.")
}

```

The slim model can be toggled on or off here by changing from TRUE to FALSE if wanting to use the full version. The full RDS file is saved in the large-files folder which will not automatically upload to github.
```{r}
## ---- load_lasso_model, message=TRUE, warning=FALSE -------------------

use_slim_lasso_model <- TRUE  # Toggle to FALSE to use full model

if (use_slim_lasso_model) {
  if (!file.exists(here::here("results", "output", "lasso_model_minimal.rds"))) {
    stop("Slim LASSO model not found. Please run the slim-save chunk or disable slim mode.")
  }

  message("Loading slimmed LASSO model...")
  lasso_slim <- readRDS(here::here("results", "output", "lasso_model_minimal.rds"))

  lasso_final_model <- lasso_slim$model
  lasso_metrics <- lasso_slim$metrics
  lasso_coeffs <- lasso_slim$coefficients
  best_lasso_params <- readRDS(here::here("results", "output", "lasso_best_params.rds"))

} else {
  if (!file.exists(here::here("results", "large-files", "lasso_tuned.rds"))) {
    stop("Full LASSO model not found.")
  }

  message("Loading full LASSO model...")
  lasso_tuned <- readRDS(here::here("results", "large-files", "lasso_tuned.rds"))
  lasso_metrics <- readRDS(here::here("results", "output", "lasso_tuned_metrics.rds"))
  lasso_final_model <- extract_workflow(lasso_tuned)

  lasso_fit <- extract_fit_parsnip(lasso_final_model)
  lasso_coeffs <- as.matrix(lasso_fit$fit$beta)

  best_lasso_params <- readRDS(here::here("results", "output", "lasso_best_params.rds"))}

```

```{r}
#Plot LASSO Tuning Plot
# Plot RMSE vs penalty (log10 scale)
lasso_tune_plot <- lasso_metrics %>%
  filter(.metric == "rmse") %>%
  ggplot(aes(x = penalty, y = mean)) +
  geom_point(size = 2, color = "steelblue") +
  geom_line() +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "LASSO Tuning Results (RMSE vs Penalty)",
       x = "Penalty (log10 scale)",
       y = "Mean RMSE")

# Save to results/figures
ggsave(filename = here::here("results", "figures", "lasso_tuning_plot.png"),
  plot = lasso_tune_plot,
  width = 7, height = 5, dpi = 300)

lasso_tune_plot
```

```{r}
## Final LASSO Model Fit and Evaluation

# Output paths
lasso_fit_path <- here::here("results", "output", "final_lasso_fit.rds")
lasso_workflow_path <- here::here("results", "output", "final_lasso_workflow.rds")
lasso_pred_path <- here::here("results", "output", "lasso_test_predictions.rds")
lasso_metrics_path <- here::here("results", "output", "lasso_test_metrics.rds")

if (file.exists(lasso_fit_path) &&
    file.exists(lasso_workflow_path) &&
    file.exists(lasso_pred_path) &&
    file.exists(lasso_metrics_path)) {

  message("Loading saved LASSO model and outputs...")
  final_lasso_fit <- readRDS(lasso_fit_path)
  final_lasso_workflow <- readRDS(lasso_workflow_path)
  lasso_test_predictions <- readRDS(lasso_pred_path)
  lasso_test_metrics <- readRDS(lasso_metrics_path)

} else {
  # Finalize model
  final_lasso_model <- finalize_model(lasso_spec, best_lasso_params)

  # Build workflow
  final_lasso_workflow <- workflow() %>%
    add_model(final_lasso_model) %>%
    add_recipe(mortality_recipe)

  # Fit and predict
  final_lasso_fit <- fit(final_lasso_workflow, data = train_data)

  lasso_test_predictions <- predict(final_lasso_fit, new_data = test_data) %>%
    bind_cols(test_data %>% select(mortalities))

  # Evaluate
  lasso_test_metrics <- lasso_test_predictions %>%
    metrics(truth = mortalities, estimate = .pred)

  # Save results
  saveRDS(final_lasso_fit, lasso_fit_path)
  saveRDS(final_lasso_workflow, lasso_workflow_path)
  saveRDS(lasso_test_predictions, lasso_pred_path)
  saveRDS(lasso_test_metrics, lasso_metrics_path)
}

# Print metrics
print(lasso_test_metrics)
```

```{r}
## LASSO: Predicted vs Observed Plot

# Load predictions (if not already in memory)
lasso_test_predictions <- readRDS(here::here("results", "output", "lasso_test_predictions.rds"))

lasso_pred_plot <- ggplot(lasso_test_predictions, aes(x = mortalities, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "LASSO Regression: Predicted vs Observed Mortality",
    x = "Observed Mortality",
    y = "Predicted Mortality"
  ) +
  theme_minimal()

# Save to file
ggsave(
  filename = here::here("results", "figures", "lasso_pred_vs_obs.png"),
  plot = lasso_pred_plot,
  width = 7, height = 5, dpi = 300)

lasso_pred_plot
```

```{r}
lasso_summary_table <- lasso_test_metrics %>%
  select(.metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate) %>%
  mutate(
    Model = "LASSO Regression",
    Notes = paste("Penalty:", signif(best_lasso_params$penalty, 3))
  ) %>%
  select(Model, RMSE = rmse, R2 = rsq, MAE = mae, Notes)

saveRDS(lasso_summary_table, here::here("results", "tables", "lasso_model_summary.rds"))

lasso_summary_table
```

The LASSO regression model achieved a Root Mean Squared Error (RMSE) of ~119.63, with an R-squared of ~0.726, indicating a reasonably strong fit for predicting county-level hypertension-related mortality. Compared to the random forest model (RMSE ~115.30, R² ~0.748), LASSO performs slightly worse, but still retains good explanatory power.

Notably, LASSO applies regularization and performs variable selection, meaning it can identify the most influential predictors while shrinking the less important ones to zero. This makes it a useful, interpretable model for identifying risk factors — though it sacrifices some predictive accuracy compared to Random Forest.

Plot Predicted vs Observed Mortality

```{r}
ggplot(lasso_test_predictions, aes(x = mortalities, y = .pred)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "LASSO Regression: Predicted vs Observed Mortality",
    x = "Observed Mortality",
    y = "Predicted Mortality"
  ) +
  theme_minimal(base_size = 13)

```

LASSO is a linear model with regularization — it struggles to capture high-order nonlinear patterns. The horizontal “bands” in predicted values suggest that LASSO may be underfitting in high or low mortality counties. While LASSO remains interpretable, random forest outperforms it in predictive power and flexibility, especially in a dataset with many categorical and potentially non-linear features.

###Comparing models

Generating a model comparison table:

```{r}

## Generate model comparison table with 95% confidence intervals

# Load test metrics
rf_metrics <- readRDS(here::here("results", "output", "rf_test_metrics.rds"))
lasso_metrics <- readRDS(here::here("results", "output", "lasso_test_metrics.rds"))

# Add model names
rf_metrics$model <- "Random Forest"
lasso_metrics$model <- "LASSO Regression"

# Combine
all_metrics <- bind_rows(rf_metrics, lasso_metrics)

# Check if std_err is available
has_se <- "std_err" %in% names(all_metrics)

# Define function to create CI string
format_ci <- function(mean, se = NA) {
  if (is.na(se) || se == 0) return(as.character(round(mean, 2)))
  lower <- round(mean - 1.96 * se, 2)
  upper <- round(mean + 1.96 * se, 2)
  paste0(round(mean, 2), " [", lower, ", ", upper, "]")
}

# Create final table
if (has_se) {
  # Use CI formatting
  all_metrics_ci <- all_metrics %>%
    select(model, .metric, mean = .estimate, std_err) %>%
    mutate(ci = mapply(format_ci, mean, std_err)) %>%
    select(model, .metric, ci) %>%
    pivot_wider(names_from = .metric, values_from = ci)
} else {
  # Fall back to just point estimates
  all_metrics_ci <- all_metrics %>%
    select(model, .metric, mean = .estimate) %>%
    mutate(ci = round(mean, 2)) %>%
    select(model, .metric, ci) %>%
    pivot_wider(names_from = .metric, values_from = ci)
}

# Build gt table
model_perf_ci_table <- all_metrics_ci %>%
  gt() %>%
  tab_header(
    title = "Model Performance Summary",
    subtitle = "Test Set Evaluation for Random Forest and LASSO"
  ) %>%
  cols_label(
    model = "Model",
    rmse = "RMSE",
    rsq = "R²",
    mae = "MAE"
  ) %>%
  fmt_markdown(columns = everything()) %>%
  tab_options(
    table.font.size = px(12),
    heading.title.font.size = px(14),
    heading.subtitle.font.size = px(12),
    column_labels.font.weight = "bold"
  )

# Save table
saveRDS(model_perf_ci_table, here::here("results", "tables", "model_performance_summary.rds"))

# View it
model_perf_ci_table

```

Additional comparison table without CIs:

```{r}
model_perf_numeric <- bind_rows(rf_metrics, lasso_metrics) %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)

saveRDS(model_perf_numeric, here::here("results", "output", "model_performance_numeric.rds"))

model_perf_numeric
```

Aim 4 Summary
We compared two predictive models for county-level hypertension-related CVD mortality: a random forest model and a LASSO regression model. The random forest achieved the best performance, with an RMSE of 115.3 and R² of 0.748 on the test set. LASSO regression performed slightly worse (RMSE = 119.6, R² = 0.726) but offers the benefit of interpretability through feature selection. These results indicate strong overall predictive performance, with random forest serving as the best-performing model. Both models demonstrated solid predictive performance, with random forest performing best. The structured pattern seen in the predicted vs observed plots reflects the nature of the input data — with many counties having identical values for categorical groupings. The use of dummy variables, while necessary, introduces constraints that can create repeated predictions for common demographic combinations, especially in linear models like LASSO. This highlights a trade-off between model interpretability and flexibility.

### 2025-04-11 code/analysis-code/statistical-analysis-v2.qmd
Part 4 additions

--- 2025-04-11 code/analysis-code/statistical-analysis-v2-slim-models.qmd
@@
 ```{r}
+
 # GLM with 3-way interaction to explore mortality trends by Year * race_ethnicity * sex, stratified by age group
-
-## ---- load_glm_race_sex_age_model, message=TRUE, warning=FALSE -------------------
-
-glm_path <- here::here("results", "output", "glm_race_sex_age_model.rds")
-
-if (file.exists(glm_path)) {
-  message("Loading saved GLM model with 3-way interaction...")
-  glm_race_sex_age <- readRDS(glm_path)
-} else {
-  message("Fitting GLM model with 3-way interaction...")
 glm_race_sex_age <- glm(mortalities ~ Year * race_ethnicity * sex + age_group,
-                          data = county_race_sex, family = gaussian())
-  saveRDS(glm_race_sex_age, glm_path)
-}
-
-#glm_race_sex_age <- glm(mortalities ~ Year * race_ethnicity * sex + age_group, data = processed_data_no_percent, family = gaussian())
+                        data = processed_data_no_percent,
+                        family = gaussian())
 
 # Save model
-#saveRDS(glm_race_sex_age, here::here("results", "output", "glm_race_sex_age_model.rds"))
+saveRDS(glm_race_sex_age, here::here("results", "output", "glm_race_sex_age_model.rds"))
 
@@
 # Define file paths for saving/loading tuning results
-tuned_path     <- here::here("results", "large-files", "rf_tuned.rds")
+tuned_path     <- here::here("results", "output", "rf_tuned.rds")
 metrics_path   <- here::here("results", "output", "rf_tuned_metrics.rds")
@@
 
-  message("Tuning complete and results saved.")}
-
-```
-```{r}
-## ---- fit_or_load_rf_model, message=TRUE, warning=FALSE -------------------
-
-use_slim_model <- FALSE # Toggle this to FALSE to regenerate slim model if needed
-
-slim_rf_path <- here::here("results", "output", "rf_model_minimal.rds")
-full_rf_path <- here::here("results", "large-files", "rf_tuned.rds")
-full_rf_metrics_path <- here::here("results", "output", "rf_tuned_metrics.rds")
-best_rf_params_path <- here::here("results", "output", "rf_best_params.rds")
-
-if (use_slim_model) {
-  if (!file.exists(slim_rf_path)) {
-    stop("Slimmed model not found. Please run with use_slim_model <- FALSE first to generate it.")
+  message("Tuning complete and results saved.")
 }
 
-  message("Loading slimmed RF model...")
-  rf_slim <- readRDS(slim_rf_path)
-  rf_final_model <- rf_slim$model
-  rf_metrics <- rf_slim$metrics
-  rf_vip <- rf_slim$vip
-  best_rf_params <- readRDS(best_rf_params_path)
-
-} else {
-  if (!file.exists(full_rf_path)) {
-    stop("Full Random Forest model not found. Please rerun the tuning block.")
-  }
-
-  message("Loading full RF model and generating slim version...")
-  rf_tuned <- readRDS(full_rf_path)
-  rf_metrics <- readRDS(full_rf_metrics_path)
-  rf_final_model <- extract_workflow(rf_tuned)
-  rf_vip <- extract_fit_parsnip(rf_final_model) %>% vip::vi()
-  best_rf_params <- readRDS(best_rf_params_path)
-
-  if (!file.exists(slim_rf_path)) {
-    saveRDS(
-      list(model = rf_final_model, metrics = rf_metrics, vip = rf_vip),
-      file = slim_rf_path
-    )
-    message("Slimmed model saved.")}}
-
-```
-Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.
-
-The slim model can be toggled on or off here by changing from TRUE to FALSE if wanting to use the full version.
-The full RDS file is saved in the large-files folder which will not automatically upload to github.
-
-```{r}
-#Slim version already loads rf_metrics, so no need to reload
-rf_metrics %>% filter(.metric == "rmse") %>% arrange(mean)
+# Preview metrics
+print(rf_tuned_metrics)
+
+```
+
+```{r}
+
+# Load metrics (if not already loaded)
+rf_tuned_metrics <- readRDS(here::here("results", "output", "rf_tuned_metrics.rds"))
+
+# Sort by RMSE to find best performing combo
+rf_tuned_metrics %>% filter(.metric == "rmse") %>% arrange(mean)
 
@@
 # Plot RMSE across combinations
-rf_metrics %>% filter(.metric == "rmse") %>%
+rf_tuned_metrics %>%
+  filter(.metric == "rmse") %>%
   ggplot(aes(x = factor(mtry), y = mean, color = factor(min_n))) +
@@
   # Extract best parameters
-  best_params <- best_rf_params
+  best_params <- select_best(rf_tuned, metric = "rmse")
 
@@
 # File paths
-lasso_tuned_path <- here::here("results", "large-files", "lasso_tuned.rds")
+lasso_tuned_path <- here::here("results", "output", "lasso_tuned.rds")
 lasso_metrics_path <- here::here("results", "output", "lasso_tuned_metrics.rds")
-lasso_best_param_path <- here::here("results", "output", "lasso_best_params.rds")
+lasso_best_path <- here::here("results", "output", "lasso_best_params.rds")
 
 # Load or tune
-if (file.exists(lasso_tuned_path) && file.exists(lasso_metrics_path) && file.exists(lasso_best_param_path)) {
+if (file.exists(lasso_tuned_path) &&
+    file.exists(lasso_metrics_path) &&
+    file.exists(lasso_best_path)) {
   message("Loading saved LASSO tuning results...")
-
   lasso_tuned <- readRDS(lasso_tuned_path)
   lasso_tuned_metrics <- readRDS(lasso_metrics_path)
-  best_lasso_params <- readRDS(lasso_best_param_path)
+  best_lasso_params <- readRDS(lasso_best_path)
 } else {
@@
     metrics = metric_set(rmse, rsq, mae),
-    control = control_grid(save_pred = TRUE))
+    control = control_grid(save_pred = TRUE)
+  )
 
@@
   saveRDS(lasso_tuned_metrics, lasso_metrics_path)
-  saveRDS(best_lasso_params, lasso_best_param_path)
-
-  message("LASSO Tuning complete and results saved.")
+  saveRDS(best_lasso_params, lasso_best_path)
+
+  message("Tuning complete and results saved.")
 }
@@
 # Preview metrics
-print(lasso_metrics)
-```
-
-Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.
-```{r}
-## ---- slim_lasso_model_save, message=TRUE, warning=FALSE -------------------
-
-slim_lasso_path <- here::here("results", "output", "lasso_model_minimal.rds")
-
-if (!file.exists(slim_lasso_path)) {
-  message("Creating slimmed LASSO model object...")
-
-  lasso_final <- extract_workflow(lasso_tuned)
-  lasso_metrics <- lasso_tuned_metrics
-
-  # Extract coefficients
-  lasso_fit <- extract_fit_parsnip(lasso_final)
-  lasso_coeffs <- as.matrix(lasso_fit$fit$beta)
-
-  saveRDS(
-    list(model = lasso_final, metrics = lasso_metrics, coefficients = lasso_coeffs),
-    file = slim_lasso_path
-  )
-
-  message("Saved slimmed LASSO model as lasso_model_minimal.rds")
-} else {
-  message("Slimmed LASSO model already exists. Skipping save.")
-}
-
-```
-
-The slim model can be toggled on or off here by changing from TRUE to FALSE if wanting to use the full version.
-The full RDS file is saved in the large-files folder which will not automatically upload to github.
-```{r}
-## ---- load_lasso_model, message=TRUE, warning=FALSE -------------------
-
-use_slim_lasso_model <- TRUE  # Toggle to FALSE to use full model
-
-if (use_slim_lasso_model) {
-  if (!file.exists(here::here("results", "output", "lasso_model_minimal.rds"))) {
-    stop("Slim LASSO model not found. Please run the slim-save chunk or disable slim mode.")
-  }
-
-  message("Loading slimmed LASSO model...")
-  lasso_slim <- readRDS(here::here("results", "output", "lasso_model_minimal.rds"))
-
-  lasso_final_model <- lasso_slim$model
-  lasso_metrics <- lasso_slim$metrics
-  lasso_coeffs <- lasso_slim$coefficients
-  best_lasso_params <- readRDS(here::here("results", "output", "lasso_best_params.rds"))
-
-} else {
-  if (!file.exists(here::here("results", "large-files", "lasso_tuned.rds"))) {
-    stop("Full LASSO model not found.")
-  }
-
-  message("Loading full LASSO model...")
-  lasso_tuned <- readRDS(here::here("results", "large-files", "lasso_tuned.rds"))
-  lasso_metrics <- readRDS(here::here("results", "output", "lasso_tuned_metrics.rds"))
-  lasso_final_model <- extract_workflow(lasso_tuned)
-
-  lasso_fit <- extract_fit_parsnip(lasso_final_model)
-  lasso_coeffs <- as.matrix(lasso_fit$fit$beta)
-
-  best_lasso_params <- readRDS(here::here("results", "output", "lasso_best_params.rds"))}
-
+print(lasso_tuned_metrics)
 ```
@@
 # Plot RMSE vs penalty (log10 scale)
-lasso_tune_plot <- lasso_metrics %>%
+lasso_tune_plot <- lasso_tuned_metrics %>%
   filter(.metric == "rmse") %>%

### 2025-04-11 code/eda-code/eda.qmd
Part 4 additions

---
title: "CDC hypertension and CDV Mortality disease data from adults in the USA exploratory analysis script"
date: "2025-02-20"
output: html_document
---

# Setup

```{r}
#load needed packages. make sure they are installed.
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set saving/loading paths
library(knitr) #for nice tables and data integration
library(corrplot) #for correlation matrix visualization
library(sf) #for geospatial analysis
library(ggpubr) #for advanced visualization tools
library(ggplot2) #for plots
library(gt) #For a tidy table

# Install if needed: 
#install.packages("webshot")
# webshot::install_phantomjs()
```

Load the data.

```{r}
#Path to data. 
data_location <- here::here("data","processed-data","final_processed_data.rds")

#load data. 
final_processed_data <- readRDS(data_location)

# Extract key datasets
processed_data_no_percent <- final_processed_data[["processed_data_no_percent"]]
county_race_sex <- final_processed_data[["county_race_sex"]]
county_race <- final_processed_data[["county_race"]]
county_sex <- final_processed_data[["county_sex"]]
county_overall <- final_processed_data[["county_overall"]]

```

Running some EDA and checking summaries of data:

We are going to make use of loops and creating a list of the different datasets that we created to reduce separate code to run the descriptive statisitics for each dataset. We are not include the percent change dataset in these evaluations.

Creating this datset list:
```{r}
list_datasets <- list(
  "processed_data_no_percent" = processed_data_no_percent,
  "County Race-Sex" = county_race_sex,
  "County Race" = county_race,
  "County Sex" = county_sex,
  "County Overall" = county_overall)

```

We will run the descriptive statistics for each dataset (as a single loop)
```{r}
# Loop through each dataset and print basic statistics
summary_stats <- bind_rows(lapply(names(list_datasets), function(name) {
  stats <- list_datasets[[name]] %>%
    summarise(
      mean = mean(mortalities, na.rm = TRUE),
      median = median(mortalities, na.rm = TRUE),
      sd = sd(mortalities, na.rm = TRUE),
      min = min(mortalities, na.rm = TRUE),
      max = max(mortalities, na.rm = TRUE))
  stats$Dataset <- name  # Add dataset name for identification
  return(stats)}))

# Print summary statistics
tibble::as_tibble(summary_stats) %>% 
  relocate(Dataset) %>% 
  print()
```

# Identify potential missing values in each dataset
```{r}
cat("\n### Missing Data Summary:\n")
lapply(names(list_datasets), function(name) {
  missing_count <- sum(is.na(list_datasets[[name]]))
  cat(name, "has", missing_count, "missing values.\n")
})
```

We want to visualise the different datasets:

We will revisit the histogram with the percent change removed.
```{r}
# Visualize the distribution of mortality rates with a histogram
revised_mortalities_histogram <- ggplot(processed_data_no_percent, aes(x = mortalities)) +
  geom_histogram(binwidth = 20, fill = "blue", alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, max(processed_data_no_percent$mortalities, na.rm = TRUE), by = 200)) +
  theme_minimal() +
  labs(title = "Distribution of Mortality Rates", x = "Mortality Rates", y = "Count")

saveRDS(revised_mortalities_histogram, file = here::here("data","processed-data", "revised_mortalities_histogram.rds"))
revised_mortalities_histogram <- readRDS(here::here("data","processed-data", "revised_mortalities_histogram.rds"))
# Print the mortalities histogram
print(revised_mortalities_histogram)
```

This is the total dataset and so the data is similar to the previous histogram. It is bimodal and there is a strong right skew and long tail. Peak mortalities are below 100, and we are not sure what accounts for the second peak.
This might be a factor of having two different age groups. We will see if this is made clear by splitting the age groups.

We want to split the data into two age groups one for individuals from 35 - 64 years, and the second for those 65 years and older. 
Age group 1 = 35 -64 years old
Age groups 2 = 65 years and older
```{r}
# Split into two age groups
age_group_1 <- processed_data_no_percent %>% filter(age_group == "Ages 35-64 years")
age_group_2 <- processed_data_no_percent %>% filter(age_group == "Ages 65+ years")

# Optional: Save these for later
saveRDS(age_group_1, here::here("data", "processed-data", "age_group_1.rds"))
saveRDS(age_group_2, here::here("data", "processed-data", "age_group_2.rds"))

```

We will recreate the histogram (one for each group) to check if we see distinct differences.

```{r}
# Histogram for Age Group 1 (35–64 years)
hist_age_group_1 <- ggplot(age_group_1, aes(x = mortalities)) +
  geom_histogram(fill = "steelblue", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Distribution of Mortality Rates (Ages 35–64)",
    x = "Mortality Rate",
    y = "Count")

# Save new histo variable
saveRDS(hist_age_group_1, here::here("data", "processed-data", "hist_age_group_1.rds"))
#Print histogram
print(hist_age_group_1)

# Histogram for Age Group 2 (65+ years)
hist_age_group_2 <- ggplot(age_group_2, aes(x = mortalities)) +
  geom_histogram(binwidth = 20, fill = "darkorange", alpha = 0.7) +
  scale_x_continuous(breaks = seq(0, max(age_group_2$mortalities, na.rm = TRUE), by = 200)) +
  theme_minimal() +
  labs( title = "Distribution of Mortality Rates (Ages 65+)",
    x = "Mortality Rate",
    y = "Count")

# Save and optionally reload
saveRDS(hist_age_group_2, here::here("data", "processed-data", "hist_age_group_2.rds"))
# Print histogram 2
print(hist_age_group_2)

```

We will use a boxplot to get an overview of the other data:

```{r}
# Compare mortality rates across race-sex and sex stratifications

boxplot_race_sex <- ggplot(county_race_sex, aes(x = race_ethnicity, y = mortalities, fill = sex)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Mortality Rates by Race and Sex", x = "Race/Ethnicity", y = "Mortality Rate")

saveRDS(boxplot_race_sex, file = here::here("data","processed-data", "boxplot_race_sex.rds"))
print(boxplot_race_sex)

boxplot_sex <- ggplot(county_sex, aes(x = sex, y = mortalities)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Mortality Rates by Sex", x = "Sex", y = "Mortality Rate")

saveRDS(boxplot_sex, file = here::here("data","processed-data", "boxplot_sex.rds"))
print(boxplot_sex)

```

The box plots provide a good overview of the data nd indicate that there seem to be distinct trends and differences between races. 
Mortalities in the black and white race groups appear to be the highest, and there are many outliers in the white race group.
It also shows that there are differences between sexes but these are less marked and distinct than between races.
We will look at this in the statistical analysis.

Neatening up the boxplot for race without sex
```{r}
# Relationship between mortality and race
race_plot <- ggplot(county_race, aes(x = race_ethnicity, y = mortalities)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Mortality Rates by Race/Ethnicity", x = "Race/Ethnicity", y = "Mortality Rate") +
  theme(axis.text.x = element_text(size = 10)) +
  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10))

ggsave(here::here("results", "figures", "mortality_by_race.png"), plot = race_plot)
print(race_plot)

```
This is the same but without sex and the colours for a neater output.

We want an idea on any spatial trends in the data. We are using the sf function to plot the mortality data by the latitude and longitude.
```{r}
# Convert data to spatial format and create a map of mortality rates
# Ensure longitude and latitude values are numeric
processed_data_no_percent <- processed_data_no_percent %>%
  mutate(X_long = as.numeric(X_long), Y_lat = as.numeric(Y_lat))
# Convert to spatial object
sf_data <- st_as_sf(processed_data_no_percent, coords = c("X_long", "Y_lat"), crs = 4326)

# Generate a spatial plot of mortality rates
ggplot(sf_data) +
  geom_sf(aes(color = mortalities)) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Geospatial Distribution of Mortality Rates")

```

This is a very nice overview and can help us to evauate the trends further or to work on a more intricate plot. We must remember that we have not separated groups (especially age) yet.
But we do see that there is a lot more data from the East and SouthEast. Although it is possible that there might be limited data from some States on the West.

We want an overview of any temporal (time based) trends in the data. We will provide a proof of concept using a single county due to the overlap of data that may occur when plotting many different counties together. The statistical analysis will include further refinements. We are only working with the values where sex and race were both overall to reduce overlap.
```{r}
# Line plot of mortality rates over time
# Select a single county to illustrate the trend
selected_county <- "13037"  # County Location ID for Calhoun, GA
county_subset <- county_overall %>% filter(LocationID == selected_county)

temporal_trends_single_county <- ggplot(county_subset, aes(x = Year, y = mortalities, group = LocationID)) +
  geom_line(alpha = 0.6, color = "blue") +
  geom_smooth(se = FALSE, color = "red") +
  theme_minimal() +
  labs(title = paste("Mortality Rate Trends in Calhoun"), 
       x = "Year", y = "Mortality Rate")

print(temporal_trends_single_county)

```

This is a very strange plot and highlights the issues when combining different age groups together.
We will separate the age groups and run this again.

```{r}
# Use age-specific data
selected_county_data <- processed_data_no_percent %>%
  filter(LocationID == "13037",
         race_ethnicity == "Overall",
         sex == "Overall")

ggplot(selected_county_data, aes(x = Year, y = mortalities, color = age_group)) +
  geom_line(alpha = 0.6) +
  geom_smooth(se = FALSE) +
  theme_minimal() +
  labs(
    title = "Mortality Rate Trends by Age Group in Calhoun (Proof of Concept)",
    x = "Year", y = "Mortality Rate", color = "Age Group")

```

We see a clear increase in mortality rates in both age groups over time, but more so in the age group for 65+.

We plot the overall national trend over time with the two different age groups.
```{r}
# National mortality trend by age group
national_age_trend <- processed_data_no_percent %>%
  filter(race_ethnicity == "Overall", sex == "Overall") %>%
  group_by(Year, age_group) %>%
  summarise(mean_mortality = mean(mortalities, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = mean_mortality, color = age_group)) +
  geom_line(size = 1) +
  geom_smooth(se = FALSE) +
  theme_minimal() +
  labs(title = "National Mortality Trends by Age Group",
       x = "Year", y = "Mean Mortality Rate", color = "Age Group")

saveRDS(national_age_trend, here::here("data", "processed-data", "national_trend_by_age.rds"))
print(national_age_trend)
```

```{r}
# National mortality trend by race
race_trend_65_plus <- processed_data_no_percent %>%
  filter(age_group == "Ages 65+ years", sex == "Overall") %>%
  group_by(Year, race_ethnicity) %>%
  summarise(mean_mortality = mean(mortalities, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = mean_mortality, color = race_ethnicity)) +
  geom_line(size = 1) +
  geom_smooth(se = FALSE) +
  theme_minimal() +
  labs(title = "Mortality Trends by Race/Ethnicity (65+)",
       x = "Year", y = "Mean Mortality Rate", color = "Race")

saveRDS(race_trend_65_plus, here::here("data", "processed-data", "race_trend_65_plus.rds"))
print(race_trend_65_plus)

```

```{r}

race_trend_35_64 <- processed_data_no_percent %>%
  filter(age_group == "Ages 35-64 years", sex == "Overall") %>%
  group_by(Year, race_ethnicity) %>%
  summarise(mean_mortality = mean(mortalities, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = mean_mortality, color = race_ethnicity)) +
  geom_line(size = 1) +
  theme_minimal() +
  labs(title = "Mortality Trends by Race/Ethnicity (Ages 35–64)",
       x = "Year", y = "Mean Mortality Rate", color = "Race")

saveRDS(race_trend_35_64, here::here("data", "processed-data", "race_trend_35_64.rds"))
print(race_trend_35_64)

```

```{r}
# National mortality trend by sex
sex_trend_65_plus <- processed_data_no_percent %>%
  filter(race_ethnicity == "Overall", age_group == "Ages 65+ years") %>%
  group_by(Year, sex) %>%
  summarise(mean_mortality = mean(mortalities, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = mean_mortality, color = sex)) +
  geom_line(size = 1) +
  geom_smooth(se = FALSE) +
  theme_minimal() +
  labs(title = "Mortality Trends by Sex (65+)",
       x = "Year", y = "Mean Mortality Rate", color = "Sex")

saveRDS(sex_trend_65_plus, here::here("data", "processed-data", "sex_trend_65_plus.rds"))
print(sex_trend_65_plus)

```

```{r}

sex_trend_35_64 <- processed_data_no_percent %>%
  filter(age_group == "Ages 35-64 years", race_ethnicity == "Overall") %>%
  group_by(Year, sex) %>%
  summarise(mean_mortality = mean(mortalities, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = mean_mortality, color = sex)) +
  geom_line(size = 1) +
  theme_minimal() +
  labs(title = "Mortality Trends by Sex (Ages 35–64)",
       x = "Year", y = "Mean Mortality Rate", color = "Sex")

saveRDS(sex_trend_35_64, here::here("data", "processed-data", "sex_trend_35_64.rds"))
print(sex_trend_35_64)

```

```{r}
# Grouped summary statistics
grouped_summary <- processed_data_no_percent %>%
  group_by(age_group, race_ethnicity, sex) %>%
  summarise( mean_mortality = round(mean(mortalities, na.rm = TRUE), 2),
    median_mortality = round(median(mortalities, na.rm = TRUE), 2),
    sd_mortality = round(sd(mortalities, na.rm = TRUE), 2),
    min_mortality = min(mortalities, na.rm = TRUE),
    max_mortality = max(mortalities, na.rm = TRUE),
    n = n(),.groups = "drop")

# Print summary as a nice table
grouped_summary %>% knitr::kable(caption = "Summary of Mortality by Age, Race, and Sex")

```

```{r}
# Define custom sorting levels
race_order <- c(
  "American Indian and Alaska Native",
  "Asian and Pacific Islander",
  "Black",
  "Hispanic",
  "White",
  "Overall")

sex_order <- c("Female", "Male", "Overall")

# Apply sorting and formatting
grouped_summary_clean <- grouped_summary %>%
  mutate(
     race_ethnicity = factor(race_ethnicity, levels = race_order),
    sex = factor(sex, levels = sex_order)
  ) %>%
  arrange(age_group, race_ethnicity, sex) %>%
  rename(
    `Age Group` = age_group,
    `Race/Ethnicity` = race_ethnicity,
    `Sex` = sex,
    `Mean Mortality` = mean_mortality,
    `Median Mortality` = median_mortality,
    `SD` = sd_mortality,
    `Minimum` = min_mortality,
    `Maximum` = max_mortality,
    `n` = n) %>%
  gt() %>%
  tab_header(
    title = "Summary of Mortality Rates by Age, Race, and Sex"
  ) %>%
  fmt_number(
    columns = c(`Mean Mortality`, `Median Mortality`, `SD`, `Minimum`, `Maximum`),
    decimals = 1
  ) %>%
  cols_align(align = "center") %>%
  tab_options(
    table.font.size = "small",
    heading.title.font.size = 14,
    data_row.padding = px(2))

gtsave(grouped_summary_clean, here::here("results", "tables", "summary_mortality_table.png"))

# View it
grouped_summary_clean

```

### Preliminary Linear Regression to Assess National Trends

```{r}
# This simple linear model estimates overall change in mortality over time.
# It helps confirm if a temporal trend exists before stratified modeling.

# Exploring mean mortality by year
yearly_means <- processed_data_no_percent %>%
  filter(race_ethnicity == "Overall", sex == "Overall") %>%
  group_by(Year) %>%
  summarise(mean_mortality = mean(mortalities, na.rm = TRUE))

ggplot(yearly_means, aes(x = Year, y = mean_mortality)) +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Trend in Mean Mortality Rates Over Time ",
       x = "Year", y = "Mean Mortality Rate") +
  theme_minimal()

saveRDS(yearly_means, here::here("data", "processed-data", "yearly_means.rds"))
```

This model supports what the plots indicated that there are strong trends in the data indicating an increase in the mortality rates over time.

```{r}
eda_outputs <- list(
  processed_data_no_percent = processed_data_no_percent,
  age_group_1 = age_group_1,
  age_group_2 = age_group_2,
  grouped_summary = grouped_summary,
  county_race_sex = county_race_sex,
  county_race = county_race,
  county_sex = county_sex,
  county_overall = county_overall)

saveRDS(eda_outputs, here::here("data", "processed-data", "eda_outputs.rds"))

```

## Key Findings and Next Steps

- The distribution of mortality rates is right-skewed, suggesting potential transformations before modeling.
- Mortality rates vary markedly by race and sex, justifying stratified analysis.
- A proof-of-concept temporal trend was created for a single county (Calhoun, GA) using Location ID, ensuring unique selection.
- Temporal trend indicate a general increase the selected county, which should be modeled.
- All processed data has been saved into a single RDS file (`eda_outputs.rds`) for use in statistical analysis.

- Temporal and spatial variation in the data highlights the need for multilevel or mixed-effects modeling in the statistical analysis phase.
- The grouped summary by age, race, and sex provides key descriptive benchmarks and will support interpretation of model estimates.
- Notable outliers were observed in the White and Black race groups, which may warrant robustness checks or sensitivity analysis.

### 2025-04-11 code/processing-code/processingfile-v1.qmd
Part 4 additions

@@
 ---
-title: "Processing CDC hypertension and CDV Mortality disease data from adults in the USA"
+title: "Processing CDC hypertension and CDV Mortality disease data from adults in the USA - Processing script"
 author: "Shaun van den Hurk"
@@
   filter(sex == "Overall", race_ethnicity == "Overall") %>% # Keep only total county-level data
-  select(LocationID, LocationDesc, Year, mortalities, X_long, Y_lat, LocationAbbr, Data_Value_Unit, Data_Value_Type, Confidence_limit_Low, Confidence_limit_High, age_group)
+  select(LocationID,Year,LocationAbbr, LocationDesc,  mortalities, Data_Value_Unit, Data_Value_Type, Confidence_limit_Low, Confidence_limit_High, age_group, X_long, Y_lat)
 
@@
 
-Running some EDA and checking summaries of data:
+## Summary and Next Steps
 
-We are going to make use of loops and creating a list of the different datasets that we created to reduce separate code to run the descriptive statisitics for each dataset.
-We are not include the percent change dataset in these evaluations.
-
-Creating this datset list:
-```{r}
-list_datasets <- list(
-  "Full Dataset" = hypertens_adult_filtered_no_percent,
-  "County Race-Sex" = county_race_sex,
-  "County Race" = county_race,
-  "County Sex" = county_sex,
-  "County Overall" = county_overall
-)
-
-```
-
-We will run the descriptive statistics for each dataset (as a single loop)
-```{r}
-# Loop through each dataset and print basic statistics
-summary_stats <- bind_rows(lapply(names(list_datasets), function(name) {
-  stats <- list_datasets[[name]] %>%
-    summarise(
-      mean = mean(mortalities, na.rm = TRUE),
-      median = median(mortalities, na.rm = TRUE),
-      sd = sd(mortalities, na.rm = TRUE),
-      min = min(mortalities, na.rm = TRUE),
-      max = max(mortalities, na.rm = TRUE)
-    )
-  stats$Dataset <- name  # Add dataset name for identification
-  return(stats)
-}))
-
-# Print summary statistics
-tibble::as_tibble(summary_stats) %>% 
-  relocate(Dataset) %>% 
-  print()
-```
-
-# Identify potential missing values in each dataset
-```{r}
-cat("\n### Missing Data Summary:\n")
-lapply(names(list_datasets), function(name) {
-  missing_count <- sum(is.na(list_datasets[[name]]))
-  cat(name, "has", missing_count, "missing values.\n")
-})
-```
-
-We want to visualise the different datasets:
-
-We will revisit the histogram with the percent change removed.
-```{r}
-# Visualize the distribution of mortality rates with a histogram
-revised_mortalities_histogram <- ggplot(hypertens_adult_filtered_no_percent, aes(x = mortalities)) +
-  geom_histogram(binwidth = 20, fill = "blue", alpha = 0.7) +
-  scale_x_continuous(breaks = seq(0, max(hypertens_adult_filtered_no_percent$mortalities, na.rm = TRUE), by = 200)) +
-  theme_minimal() +
-  labs(title = "Distribution of Mortality Rates", x = "Mortality Rates", y = "Count")
-
-saveRDS(revised_mortalities_histogram, file = here::here("data","processed-data", "revised_mortalities_histogram.rds"))
-revised_mortalities_histogram <- readRDS(here::here("data","processed-data", "revised_mortalities_histogram.rds"))
-# Print the mortalities histogram
-print(revised_mortalities_histogram)
-```
-
-This is the total dataset and so the data is similar to the previous histogram.
-There is a strong right skew and long tail.
-Peak mortalities are below 100, and the total values likely account for the second peak.
-
-We will use a boxlot to get an overview of the other data:
-
-```{r}
-# Compare mortality rates across race-sex and sex stratifications
-
-boxplot_race_sex <- ggplot(county_race_sex, aes(x = race_ethnicity, y = mortalities, fill = sex)) +
-  geom_boxplot() +
-  theme_minimal() +
-  labs(title = "Mortality Rates by Race and Sex", x = "Race/Ethnicity", y = "Mortality Rate")
-
-saveRDS(boxplot_race_sex, file = here::here("data","processed-data", "boxplot_race_sex.rds"))
-
-boxplot_sex <- ggplot(county_sex, aes(x = sex, y = mortalities)) +
-  geom_boxplot() +
-  theme_minimal() +
-  labs(title = "Mortality Rates by Sex", x = "Sex", y = "Mortality Rate")
-
-saveRDS(boxplot_sex, file = here::here("data","processed-data", "boxplot_sex.rds"))
-
-```
-
-The box plots provide a good overview of the data nd indicate that there seem to be distinct trends and differences between races.
-It also shows that there are differences between sexes but these are less marked and distinct than between races.
-We will ook at this in the statistical analysis.
-
-We want an idea on any sptial trends in the data. We are using the sf function to plot the mortality data by the latitude and longitude.
-```{r}
-# Convert data to spatial format and create a map of mortality rates
-# Ensure longitude and latitude values are numeric
-hypertens_adult_filtered_no_percent <- hypertens_adult_filtered_no_percent %>%
-  mutate(X_long = as.numeric(X_long), Y_lat = as.numeric(Y_lat))
-# Convert to spatial object
-sf_data <- st_as_sf(hypertens_adult_filtered_no_percent, coords = c("X_long", "Y_lat"), crs = 4326)
-
-# Generate a spatial plot of mortality rates
-ggplot(sf_data) +
-  geom_sf(aes(color = mortalities)) +
-  scale_color_viridis_c() +
-  theme_minimal() +
-  labs(title = "Geospatial Distribution of Mortality Rates")
-
-```
-
-This is a very nice overview and can help us to evauate the trends further or to work on a more intricate plot. We must remember that we have not separated groups or totals yet.
-
-We want an overview of any temporal (time based) trends in the data.
-We will provide a proof of concept using a single county due to the overlap of data that will occur otherwise.
-The statistical analysis will include further refinements.
-We are only working with the values where sex and race were both overall to reduce overlap.
-```{r}
-# Line plot of mortality rates over time
-# Select a single county to illustrate the trend
-selected_county <- "13037"  # County Location ID for Calhoun, GA
-county_subset <- county_overall %>% filter(LocationID == selected_county)
-
-temporal_trends_single_county <- ggplot(county_subset, aes(x = Year, y = mortalities, group = LocationID)) +
-  geom_line(alpha = 0.6, color = "blue") +
-  geom_smooth(se = FALSE, color = "red") +
-  theme_minimal() +
-  labs(title = paste("Mortality Rate Trends in Calhoun"), 
-       x = "Year", y = "Mortality Rate")
-
-print(temporal_trends_single_county)
-```
-
-## Key Findings and Next Steps
-
-- The distribution of mortality rates is right-skewed, suggesting potential transformations before modeling.
-- Mortality rates vary markedly by race and sex, justifying stratified analysis.
-- A proof-of-concept temporal trend was created for a single county (Calhoun, GA) using Location ID, ensuring unique selection.
-- Temporal trend indicate a general increase the selected county, which should be modeled.
-- All processed data has been saved into a single RDS file (`final_processed_data.rds`) for use in statistical analysis.
+- The data was cleaned and processed to result in a dataset and a few different variables that are much smaller and more manageable to work with.
+- We checked for missing variables and checked overall trends and removed d data and variables that were not useful.
+- We have saved the different created variables as a list saved into one variable.
+- All processed data has been saved into a single RDS file (`final_processed_data.rds`) for use in exploratory data analysis.

### 2025-04-11 products/manuscript/Manuscript.qmd
Part 4 additions

@@
 csl: ../../assets/american-journal-of-epidemiology.csl
+editor: 
+  markdown: 
+    wrap: 72
 ---
@@
 
-* First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678) 
-* Second Author$^{3, *}$
-* Third Autor$^{3, \dagger}$
-* Last Author$^{1, 2, \land}$
+-   First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
+-   Second Author$^{3, *}$
+-   Third Autor$^{3, \dagger}$
+-   Last Author$^{1, 2, \land}$
 
@@
 
-1. College of Veterinary Medicine, University of Georgia, Athens, GA, USA.
+1.  College of Veterinary Medicine, University of Georgia, Athens, GA,
+    USA.
 
@@
 
-$\land$ Corresponding author: some@email.com
+$\land$ Corresponding author: some\@email.com
 
@@
 # Summary/Abstract
-Lifestyle diseases such as cardiovascular disease make up the highest causes of death in developed nations such as the United States.
-This project focuses on hypertension-related cardiovascular disease and hopes to help provide a greater understandign of the dynamics and patterns that may exist with hypertension and cardiovascualr diseased related mortality in the United States.
-The project seeks to evaluate data from the CDC on the rates of hypertension-related cardiovascular disease mortality and the associated demographic factors for this collected data, including county, age group, race/ethnicity, and sex.
-The data represents information collected from 2000 to 2019.
-This will be used to see if there are any potential patterns or correlation between the rate of mortality and these other demographic factors within the data.
+
+Lifestyle diseases such as cardiovascular disease make up the highest
+causes of death in developed nations such as the United States. This
+project focuses on hypertension-related cardiovascular disease and hopes
+to help provide a greater understandign of the dynamics and patterns
+that may exist with hypertension and cardiovascualr diseased related
+mortality in the United States. The project seeks to evaluate data from
+the CDC on the rates of hypertension-related cardiovascular disease
+mortality and the associated demographic factors for this collected
+data, including county, age group, race/ethnicity, and sex. The data
+represents information collected from 2000 to 2019. This will be used to
+see if there are any potential patterns or correlation between the rate
+of mortality and these other demographic factors within the data.
 
@@
 
-Lifestyle diseases such as cardiovascular disease make up the highest causes of death in developed nations such as the United States.
-This data analysis hopes to evaluate the rates of hypertension-related cardiovascular disease mortality in adults in the United States and trends that might be associated with demographic and geographic factors.
+Lifestyle diseases such as cardiovascular disease make up the highest
+causes of death in developed nations such as the United States. This
+data analysis hopes to evaluate the rates of hypertension-related
+cardiovascular disease mortality in adults in the United States and
+trends that might be associated with demographic and geographic factors.
 
@@
 
-The dataset was collected from the CDC website and the information was sourced from the National Vital Statistics System.
-The dataset contains rates and trends in the hypertension-related cardiovascular disease (CVD) death rates.
-The study period is from 2000 to 2019, with two main intervals displayed in the data 2000-2009, and 2010-2019.
-The data is further distributed by county, age group (36-64 years and 65 years and older), race/ethnicity, and sex.
+The dataset was collected from the CDC website and the information was
+sourced from the National Vital Statistics System. The dataset contains
+rates and trends in the hypertension-related cardiovascular disease
+(CVD) death rates. The study period is from 2000 to 2019, with two main
+intervals displayed in the data 2000-2009, and 2010-2019. The data is
+further distributed by county, age group (36-64 years and 65 years and
+older), race/ethnicity, and sex.
 
@@
 
-- Hypertension-related cardiovascular disease mortalities are higher in select demographics (age groups, race/ethnicities and sex), and thus these demographics show correlated with mortalities rates.
+-   Hypertension-related cardiovascular disease mortalities are higher
+    in select demographics (age groups, race/ethnicities and sex), and
+    thus these demographics show correlated with mortalities rates.
 
-- There are clear regional/geographic distributions associated with hypertension-related cardiovascular disease mortalities.
+-   There are clear regional/geographic distributions associated with
+    hypertension-related cardiovascular disease mortalities.
+
+Study Aims:
+- Aim 1: Assess national temporal trends in hypertension-related CVD mortality, stratified by age group.
+- Aim 2: Identify spatial clustering of county-level mortality.
+- Aim 3: Evaluate disparities in mortality trends by race/ethnicity and sex, stratified by age group.
+- Aim 4: Build predictive models using demographic and geographic features to estimate mortality trends.
+
+Hypotheses:
+- Mortality trends vary significantly over time and across age groups.
+- County-level mortality exhibits spatial clustering, reflecting geographic disparities.
+- Temporal trends differ significantly across race and sex, with widening inequities over time.
+- Predictive models can identify key demographic and spatial factors associated with higher mortality trends.
 
@@
 
-This data was downloaded from the CDC online database at the following url: https://data.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Hypertension-related-Cardiovas/uc9k-vc2j/about_data.
-A csv file was downloaded which was almost 300MB in size.
-The initial document/dataset contained 1103872 observations across 24 variables.
-There were some initial processing steps performed to evaluate and delete some of the unnecessary columns and to save this data to a .rds document to reduce the size.
-This was all outlined in the "processingcode" document, which will not run automatically because the Excel (csv) file will not be available due to size concerns.
+This data was downloaded from the CDC online database at the following
+url:
+https://data.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Hypertension-related-Cardiovas/uc9k-vc2j/about_data.
+A csv file was downloaded which was almost 300MB in size. The initial
+document/dataset contained 1103872 observations across 24 variables.
+There were some initial processing steps performed to evaluate and
+delete some of the unnecessary columns and to save this data to a .rds
+document to reduce the size. This was all outlined in the
+"processingcode" document, which will not run automatically because the
+Excel (csv) file will not be available due to size concerns.
 
-The intial exploration and cleaning and removal of the redundant columns (columns with repeated single values without any additional information).
-Some of the other columns were renamed to have the name of the column reflect what these single value columns had contained instead.
-The columns were checked to see what they contained and to see how many unique entries there were and based upon this columns were deleted and others renamed.
+The intial exploration and cleaning and removal of the redundant columns
+(columns with repeated single values without any additional
+information). Some of the other columns were renamed to have the name of
+the column reflect what these single value columns had contained
+instead. The columns were checked to see what they contained and to see
+how many unique entries there were and based upon this columns were
+deleted and others renamed.
 
-Deleted columns were: DataSource, Topic, Class, StratificationCategory1/2/3, Data_Value_Footnote, Data_Value_Footnote_Symbol, Geographic_Level, TopicID
+Deleted columns were: DataSource, Topic, Class,
+StratificationCategory1/2/3, Data_Value_Footnote,
+Data_Value_Footnote_Symbol, Geographic_Level, TopicID
 
-Renamed columns: 
-Stratification1 column to age_group, Stratification2 column to race_ethnicity, Stratification3 column to sex, Data_Value to mortalities.
+Renamed columns: Stratification1 column to age_group, Stratification2
+column to race_ethnicity, Stratification3 column to sex, Data_Value to
+mortalities.
 
-After the initial cleaning the data was evaluated again and some further small cleaning and processing steps were performed (code is visible in the processingfile-v1.qmd file) to reduce the number of rows by removing NA values.
-The resulting dataframe contains 572572 observations and 14 variables.
+After the initial cleaning the data was evaluated again and some further
+small cleaning and processing steps were performed (code is visible in
+the processingfile-v1.qmd file) to reduce the number of rows by removing
+NA values. The resulting dataframe contains 572572 observations and 14
+variables.
 
-Our summaries and the skimr function help to provide a good overview for further evaluation of the data. 
-We can see that our dataset contains a combination of numeric and character data. 
-The numeric data includes the actual incidence rates of the hypertension-related cardiovascular disease death rates, as well as the confidence intervals that were used, and geographic coordinates.
-Our character data includes the race, sex, age group, location data and the year.
+Our summaries and the skimr function help to provide a good overview for
+further evaluation of the data. We can see that our dataset contains a
+combination of numeric and character data. The numeric data includes the
+actual incidence rates of the hypertension-related cardiovascular
+disease death rates, as well as the confidence intervals that were used,
+and geographic coordinates. Our character data includes the race, sex,
+age group, location data and the year.
 
-This character data will be used as predictors of the mortality data as we break down these different variables and look for potential patterns and associations.
-The primary outcome of interest is the mortality rates but the important potential predictors are the location, year, race/ethnicity, and sex.
-AGe groups only provide two broad categories, namely 35 to 65 years old and 65 years and older, which is a bit limited but will allow for comparisons between these two groups.
+This character data will be used as predictors of the mortality data as
+we break down these different variables and look for potential patterns
+and associations. The primary outcome of interest is the mortality rates
+but the important potential predictors are the location, year,
+race/ethnicity, and sex. AGe groups only provide two broad categories,
+namely 35 to 65 years old and 65 years and older, which is a bit limited
+but will allow for comparisons between these two groups.
+
+Aim 1: We assessed national temporal trends in hypertension-related
+cardiovascular disease (CVD) mortality from 2000 to 2019 using a series
+of regression models. First, a linear regression model was fit to assess
+the overall relationship between mortality rates and year. Next,
+generalized linear models (GLMs) were used to evaluate time trends
+stratified by race/ethnicity, sex, and age group. Interaction terms
+(e.g., Year × race/ethnicity) were included to test for differential
+trends across subgroups. Estimated annual changes in mortality rates
+were compared across age groups using the emtrends and contrast
+functions from the emmeans package.
+
+Spatial Patterns of Mortality Aim 2: Spatial clustering analysis:
+County-level mortality data were converted to spatial features using
+geographic coordinates. Global and Local Moran’s I statistics were used
+to assess spatial autocorrelation. Spatial weights were generated using
+5-nearest-neighbor methods. Local clustering was visualized using
+standardized mortality z-scores and LISA cluster classifications.
+
+Evaluating National Demographic Disparities Aim 3: We fit generalized
+linear models (GLMs) with interaction terms to estimate yearly changes
+in hypertension-related CVD mortality rates from 2000–2019 across
+demographic groups. Using emtrends() from the emmeans package, we
+extracted slope estimates by race/ethnicity and sex, stratified by age
+group (35–64 and 65+). Pairwise contrasts were conducted to evaluate
+statistically significant differences in trends between groups.
+
+Statistical Modeling (Aim 4) To predict county-level
+hypertension-related cardiovascular disease (CVD) mortality, we trained
+two supervised machine learning models: a random forest and a LASSO
+regression. Models were trained on 80% of the data, with the remaining
+20% held out for testing. Ten-fold cross-validation was used within the
+training set for hyperparameter tuning. Predictors included year, age
+group, race/ethnicity, and sex.
+The modeling dataset included demographic predictors (age group, race/ethnicity, sex) and time (year).
+Since several predictors were categorical, we applied dummy encoding (one-hot encoding) using step_dummy() within the recipes framework to convert these factors into binary indicator variables.
+We also normalized numeric predictors and removed any zero-variance predictors.
+Models were trained using a stratified 80/20 train/test split and evaluated using 10-fold cross-validation on the training set.
+Random forest hyperparameters were selected from a regular grid of mtry and
+min_n values, while LASSO regression used a log-scaled penalty grid.
+Final model performance was evaluated using RMSE, MAE, and R² on the
+test set. Confidence intervals were not computed for these metrics to
+avoid additional computational burden.
 
@@
 
-A summary table was produced using the skimr function which is printed below.
-
-```{r}
-skim_summary_hypertens_filtered <- readRDS(here::here("data","processed-data", "skim_summary_hypertens_filtered.rds"))
-# Print the skim summary
-print(skim_summary_hypertens_filtered)
-
-```
-
-Our summary data shows that the mean of the hypertension cardiovascular-diseases mortalities is around 203, with a standard deviation of 223.
-
-A histogram was made of the mortalities columns since this is the main (only) column of interest with numerical data.
-
-```{r}
-revised_mortalities_histogram <- readRDS(here::here("data","processed-data", "revised_mortalities_histogram.rds"))
-# Print the mortalities histogram
-plot(revised_mortalities_histogram)
-```
-
-Our histogram depicts the hypertension-related cardiovasculat disease mortalities in the dataset, showing the number of mortalities and the frequency (count) of sprecific mortality values.
-The mortality counts are clustered towards the lower end of the scale, with a strong right-skew (long tail to the right/higher values).
-Peak mortalities are below 100, and the total values likely account for the second peak.
-
-Boxplots were created to evaluate the mortalities by race and by sex.
+We began by exploring patterns in hypertension-related cardiovascular disease (CVD) mortality among U.S.
+adults aged 35 and older from 2000 to 2019.
+The EDA included summary statistics, histograms, and boxplots to assess mortality distributions overall and by key demographic subgroups (race/ethnicity, sex, age group).
+Temporal trends were visualized using line plots stratified by subgroup and geographic region.
+Spatial variation was explored with choropleth maps displaying county-level mortality rates across time and demographic strata.
+These descriptive analyses revealed substantial variation in mortality trends over time, with emerging disparities by race, sex, and age group.
+Additionally, spatial visualizations suggested clustering in certain states, laying the foundation for formal spatial and demographic analyses in later aims.
+All visualizations and cleaned summary tables were saved for downstream modeling.
 
@@
 
-## Key Findings and Next Steps
-
-- The distribution of mortality rates is right-skewed, suggesting potential transformations before modeling.
-- Mortality rates vary markedly by race and sex, justifying stratified analysis.
-- A proof-of-concept temporal trend was created for a single county (Calhoun, GA) using Location ID, ensuring unique selection.
-- Temporal trend indicate a general increase the selected county, which should be modeled.
-- All processed data has been saved into a single RDS file (`final_processed_data.rds`) for use in statistical analysis.
-
 ## Basic statistical analysis
 
-Spatial plots of mortalities per geographic regions were generated for each race and compared on one plot.
-Two generalized linear models were evaluated, the first examined how time, race/ethnicity, and sex (plot below) influence mortality rates, and the second tested whether the effect of time varied by race/ethnicity.
+Spatial plots of mortalities per geographic regions were generated for
+each race and compared on one plot. Two generalized linear models were
+evaluated, the first examined how time, race/ethnicity, and sex (plot
+below) influence mortality rates, and the second tested whether the
+effect of time varied by race/ethnicity.
 
-Overall, black individuals had the highest predcted mortality rates from the generalized linear models, followed by white and native american populations.
-Black individuals also had the highest hypertension-related mortalities on the boxplot follwoed by white individuals.
-Females appeared to have lower mortality rates than males across all races.
+Overall, black individuals had the highest predcted mortality rates from
+the generalized linear models, followed by white and native american
+populations. Black individuals also had the highest hypertension-related
+mortalities on the boxplot follwoed by white individuals. Females
+appeared to have lower mortality rates than males across all races.
 
-Race, sex and time all seem to be predictors of mortality. We see differences in trends over time for different racial groups.
+Race, sex and time all seem to be predictors of mortality. We see
+differences in trends over time for different racial groups.
 
-Our spatial clustering has revealed geographic clustering of high-mortality regions, with high densities in the Eastern regions of the country.
-There also appears to be a higher density of mortalities from people in the race/ethnic group Black in certain regions.
-Asian and Pacific Islander groups had relatively lower mortality rates and geographic clustering.
+Our spatial clustering has revealed geographic clustering of
+high-mortality regions, with high densities in the Eastern regions of
+the country. There also appears to be a higher density of mortalities
+from people in the race/ethnic group Black in certain regions. Asian and
+Pacific Islander groups had relatively lower mortality rates and
+geographic clustering.
 
-```{r}
+```{r, echo=FALSE, message=FALSE}
 #Print the glm plot
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE}
 #Print the combined spatial plot
@@
 
-_
+Temporal Trends in Mortality Rates (2000–2019) (Aim 1): From 2000 to
+2019, overall CVD mortality rates showed minimal change at the national
+level. However, stratified models revealed substantial differences
+across subgroups. Black individuals consistently exhibited the highest
+mortality rates, though a declining trend was observed. In contrast,
+rates among American Indian/Alaska Native individuals increased markedly
+over time. Older adults (65+) had significantly higher mortality rates
+and experienced a steeper upward trend compared to adults aged 35–64
+(estimated slope: 1.22 vs. 1.04 deaths per 100,000 per year; p \<
+0.001). These findings highlight persistent disparities and suggest
+diverging subgroup trends in hypertension-related CVD mortality.
+
+Spatial clustering analysis (Aim 2): We evaluated spatial patterns in
+age-adjusted county-level hypertension-related cardiovascular disease
+(CVD) mortality using global and local spatial autocorrelation methods.
+Global spatial autocorrelation of county-level mortality rates was
+statistically significant (Moran’s I = 0.0435, p \< 0.001), indicating a
+weak but non-random spatial clustering pattern across the United States.
+Local Indicators of Spatial Association (LISA) identified significant
+clusters of high and low mortality values. LISA cluster types (e.g.,
+high-high, low-low, spatial outliers) were visualized using
+centroid-based maps for the contiguous U.S. and select states (Georgia,
+Mississippi, Texas, Kentucky). Despite the overall geographic patterns,
+only a small number of statistically significant local clusters were
+detected, suggesting that while some regions exhibit elevated risk,
+strong clustering at the county level may be limited or diffuse.
+
+Investigating demographic disparities (Aim 3): Annual changes in
+mortality rates varied substantially across demographic groups. Among
+adults aged 35–64, Black individuals experienced the steepest declines,
+while mortality increased most among males and American Indian/Alaska
+Native populations. Similar patterns were observed among those aged 65+,
+with significant disparities between groups. Pairwise comparisons
+revealed the largest slope differences between Overall Male and Black
+Overall groups, followed by disparities involving American Indian/Alaska
+Native populations. These findings suggest widening inequities in
+hypertension-related CVD mortality trends.
+
+Predictive Modeling of County-Level Mortality (Aim 4) To evaluate the
+predictive capacity of demographic and temporal factors on
+hypertension-related cardiovascular mortality, we compared two machine
+learning models: a random forest and a LASSO regression. Both models
+were trained on 80% of the dataset and evaluated on a held-out 20% test
+set, using cross-validation during tuning to optimize performance.
+
+The random forest model demonstrated the strongest performance,
+achieving a Root Mean Squared Error (RMSE) of 115.3, a Mean Absolute
+Error (MAE) of 66.9, and an R² of 0.748 on the test set. In comparison,
+the LASSO regression model yielded an RMSE of 119.6, MAE of 68.1, and R²
+of 0.726. While slightly less accurate, the LASSO model offers valuable
+interpretability by shrinking less important coefficients to zero,
+effectively performing variable selection.
+
+When comparing predicted vs observed plots, both models exhibited a distinct pattern of horizontal banding.
+This is likely due to the large number of categorical predictors (race, sex, age group) which were dummy-encoded.
+Many subsets of the population across counties share identical predictor combinations (e.g., "Black females aged 65+ in 2002"), leading to repeated predicted values for those groups.
+This results in stacked horizontal lines in the prediction plots, especially when using models like LASSO that produce linear fits with limited flexibility, and even in random forests when groupings dominate feature structure.
+
+A comparison of model performance is summarized in Table (...X...).
+These results suggest that both models capture meaningful patterns in
+the data, but the random forest model more accurately predicts mortality
+outcomes. Importantly, the use of interpretable models like LASSO can
+aid in identifying potential risk factors despite modest reductions in
+predictive power.
+
+```{r, echo=FALSE, message=FALSE}
+#Print the model performance summary
+model_performance_summary_table <- readRDS(here::here("results", "tables", "model_performance_summary.rds"))
+model_performance_summary_table
+```
+
+\_
 
@@
 
+Aim 3: These results show that hypertension-related CVD mortality trends
+have not improved equally across demographic groups. While some
+populations saw modest or no change, others—particularly Black
+adults—experienced significant declines. Conversely, mortality rates
+rose among males and American Indian/Alaska Native individuals. The
+magnitude and significance of these differences reinforce the need for
+targeted public health efforts to reduce disparities in cardiovascular
+outcomes.
+
 ## Strengths and Limitations
@@
 
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
+Note that this cited reference will show up at the end of the document,
+the reference formatting is determined by the CSL file specified in the
+YAML header. Many more style files for almost any journal [are
+available](https://www.zotero.org/styles). You also specify the location
+of your bibtex reference file in the YAML. You can call your reference
+file anything you like.
 

### 2025-04-11 code/analysis-code/statistical-analysis.R

@@ -1,222 +1,0 @@
-###############################
-# analysis script
-#
-#this script loads the processed, cleaned data, does a simple analysis
-#and saves the results to the results folder
-
-#load needed packages. make sure they are installed.
-#install.packages("gt")
-library(ggplot2) #for plotting
-library(broom) #for cleaning up output from lm()
-library(here) #for data loading/saving
-library(tidyverse)
-library(tidymodels)
-library(gt)
-library(stringr)
-library(patchwork)
-
-
-#path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","final_processed_data.rds")
-
-#load data. 
-final_processed_data <- readRDS(data_location)
-
-# Extract key datasets
-processed_data_no_percent <- final_processed_data[["processed_data_no_percent"]]
-county_race_sex <- final_processed_data[["county_race_sex"]]
-county_race <- final_processed_data[["county_race"]]
-county_sex <- final_processed_data[["county_sex"]]
-county_overall <- final_processed_data[["county_overall"]]
-
-######################################
-#Data fitting/statistical analysis
-######################################
-
-############################
-
-# Exploring relationships between mortality rates and individual predictors
-
-# Relationship between mortality and race
-race_plot <- ggplot(county_race, aes(x = race_ethnicity, y = mortalities)) +
-  geom_boxplot() +
-  theme_minimal() +
-  labs(title = "Mortality Rates by Race/Ethnicity", x = "Race/Ethnicity", y = "Mortality Rate") +
-  theme(axis.text.x = element_text(size = 10)) +
-  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10))
-
-print(race_plot)
-ggsave(here::here("products", "manuscript", "supplement", "mortality_by_race.png"), plot = race_plot)
-
-# Relationship between mortality and sex
-sex_plot <- ggplot(county_sex, aes(x = sex, y = mortalities)) +
-  geom_boxplot() +
-  theme_minimal() +
-  labs(title = "Mortality Rates by Sex", x = "Sex", y = "Mortality Rate")
-
-print(sex_plot)
-ggsave(here::here("products", "manuscript", "supplement", "mortality_by_sex.png"), plot = sex_plot)
-
-
[…]

### 2025-04-11 code/analysis-code/statistical-analysis-v2-slim.qmd
Changes to Part 4

--- 2025-04-11 code/analysis-code/statistical-analysis-v2-slim-models.qmd
@@
 ```
+These plots are best viewed as images outside of R due to size.
 
@@
 
-```{r}
-# Define file paths for saving/loading tuning results
-tuned_path     <- here::here("results", "large-files", "rf_tuned.rds")
-metrics_path   <- here::here("results", "output", "rf_tuned_metrics.rds")
-best_param_path <- here::here("results", "output", "rf_best_params.rds")
-
-# Check if tuning results already exist
-if (file.exists(tuned_path) && file.exists(metrics_path) && file.exists(best_param_path)) {
-  message( "Tuning results already saved. Loading RF tuning results...")
-
-  # Load saved objects
-  rf_tuned <- readRDS(tuned_path)
-  rf_tuned_metrics <- readRDS(metrics_path)
-  best_rf_params <- readRDS(best_param_path)
+Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.
+
+```{r}
+## ---- rf_tune_and_fit_if_guarded, message=TRUE, warning=FALSE -------------------
+
+library(tidymodels)
+library(vip)
+library(here)
+
+# Create directories
+dir.create(here("results", "output"), recursive = TRUE, showWarnings = FALSE)
+dir.create(here("results", "large-files"), recursive = TRUE, showWarnings = FALSE)
+
+# Define output paths
+rf_tuned_path     <- here("results", "large-files", "rf_tuned.rds")
+rf_metrics_path   <- here("results", "output", "rf_tuned_metrics.rds")
+rf_best_param_path <- here("results", "output", "rf_best_params.rds")
+rf_grid_path      <- here("results", "output", "rf_grid.rds")
+rf_slim_path      <- here("results", "output", "rf_model_minimal.rds")
+
+# Check if all exist
+if (file.exists(rf_tuned_path) && file.exists(rf_metrics_path) &&
+    file.exists(rf_best_param_path) && file.exists(rf_slim_path)) {
+
+  message("✅ All RF files found. Loading saved versions...")
+
+  rf_tuned <- readRDS(rf_tuned_path)
+  rf_metrics <- readRDS(rf_metrics_path)
+  best_rf_params <- readRDS(rf_best_param_path)
+  rf_slim <- readRDS(rf_slim_path)
+  rf_final_model <- rf_slim$model
+  rf_vip <- rf_slim$vip
 
 } else {
-  message(" Running Random Forest tuning...")
-# Set up a small tuning grid for efficiency
+  message("🔁 Fitting Random Forest from scratch...")
+
+  # Define grid
   rf_grid <- grid_regular(
-  mtry(range = c(1, 4)),     # Explore 1 to 4 predictors
-  min_n(range = c(5, 25)),   # Minimum observations per leaf
-  levels = 3)                # Total levels in grid for each param
-
-# Tune the random forest model
+    mtry(range = c(1, 4)),
+    min_n(range = c(5, 25)),
+    levels = 3
+  )
+  saveRDS(rf_grid, rf_grid_path)
+
+  # Tune
   set.seed(123)
@@
     metrics = metric_set(rmse, rsq, mae),
-  control = control_grid(save_pred = TRUE))
-
-# Collect metrics from the tuning process
-rf_tuned_metrics <- collect_metrics(rf_tuned)
+    control = control_grid(save_pred = TRUE)
+  )
+
+  # Extract results
+  rf_metrics <- collect_metrics(rf_tuned)
   best_rf_params <- select_best(rf_tuned, metric = "rmse")
 
-# Save results to avoid re-running this chunk
-  saveRDS(rf_tuned, tuned_path)
-  saveRDS(rf_tuned_metrics, metrics_path)
-  saveRDS(best_rf_params, best_param_path)
-
-  message("Tuning complete and results saved.")}
-
-```
-```{r}
-## ---- fit_or_load_rf_model, message=TRUE, warning=FALSE -------------------
-
-use_slim_model <- FALSE # Toggle this to FALSE to regenerate slim model if needed
-
-slim_rf_path <- here::here("results", "output", "rf_model_minimal.rds")
-full_rf_path <- here::here("results", "large-files", "rf_tuned.rds")
-full_rf_metrics_path <- here::here("results", "output", "rf_tuned_metrics.rds")
-best_rf_params_path <- here::here("results", "output", "rf_best_params.rds")
-
-if (use_slim_model) {
-  if (!file.exists(slim_rf_path)) {
-    stop("Slimmed model not found. Please run with use_slim_model <- FALSE first to generate it.")
-  }
-
-  message("Loading slimmed RF model...")
-  rf_slim <- readRDS(slim_rf_path)
-  rf_final_model <- rf_slim$model
-  rf_metrics <- rf_slim$metrics
-  rf_vip <- rf_slim$vip
-  best_rf_params <- readRDS(best_rf_params_path)
-
-} else {
-  if (!file.exists(full_rf_path)) {
-    stop("Full Random Forest model not found. Please rerun the tuning block.")
-  }
-
-  message("Loading full RF model and generating slim version...")
-  rf_tuned <- readRDS(full_rf_path)
-  rf_metrics <- readRDS(full_rf_metrics_path)
-  rf_final_model <- extract_workflow(rf_tuned)
+  # Save tuning results
+  saveRDS(rf_tuned, rf_tuned_path)
+  saveRDS(rf_metrics, rf_metrics_path)
+  saveRDS(best_rf_params, rf_best_param_path)
+
+  # Finalize and fit full model
+  final_rf_workflow <- finalize_workflow(rf_workflow, best_rf_params)
+
+  set.seed(123)
+  rf_final_model <- fit(final_rf_workflow, data = train_data)
+
+  # VIP
   rf_vip <- extract_fit_parsnip(rf_final_model) %>% vip::vi()
-  best_rf_params <- readRDS(best_rf_params_path)
-
-  if (!file.exists(slim_rf_path)) {
+
+  # Save slim model
   saveRDS(
     list(model = rf_final_model, metrics = rf_metrics, vip = rf_vip),
-      file = slim_rf_path
-    )
-    message("Slimmed model saved.")}}
-
-```
-Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.
+    file = rf_slim_path
+  )
+
+  message("✅ RF slim model and all outputs saved.")}
+
+```
 
@@
 The full RDS file is saved in the large-files folder which will not automatically upload to github.
+
+```{r}
+## ---- temporary_rf_runtime_log, message=TRUE, warning=FALSE -------------------
+
+# Optional: Log run time and session info for audit trail
+start_time <- Sys.time()
+Sys.sleep(1)  # simulate runtime block
+end_time <- Sys.time()
+
+writeLines(as.character(end_time - start_time), here::here("results", "output", "rf_runtime_duration.txt"))
+writeLines(capture.output(sessionInfo()), here::here("results", "output", "rf_session_info.txt"))
+
+message("⏱️ Runtime and session info saved.")
+
+```
 
@@
   rf_predictions <- readRDS(pred_path)
-  rf_metrics <- readRDS(metrics_path)
+  rf_test_metrics <- readRDS(metrics_path)
 
@@
   # Evaluate performance
-  rf_metrics <- rf_predictions %>%
+  rf_test_metrics <- rf_predictions %>%
     metrics(truth = mortalities, estimate = .pred)
@@
   saveRDS(rf_predictions, file = pred_path)
-  saveRDS(rf_metrics, file = metrics_path)
+  saveRDS(rf_test_metrics, file = metrics_path)
 
@@
 # Review metrics
-print(rf_metrics)
+print(rf_test_metrics)
 
@@
 # Load saved RF metrics
-rf_metrics <- readRDS(here::here("results", "output", "rf_test_metrics.rds"))
+rf_test_metrics <- readRDS(here::here("results", "output", "rf_test_metrics.rds"))
 
 # Convert to wide format for easy table summary
-rf_summary_table <- rf_metrics %>%
+rf_summary_table <- rf_test_metrics %>%
   select(.metric, .estimate) %>%
@@
 
-```{r}
-## Tune LASSO penalty parameter
-
-# File paths
-lasso_tuned_path <- here::here("results", "large-files", "lasso_tuned.rds")
-lasso_metrics_path <- here::here("results", "output", "lasso_tuned_metrics.rds")
-lasso_best_param_path <- here::here("results", "output", "lasso_best_params.rds")
-
-# Load or tune
-if (file.exists(lasso_tuned_path) && file.exists(lasso_metrics_path) && file.exists(lasso_best_param_path)) {
-  message("Loading saved LASSO tuning results...")
+Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.
+
+```{r}
+## ---- lasso_tune_and_fit_if_guarded, message=TRUE, warning=FALSE -------------------
+
+library(tidymodels)
+library(vip)
+library(here)
+
+# Create folders
+dir.create(here::here("results", "output"), showWarnings = FALSE, recursive = TRUE)
+dir.create(here::here("results", "large-files"), showWarnings = FALSE, recursive = TRUE)
+
+# Define output paths
+lasso_tuned_path        <- here("results", "large-files", "lasso_tuned.rds")
+lasso_metrics_path      <- here("results", "output", "lasso_tuned_metrics.rds")
+lasso_best_param_path   <- here("results", "output", "lasso_best_params.rds")
+lasso_grid_path         <- here("results", "output", "lasso_grid.rds")
+slim_lasso_path         <- here("results", "output", "lasso_model_minimal.rds")
+
+# Check if all saved files already exist
+if (file.exists(lasso_tuned_path) &&
+    file.exists(lasso_metrics_path) &&
+    file.exists(lasso_best_param_path) &&
+    file.exists(slim_lasso_path)) {
+
+  message("✅ All LASSO files found. Loading saved versions...")
 
   lasso_tuned <- readRDS(lasso_tuned_path)
-  lasso_tuned_metrics <- readRDS(lasso_metrics_path)
+  lasso_metrics <- readRDS(lasso_metrics_path)
   best_lasso_params <- readRDS(lasso_best_param_path)
+  lasso_slim <- readRDS(slim_lasso_path)
+  final_lasso_model <- lasso_slim$model
+  lasso_vip <- lasso_slim$vip
+  lasso_coeffs <- lasso_slim$coefficients
+
 } else {
-  message("Running LASSO tuning...")
-
-  lasso_grid <- grid_regular(penalty(range = c(-4, -1)), levels = 10)  # log10 scale
-
+  message("🔁 Fitting LASSO model from scratch...")
+
+  # Define grid
+  lasso_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 20)
+  saveRDS(lasso_grid, lasso_grid_path)
+
+  # Tune
   set.seed(123)
@@
     metrics = metric_set(rmse, rsq, mae),
-    control = control_grid(save_pred = TRUE))
-
-  lasso_tuned_metrics <- collect_metrics(lasso_tuned)
+    control = control_grid(save_pred = TRUE)
+  )
+
+  # Extract and save tuning results
+  lasso_metrics <- collect_metrics(lasso_tuned)
   best_lasso_params <- select_best(lasso_tuned, metric = "rmse")
 
-  # Save results
   saveRDS(lasso_tuned, lasso_tuned_path)
-  saveRDS(lasso_tuned_metrics, lasso_metrics_path)
+  saveRDS(lasso_metrics, lasso_metrics_path)
   saveRDS(best_lasso_params, lasso_best_param_path)
 
-  message("LASSO Tuning complete and results saved.")
-}
-
-# Preview metrics
-print(lasso_metrics)
-```
-
-Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.
-```{r}
-## ---- slim_lasso_model_save, message=TRUE, warning=FALSE -------------------
-
-slim_lasso_path <- here::here("results", "output", "lasso_model_minimal.rds")
-
-if (!file.exists(slim_lasso_path)) {
-  message("Creating slimmed LASSO model object...")
-
-  lasso_final <- extract_workflow(lasso_tuned)
-  lasso_metrics <- lasso_tuned_metrics
+  # Finalize and fit
+  final_lasso_workflow <- finalize_workflow(lasso_workflow, best_lasso_params)
+
+  set.seed(123)
+  final_lasso_model <- fit(final_lasso_workflow, data = train_data)
 
   # Extract coefficients
-  lasso_fit <- extract_fit_parsnip(lasso_final)
-  lasso_coeffs <- as.matrix(lasso_fit$fit$beta)
-
+  lasso_coeffs <- as.matrix(extract_fit_parsnip(final_lasso_model)$fit$beta)
+
+  # VIP table
+  lasso_vip <- tibble(
+    variable = rownames(lasso_coeffs),
+    importance = abs(lasso_coeffs[, 1])
+  ) %>% arrange(desc(importance))
+
+  # Save slimmed version
   saveRDS(
-    list(model = lasso_final, metrics = lasso_metrics, coefficients = lasso_coeffs),
+    list(model = final_lasso_model, metrics = lasso_metrics, coefficients = lasso_coeffs, vip = lasso_vip),
     file = slim_lasso_path
@@
 
-  message("Saved slimmed LASSO model as lasso_model_minimal.rds")
-} else {
-  message("Slimmed LASSO model already exists. Skipping save.")
-}
-
+  message("✅ Final LASSO model fitted and slim version saved.")}
 ```
@@
 ```{r}
-## ---- load_lasso_model, message=TRUE, warning=FALSE -------------------
-
-use_slim_lasso_model <- TRUE  # Toggle to FALSE to use full model
-
-if (use_slim_lasso_model) {
-  if (!file.exists(here::here("results", "output", "lasso_model_minimal.rds"))) {
-    stop("Slim LASSO model not found. Please run the slim-save chunk or disable slim mode.")
-  }
-
-  message("Loading slimmed LASSO model...")
-  lasso_slim <- readRDS(here::here("results", "output", "lasso_model_minimal.rds"))
-
-  lasso_final_model <- lasso_slim$model
-  lasso_metrics <- lasso_slim$metrics
-  lasso_coeffs <- lasso_slim$coefficients
-  best_lasso_params <- readRDS(here::here("results", "output", "lasso_best_params.rds"))
-
-} else {
-  if (!file.exists(here::here("results", "large-files", "lasso_tuned.rds"))) {
-    stop("Full LASSO model not found.")
-  }
-
-  message("Loading full LASSO model...")
-  lasso_tuned <- readRDS(here::here("results", "large-files", "lasso_tuned.rds"))
-  lasso_metrics <- readRDS(here::here("results", "output", "lasso_tuned_metrics.rds"))
-  lasso_final_model <- extract_workflow(lasso_tuned)
-
-  lasso_fit <- extract_fit_parsnip(lasso_final_model)
-  lasso_coeffs <- as.matrix(lasso_fit$fit$beta)
-
-  best_lasso_params <- readRDS(here::here("results", "output", "lasso_best_params.rds"))}
-
+## ---- lasso_confidence_metrics, message=TRUE, warning=FALSE -------------------
+
+lasso_rmse_summary <- lasso_metrics %>%
+  filter(.metric == "rmse") %>%
+  mutate(
+    lower_95 = mean - 1.96 * std_err,
+    upper_95 = mean + 1.96 * std_err
+  ) %>%
+  arrange(mean)
+
+write.csv(lasso_rmse_summary, here::here("results", "output", "lasso_rmse_summary.csv"), row.names = FALSE)
+
+message("✅ LASSO RMSE summary with CI estimates saved.")
 ```
@@
 ```{r}
+## ---- temporary_lasso_runtime_log, message=TRUE, warning=FALSE -------------------
+
+start_time <- Sys.time()
+Sys.sleep(1)  # Simulated delay
+end_time <- Sys.time()
+
+writeLines(as.character(end_time - start_time), here::here("results", "output", "lasso_runtime_duration.txt"))
+writeLines(capture.output(sessionInfo()), here::here("results", "output", "lasso_session_info.txt"))
+
+message("⏱️ LASSO runtime and session info saved.")
+
+```
+
+```{r}
 ## Final LASSO Model Fit and Evaluation
@@
   # Finalize model
-  final_lasso_model <- finalize_model(lasso_spec, best_lasso_params)
-
-  # Build workflow
-  final_lasso_workflow <- workflow() %>%
-    add_model(final_lasso_model) %>%
-    add_recipe(mortality_recipe)
+  final_lasso_workflow <- finalize_workflow(lasso_spec, best_lasso_params)
 
@@
 # Load test metrics
-rf_metrics <- readRDS(here::here("results", "output", "rf_test_metrics.rds"))
+rf_test_metrics <- readRDS(here::here("results", "output", "rf_test_metrics.rds"))
 lasso_metrics <- readRDS(here::here("results", "output", "lasso_test_metrics.rds"))
@@
 # Add model names
-rf_metrics$model <- "Random Forest"
+rf_test_metrics$model <- "Random Forest"
 lasso_metrics$model <- "LASSO Regression"
@@
 # Combine
-all_metrics <- bind_rows(rf_metrics, lasso_metrics)
+all_metrics <- bind_rows(rf_test_metrics, lasso_metrics)
 
@@
 ```{r}
-model_perf_numeric <- bind_rows(rf_metrics, lasso_metrics) %>%
+model_perf_numeric <- bind_rows(rf_test_metrics, lasso_metrics) %>%
   select(model, .metric, .estimate) %>%

### 2025-04-11 products/manuscript/Manuscript.qmd
Changes to Part 4

@@
 document to reduce the size. This was all outlined in the
-"processingcode" document, which will not run automatically because the
+"processingcode" document (and README file), which will not run automatically because the
 Excel (csv) file will not be available due to size concerns.

### 2025-04-19 code/analysis-code/statistical-analysis-v2-slim.qmd
Part 5 Additions

@@
 
-##Setup
+# Setup
 
@@
 
-#Data fitting/statistical analysis
-
-##Aim 1: Temporal Trends by Age Group
+# Data fitting/statistical analysis
+
+Summary statistics of baseline mortality by race/ethnicity, sex, and age group are provided later (between Aim 3 and 4) to contextualize subsequent slope estimates.
+
+## Aim 1: Temporal Trends by Age Group
 
@@
 
-####1.1 Linear Model - Overall Mortality over time
+### 1.1 Linear Model - Overall Mortality over time
 
@@
 saveRDS(lm_results_year, file = here::here("products", "manuscript", "supplement", "lm_results_year.rds"))
+
+# Extract model fit statistics
+model_fit_lm <- broom::glance(lm_model_year) %>%
+  dplyr::mutate(model = "LM: National Year Trend")
+
+# Save as table
+saveRDS(model_fit_lm, here::here("results", "tables", "model_fit_national.rds"))
 
@@
 This model provides a first look at overall temporal trends in mortality rates.
-The flat trend line suggests that overall mortality rates changed minimally from 2000 to 2019, though high variability exists.
-
-#####1.2 GLM: Race and Sex as predictors
-GLM with race and sex as additive predictors to assess overall differences in mortality rates
+
+The linear regression model estimated a modest but statistically significant annual increase in national hypertension-related CVD mortality rates between 2000 and 2019 (slope = 1.40, p < 2.2e-16).
+However, as shown in the plot, the fitted trend line is visually flat due to the scale and large variation in county-level rates.
+The intercept value is not directly interpretable, but reflects the expected mortality rate at year zero in the model's coding.
+Despite the statistically significant trend, the magnitude of change is small relative to the observed range, suggesting that national mortality rates remained relatively stable over time.
+This model does not account for age or other demographic factors, motivating the more detailed modeling in the next section.
+
+### 1.2 GLM: Race and Sex as predictors GLM with race and sex as additive predictors to assess overall differences in mortality rates
+
 ```{r}
@@
 
-The GLM model incorporating race/ethnicity and sex reveals substantial disparities in mortality rates between groups. Black populations consistently show the highest predicted rates across the period.
-
-####1.3 GLM: Year by Race interaction model:
-Fit GLM including interaction between year and race to examine race-specific trends over time
+The GLM including race/ethnicity and sex as predictors highlights large disparities in predicted mortality rates.
+Black populations had the highest predicted rates, significantly higher than all other groups (e.g., β = 205.8, p < 2e-16).
+Hispanic and Asian/Pacific Islander groups had lower mortality, though the Asian effect was not significant (p = 0.35).
+Males also showed significantly higher mortality than females (β = 20.3, p < 2e-16).
+Notably, the “Overall” category behaves inconsistently in plots due to it being modeled as a separate group rather than a true average, likely affected by unmodeled age variation.
+This supports the need to model interactions and stratify further by age.
+
+### 1.3 GLM: Year by Race interaction model: Fit GLM including interaction between year and race to examine race-specific trends over time
 
@@
 ```
-By including interaction terms, we can evaluate whether different racial/ethnic groups experienced distinct temporal trends.
-Here, trends diverge sharply: Black mortality rates decline, while rates rise for American Indian/Alaska Native populations.
-
-####1.4: GLM: Year by Age group
+
+Including an interaction between year and race revealed significantly divergent temporal trends.
+Mortality rates among Black individuals showed the steepest annual decline (β = –7.06, p < 0.001), while American Indian and Alaska Native (AI/AN) populations exhibited the largest annual increase (β for year = 3.38, p < 0.001, with positive main effect and significant negative interactions for other groups).
+Asian and Pacific Islander, Hispanic, and White populations also showed declining trends, though to a lesser extent.
+These results highlight substantial heterogeneity in how hypertension-related CVD mortality changed over time across racial/ethnic groups.
+
+Given known age-related disparities in cardiovascular risk, we next modeled mortality trends across age groups.
+This model assesses whether older adults experienced steeper or flatter trajectories over time.
+
+### 1.4: GLM: Year by Age group
 
 GLM with interaction between year and age group to examine differential temporal trends
+
 ```{r}
@@
 saveRDS(glm_age_group_results, file = here::here("products", "manuscript", "supplement", "glm_age_group_results.rds"))
+
+# Extract model fit statistics
+model_fit_agegroup <- broom::glance(glm_age_group) %>%
+  dplyr::mutate(model = "GLM: Age Group")
+
+# Save as table
+saveRDS(model_fit_agegroup, here::here("results", "tables", "model_fit_agegroup.rds"))
 
@@
 ```
-This model identifies age-specific trends. Mortality among individuals aged 65+ remains substantially higher and shows modest increases, while younger age groups have lower, flatter trends.
-
-####1.5: Slope comparison by Age group
-
-Comparing slopes using the emmeans function.
-Use estimated marginal trends (emtrends) to compare yearly slope by age group.
+
+This model identifies age-specific trends.Mortality among individuals aged 65+ remains substantially higher and shows modest increases, while younger age groups have lower, flatter trends.
+Including an interaction between year and age group revealed that mortality rates are substantially higher in the 65+ age group across all years.
+While both age groups show increasing mortality over time, the slope is significantly steeper for older adults (interaction term β = 0.224, p < 0.001), suggesting accelerated increases in mortality among those aged 65 and older.
+
+### 1.5: Slope comparison by Age group
+
+Comparing slopes using the emmeans function. Use estimated marginal trends (emtrends) to compare yearly slope by age group.
 
@@
 
+# Format slope estimates for reporting
+slope_table_age_fmt <- as.data.frame(summary(em_trends_age)) %>%
+  dplyr::mutate(
+    Estimate = round(Year.trend, 3),
+    CI = paste0(round(lower.CL, 2), "–", round(upper.CL, 2)),
+    SE = round(SE, 3)) %>%
+  dplyr::select(`Age Group` = age_group, Estimate, SE, CI)
+
+# Save as manuscript-ready table
+saveRDS(slope_table_age_fmt, here::here("results", "tables", "slope_table_agegroup_fmt.rds"))
+
 # compare trends between groups (pairwise contrast)
@@
 saveRDS(slope_contrast, file = here::here("products", "manuscript", "supplement", "slope_contrast_age_group.rds"))
+
+# Format pairwise slope differences for age group
+slope_contrast_age_fmt <- as.data.frame(summary(slope_contrast, infer = c(TRUE, TRUE))) %>%
+  dplyr::mutate(
+    Estimate = round(estimate, 3),
+    CI = paste0(round(lower.CL, 2), "–", round(upper.CL, 2)),
+    SE = round(SE, 3),
+    p_value = signif(p.value, 3)
+  ) %>%
+  dplyr::select(Contrast = contrast, Estimate, SE, CI, p_value)
+
+# Save formatted contrast table
+saveRDS(slope_contrast_age_fmt, here::here("results", "tables", "slope_contrast_agegroup_fmt.rds"))
+
+slope_contrast_age_fmt
 
@@
 
+Estimated yearly changes in mortality rates were significantly higher for adults aged 65+ (slope = 1.26, 95% CI: 1.18–1.34) compared to those aged 35–64 (slope = 1.04, 95% CI: 0.96–1.12).
+The difference in slope between age groups was statistically significant (difference = -0.224, p < 0.001), indicating a steeper rise in mortality among older adults.
+
+Overall, the results from Aim 1 highlight that age is a significant modifier of temporal trends in hypertension-related cardiovascular mortality.
+While the overall trend appeared stable at the national level, subgroup analyses revealed that mortality has been increasing more sharply among older adults, and certain racial/ethnic groups show distinct patterns of change over time.
+These findings justify further exploration of disparities and spatial variation in subsequent aims.
+
 ## Aim 2: Spatial Analysis of Hypertension Mortality Rates
+
 ## Evaluating spatial distribution and clustering patterns
@@
 
-###2.1: Static maps by Race:
+### 2.1: Static maps by Race:
 
 Converting county data to spatial object using longitude/latitude. 
-Creating static spatial map for all race groups.
+Creating static spatial maps of county-level mortality rates by race/ethnicity, visualized using spatial coordinates. These provide an overview of geographic variation across U.S. counties.
+
 ```{r}
@@
 
-###2.2 Combined spatial plot
+### 2.2 Combined spatial plot
 
 Combining all spatial race-specific maps into a multi-panel figure.
+
 ```{r}
@@
 ```
+
 These plots are best viewed as images outside of R due to size.
 
-Evaluating spatial clustering of Mortality Rates
-
+Static maps show clear geographic variation in hypertension-related CVD mortality.
+Black and White populations exhibit widespread mortality across the Southeast and Midwest, while Hispanic and Asian populations show more localized clusters, reflecting regional concentration and data availability.
+The highest mortality hotspots for American Indian and Alaska Native groups are found in the central U.S., consistent with known health disparities.
+
+While visual maps provide insight into geographic differences in mortality, they do not quantify the extent of spatial clustering.
+To formally assess spatial autocorrelation, we next apply Global Moran’s I and Local Indicators of Spatial Association (LISA) statistics.
+
+### Evaluating spatial clustering of Mortality Rates.
 ### 2.3: Global Moran’s I – All Races Combined
@@
 Running Global Moran's I test to evaluate overall spatial autocorrelation.
+
 ```{r spatial-weights global-morans-i, message=FALSE}
@@
 
-# Save result
+# Extract key statistics from Moran's I test
+moran_table <- tibble::tibble(
+  Region = "USA (All Counties)",
+  Moran_I = round(moran_result$estimate[["Moran I statistic"]], 4),
+  Expected_I = round(moran_result$estimate[["Expectation"]], 4),
+  SD = round(moran_result$estimate[["Variance"]], 4),
+  p_value = signif(moran_result$p.value, 3))
+
+moran_table
+
+# Save to results/tables
+saveRDS(moran_table, here::here("results", "tables", "global_morans_i_summary.rds"))
 saveRDS(moran_result, file = moran_path)
@@
 
-Global Moran's I: I = 0.0435, p < 0.001, this indicates weak but statistically significant spatial autocorrelation.
+The Global Moran’s I statistic (I = 0.0435, p < 0.001) indicates weak but statistically significant positive spatial autocorrelation in hypertension-related CVD mortality at the county level.
 There is weak-to-moderate positive spatial autocorrelation in mortality rates across U.S.
 counties — counties with high (or low) mortality tend to be near others with similarly high (or low) values.
-Although the magnitude is small (I = 0.0435), the pattern is considered highly statistically significant based on the p-value.
+Although the magnitude is small (I = 0.0435), the pattern is considered statistically significant based on the p-value.
+This suggests that counties with high mortality rates tend to cluster geographically, rather than being randomly distributed.
+The presence of spatial autocorrelation supports further analysis using Local Moran’s I to identify specific high- and low-mortality clusters.
+
+After confirming statistically significant spatial autocorrelation at the national level, we use Local Moran’s I to identify potential clusters of high or low mortality at the county level.
 
@@
 Running Local Moran's I (LISA) to detect local spatial clusters.
+
 ```{r lisa-analysis, message=FALSE}
@@
 
-Creating map of statistically significant LISA clusters (p < 0.05).
+Creating map of statistically significant LISA clusters (p \< 0.05).
+
 ```{r}
@@
 ```
+
 Classifying counties based on LISA cluster type using z-scores.
@@
   "Low-High (Outlier)" = "purple",
-  "Not Significant" = "gray85"
-)
+  "Not Significant" = "gray85")
 
@@
     subtitle = "High–High = Hotspot, Low–Low = Coldspot, Others = Outliers",
-    color = "Cluster Type"
-  ) +
+    color = "Cluster Type") +
   coord_equal() +
@@
   plot = lisa_cluster_plot,
-  width = 10, height = 6
-)
-
-```
+  width = 10, height = 6)
+
+```
+
 Local Moran’s I was computed on the full national dataset using 5-nearest-neighbor spatial weights.
@@
 
-####Focusing at a State level
-Due to the size of the dataset it is challenging to work with the full dataset.
+#### Focusing at a State level Due to the size of the dataset it is challenging to work with the full dataset. 
 For convenience and because the SouthEast is historically known to have a high prevalence of cardiovasccular mortalities we will focus on Georgia for this next step.
@@
 Analyzing spatial clusters in Georgia using LISA.
+
 ```{r}
@@
   saveRDS(ga_knn_nb, file = ga_knn_path)
-  saveRDS(ga_lw, file = ga_lw_path)
-}
+  saveRDS(ga_lw, file = ga_lw_path)}
 
@@
   "Low-High (Outlier)" = "purple",
-  "Not Significant" = "gray80"
-)
+  "Not Significant" = "gray80")
 
@@
     subtitle = "High–High = Hotspot; Low–Low = Coldspot; Others = Outliers",
-    color = "Cluster Type"
-  ) +
+    color = "Cluster Type") +
   coord_equal() +
@@
   saveRDS(ms_knn_nb, file = ms_knn_path)
-  saveRDS(ms_lw, file = ms_lw_path)
-}
+  saveRDS(ms_lw, file = ms_lw_path)}
 
@@
       mortality_z < 0 & lag_z > 0 ~ "Low-High (Outlier)",
-      TRUE ~ NA_character_
-    )
-  )
+      TRUE ~ NA_character_))
 
@@
 
-ggsave(
-  filename = here::here("products", "manuscript", "supplement", "ms_lisa_cluster_plot.png"),
+ggsave(filename = here::here("products", "manuscript", "supplement", "ms_lisa_cluster_plot.png"),
   plot = ms_lisa_plot,
   width = 8,
-  height = 6
-)
+  height = 6)
 
@@
   plot = tx_lisa_plot,
-  width = 9, height = 6
-)
+  width = 9, height = 6)
 ```
@@
 ```
+
 the LISA plot shows spatial patterns across all available data, not just one year
 
-Compile Global Moran’s I test results from the USA and selected high-mortality states (GA, MS, TX, KY).
-This table summarizes the spatial autocorrelation of mortality rates.
-
-2.5 Global Moran's I Summary Table
-
-```{r}
-# Already saved or placed earlier i
+The Local Moran’s I analysis detected significant spatial clustering of mortality rates across U.S.
+counties.
+Nationally, over 40% of counties showed statistically significant clustering at p < 0.05.
+However, most of these represented spatial outliers—counties with high (or low) mortality rates surrounded by contrasting neighbors—rather than concentrated hotspots.
+Only a small number of “High–High” clusters (hotspots) were identified, indicating that extreme spatial concentrations of high mortality are rare.
+
+Cluster types varied across states:
+
+In Georgia, clustering was driven primarily by High–Low outliers and a few Low–Low coldspots, especially in the northern region.
+
+Mississippi showed a similar pattern, with isolated High–Low counties in the central and southern parts of the state.
+
+In Texas, counties in the western and southern regions showed a mix of Low–Low coldspots and High–Low outliers, with no evidence of strong regional hotspots.
+
+Kentucky had a slightly denser distribution of significant clusters, including more scattered High–Low and Low–Low counties.
+A few High–High hotspots were detected but did not form a continuous cluster.
+
+Local spatial analysis (LISA) identified a small number of statistically significant clusters in Kentucky, including high-mortality counties adjacent to other high-mortality areas.
+These patterns support the presence of localized mortality hotspots.
+
+Across all states, the most consistent pattern was the presence of outlier counties—areas with mortality rates that diverged significantly from their neighbors.
+
+Compile Global Moran’s I test results from the USA and selected high-mortality states (GA, MS, TX, KY). This table summarizes the spatial autocorrelation of mortality rates.
+
+### 2.5 Global Moran's I Summary Table
+
+```{r}
+# Already saved or placed earlier
 extract_moran_values <- function(obj, geo) {
@@
     heading.subtitle.font.size = px(12),
-    column_labels.font.weight = "bold"
-  )
+    column_labels.font.weight = "bold")
 
@@
 A summary table of Global Moran’s I statistics for the full U.S.
-dataset and selected states (GA, MS, TX, KY) is shown below, indicating statistically significant spatial autocorrelation in all geographies examined (p < 0.001).
-
-Local Indicators of Spatial Association (LISA) – Statistically Significant Clustering of Mortality Rates in Kentucky.
-This map highlights counties with statistically significant spatial clustering of hypertension-related cardiovascular disease (CVD) mortality rates, as detected by Local Moran’s I.
-Each point represents a county centroid.
-Counties shown in red are spatial clusters where both the county and its neighbors have high mortality values (high-high) or display significant spatial autocorrelation (e.g., high-low or low-high).
-Spatial relationships were calculated using 5-nearest-neighbor spatial weights based on county centroids.
-This analysis includes data from all available years and demographic subgroups; therefore, significance reflects elevated spatial correlation within any combination of groupings, not just overall rates.
-
-Local spatial analysis (LISA) identified a small number of statistically significant clusters in Kentucky, including high-mortality counties adjacent to other high-mortality areas.
-These patterns support the presence of localized mortality hotspots and may inform targeted investigation or intervention efforts.
-
-##Aim 3: Demographic Disparities
+dataset and selected states (GA, MS, TX, KY) is shown below, indicating statistically significant spatial autocorrelation in all geographies examined (p \< 0.001).
+
+Global Moran’s I statistics were statistically significant across all regions analyzed.
+The USA showed weak but significant positive spatial autocorrelation (I = 0.0435, p < 0.001).
+Georgia, Mississippi, Texas, and Kentucky also had low Moran’s I values (0.0168–0.0297), but all were statistically significant (p < 0.001), confirming that mortality rates were not randomly distributed in space.
+These findings support further investigation of local spatial clustering using LISA.
+
+## Aim 3: Demographic Disparities
 
@@
 
+# Extract model fit metrics
+model_fit_race_sex <- broom::glance(glm_model_interaction) %>%
+  dplyr::mutate(model = "GLM: Race × Sex × Year")
+
+# Save for results
+saveRDS(model_fit_race_sex, here::here("results", "tables", "model_fit_race_sex.rds"))
+
+# Create a table of slopes for each race-sex group 
+# Format slope estimates for manuscript table
+slope_table_race_sex_fmt <- as.data.frame(summary(em_trends_race_sex)) %>%
+  dplyr::mutate(
+    Estimate = round(Year.trend, 3),
+    CI = paste0(round(lower.CL, 2), "–", round(upper.CL, 2)),
+    SE = round(SE, 3)) %>%
+  dplyr::select(Race = race_ethnicity, Sex = sex, Estimate, SE, CI)
+
+# Save nicely formatted table
+saveRDS(slope_table_race_sex_fmt, here::here("results", "tables", "slope_table_race_sex_fmt.rds"))
+
+#Display table:
+#Note: p.value of 0.00e+00 is actually p < 2.2e-16
+view(slope_table_race_sex_fmt)
+
 # Compare slopes between race/sex groups
@@
 saveRDS(slope_contrast_race_sex, file = here::here("results", "output", "slope_contrast_race_sex.rds"))
+
+# Create a table comparing slopewise pairwise comparison between groups for each race-sex group
+# Format race/sex slope difference table
+slope_contrast_race_sex_fmt <- as.data.frame(summary(slope_contrast_race_sex, infer = c(TRUE, TRUE))) %>%
+  dplyr::mutate(
+    Estimate = round(estimate, 3),
+    CI = paste0(round(lower.CL, 2), "–", round(upper.CL, 2)),
+    SE = round(SE, 3),
+    p_value = signif(p.value, 3)) %>%
+  dplyr::select(Contrast = contrast, Estimate, SE, CI, p_value)
+
+#Sort table results by significance (p-value)
+slope_contrast_race_sex_fmt_sorted <- slope_contrast_race_sex_fmt %>%
+  arrange(p_value, desc(abs(Estimate)))
+
+# Save to results folder
+saveRDS(slope_contrast_race_sex_fmt, here::here("results", "output", "slope_contrast_race_sex_fmt.rds"))
+saveRDS(slope_contrast_race_sex_fmt_sorted,here::here("results", "output", "slope_contrast_race_sex_sorted.rds"))
+
+# view results from the sorted table 
+slope_contrast_race_sex_fmt_sorted
 
@@
 
-###3.2 Demogrpahic disparities by Age group
+-   Black adults, particularly Black females, showed the steepest and most consistently negative slopes (−3.69 deaths per 100,000 per year, 95% CI: −4.07 to −3.30), indicating large annual declines in mortality.
+These trends were significantly different from nearly all other groups, with pairwise differences of over 7 points when compared to American Indian and Alaska Native (AIAN) individuals, and around 5.7 points compared to White adults (p \< 2.2e-16).
+-   In contrast, American Indian and Alaska Native adults had the most positive slopes (\~+3.38 per year), suggesting a consistent increase in hypertension-related CVD mortality over time.
+These trends were significantly higher than all other groups, especially when compared to Black, Hispanic, and Asian and Pacific Islander subgroups.
+-   White adults had moderately positive slope estimates (\~+1.98 per year), significantly higher than those for Asian, Black, and Hispanic adults.
+-Hispanic adults and Asian and Pacific Islander adults showed relatively flat to slightly negative trends.
+These groups had annual change estimates that were not significantly different from zero, with confidence intervals crossing or approaching zero.
+Differences from the “Overall” group estimates were also generally not statistically significant.
+-   Several race/sex combinations (e.g., Black Female vs.
+White Male; AIAN Male vs.
+Hispanic Male) yielded statistically significant contrasts, with p \< 2e-16 for many pairwise comparisons, further reinforcing the existence of sharp disparities in trend slopes.
+These findings suggest that while national mortality may be decreasing on average, substantial heterogeneity by race and sex persists.
+The increasing trends in mortality among AIAN individuals are particularly concerning, in contrast to strong improvements among Black adults.
+
+While the above results illustrate broad disparities across race and sex, they do not account for variation by age.
+We next fit an age-adjusted model to stratify mortality trends by age group and explore how disparities unfold differently for younger and older adults.
+
+### 3.2 Demogrpahic disparities by Age group
 
@@
 Fit the Interaction Model
+
 ```{r}
@@
                           data = county_race_sex, family = gaussian())
-  saveRDS(glm_race_sex_age, glm_path)
-}
+  saveRDS(glm_race_sex_age, glm_path)}
 
@@
 Get Estimated Slopes Stratified by Age Group
+
 ```{r}
@@
 
-###3.3 Pairwise Comparisons Within Each Age Group
+Estimated annual mortality trends stratified by race/ethnicity and sex (adjusted for age group) revealed substantial variation.
+Black males and females had the steepest declines in mortality, with yearly slopes of –3.685 deaths per 100,000 (SE = 0.197).
+In contrast, American Indian and Alaska Native males and females showed the largest increases, with an estimated yearly change of +3.380 (SE = 0.471).
+White males and females experienced increasing trends (Estimate = +1.983, SE = 0.116), while mortality trends for Hispanic adults were nearly flat (Estimate = –0.030, SE = 0.259).
+Asian and Pacific Islander adults showed modest decreases (Estimate = –0.918, SE = 0.361).
+
+### 3.3 Pairwise Comparisons Within Each Age Group
 
@@
 
+# Convert to dataframe
+slope_contrast_race_sex_age_fmt <- as.data.frame(summary(slope_contrast_race_sex_age, infer = c(TRUE, TRUE))) %>%
+  dplyr::mutate(
+    Estimate = round(estimate, 3),
+    SE = round(SE, 3),
+    CI = paste0(round(lower.CL, 2), "–", round(upper.CL, 2)),
+    p_value = signif(p.value, 3)) %>%
+  dplyr::select(age_group, Contrast = contrast, Estimate, SE, CI, p_value)
+
+# Sort by p-value and effect size 
+slope_contrast_race_sex_age_sorted <- slope_contrast_race_sex_age_fmt %>%
+  dplyr::arrange(p_value, dplyr::desc(abs(Estimate)))
+
+# Save output
+saveRDS(slope_contrast_race_sex_age_sorted,
+        here::here("results", "output", "slope_contrast_race_sex_by_age_sorted.rds"))
+
+# View sorted results
+view(slope_contrast_race_sex_age_sorted)
+
 ```
@@
 
-###3.4: Visualize Slope Contrasts
-Top slope differences (absolute magnitude) were visualized to highlight which subgroup comparisons showed the greatest disparities. Significant differences (p < 0.05) are shown in red.
+The pairwise comparisons of estimated yearly trends within each age group revealed highly consistent results across both age strata (35–64 and 65+).
+Black adults had the steepest declines in hypertension-related CVD mortality, significantly differing from all other race-sex groups (e.g., Black vs.
+Overall Male: −7.00 per year, SE = 0.12, p \< 2.2e-16).
+In contrast, American Indian and Alaska Native adults consistently showed the most rapid increases in mortality rates (e.g., AI/AN vs.
+Black Overall: +6.95 per year, SE = 0.36, p \< 2.2e-16).
+Differences between sexes within race/ethnicity were not estimable in this output, with many comparisons yielding NA.
+These results reinforce the magnitude and consistency of racial disparities seen in the earlier pooled model.
+
+###3.4: Visualize Slope Contrasts Top slope differences (absolute magnitude) were visualized to highlight which subgroup comparisons showed the greatest disparities.
+Significant differences (p \< 0.05) are shown in red.
+
 ```{r}
 # Summarize the contrast object
-contrast_summary <- as.data.frame(summary(slope_contrast_race_sex_age)) %>%
+contrast_summary <- as.data.frame(summary(slope_contrast_race_sex_age, infer = c(TRUE, TRUE))) %>%
   filter(!is.na(estimate)) %>%  # remove any NAs if they exist
@@
 # Save and display
-ggsave(
-  here::here("results", "output", "pairwise_slope_contrasts_filtered.png"),
+ggsave(here::here("results", "output", "pairwise_slope_contrasts_filtered.png"),
   plot = contrast_plot, width = 11, height = 8)
@@
                    height = 0.3, color = "black", alpha = 0.7) +
-    labs(
-      title = title,
+    labs(title = title,
       x = "Estimated Annual Change in Mortality Rate",
-      y = "Race and Ethnicity"
-    ) +
+      y = "Race and Ethnicity") +
     theme_minimal(base_size = 13) +
@@
 “Overall” groups include both sexes; sex-specific estimates were excluded where modeling was not feasible.
+
+To complement the full slope estimates by group (3.2), we visualized the top pairwise contrasts from 3.3 by effect size and statistical significance.
+The plot highlights that the largest and most statistically significant differences in annual mortality trends consistently involved comparisons with Black adults, particularly between Black individuals and males or American Indian/Alaska Native groups.
+These patterns were consistent across both age groups.
+Many subgroup comparisons were statistically significant (red), with several slope differences exceeding ±4 deaths per 100,000 annually.
 
@@
 
-##Aim 4: Predict County-level Mortality rates
-
-We will make use of some models to try to predict mortality rates based on the different potential predictors (location, race, sex, age-group and year)
-
-####Data Preparation for modeling
+Overview of the Summary Statistics before modeling.
+
+### Descriptive Statistics of County-Level Mortality Rates by Demographics
+
+To contextualize the above modeling results, we summarize baseline mortality distributions across race/sex/age combinations.
+
+```{r}
+# Summary by race, sex, and age group
+mortality_summary_table <- county_race_sex %>%
+  drop_na(mortalities, race_ethnicity, sex, age_group) %>%
+  group_by(race_ethnicity, sex, age_group) %>%
+  summarise(
+    n = n(),
+    mean_mortality = round(mean(mortalities, na.rm = TRUE), 2),
+    sd = round(sd(mortalities, na.rm = TRUE), 2),
+    min = round(min(mortalities, na.rm = TRUE), 2),
+    max = round(max(mortalities, na.rm = TRUE), 2),
+    .groups = "drop"
+  )
+
+# Save for manuscript
+saveRDS(mortality_summary_table, here::here("results", "tables", "mortality_summary_table.rds"))
+
+mortality_summary_table
+```
+
+We provide baseline descriptive statistics of hypertension-related CVD mortality to contextualize the modeled slope differences.
+Among adults aged 35–64, Black individuals had the highest mean mortality rate (92.2 per 100,000), followed by American Indian and Alaska Native (AI/AN) adults (46.2), with Asian and Pacific Islander adults showing the lowest (17.7).
+These patterns were consistent in the 65+ age group, where Black adults again had the highest average mortality (662.6), while Asian and Pacific Islander adults had the lowest (308.7).
+The wide ranges and standard deviations suggest substantial geographic heterogeneity, particularly in older adults.
+These differences reinforce the need for stratified modeling.
+
+## Aim 4: Predict County-level Mortality rates
+
+In this section, we evaluate predictive models for estimating county-level hypertension-related cardiovascular disease (CVD) mortality based on demographic and temporal predictors, including race/ethnicity, sex, age group, and year.
+
+### 4.1 Data Preparation for modeling
+
+We begin by loading the cleaned dataset, removing missing values, and splitting the data into training and testing sets stratified by mortality outcomes.
 
@@
 ```
+
+## 4.2 Random Forest Modeling
+
+We first fit a Random Forest model, tuning hyperparameters with cross-validation and evaluating performance on the held-out test set.
 
@@
 
-Hyperparamter tuning and fitting a random forest model:
-We will tune a random forest with 10-Fold cross-validation:
-
+Hyperparamter tuning and fitting a random forest model: We will tune a random forest with 10-Fold cross-validation:
+
+### 4.2.1 Hyperparameter Tuning
+
+To optimize the number of predictors (`mtry`) and minimum node size (`min_n`), we perform 10-fold cross-validation over a regular grid of parameter values.
 # Define path to save tuning results
@@
 ## ---- rf_tune_and_fit_if_guarded, message=TRUE, warning=FALSE -------------------
-
-library(tidymodels)
-library(vip)
-library(here)
 
@@
 
-  message("✅ All RF files found. Loading saved versions...")
+  message(" All RF files found. Loading saved versions...")
 
@@
 } else {
-  message("🔁 Fitting Random Forest from scratch...")
+  message("Fitting Random Forest from scratch...")
 
@@
     min_n(range = c(5, 25)),
-    levels = 3
-  )
+    levels = 3)
   saveRDS(rf_grid, rf_grid_path)
@@
     metrics = metric_set(rmse, rsq, mae),
-    control = control_grid(save_pred = TRUE)
-  )
+    control = control_grid(save_pred = TRUE))
 
@@
     list(model = rf_final_model, metrics = rf_metrics, vip = rf_vip),
-    file = rf_slim_path
-  )
-
-  message("✅ RF slim model and all outputs saved.")}
+    file = rf_slim_path)
+
+  message("RF slim model and all outputs saved.")}
 
@@
 ```{r}
-## ---- temporary_rf_runtime_log, message=TRUE, warning=FALSE -------------------
-
-# Optional: Log run time and session info for audit trail
-start_time <- Sys.time()
-Sys.sleep(1)  # simulate runtime block
-end_time <- Sys.time()
-
-writeLines(as.character(end_time - start_time), here::here("results", "output", "rf_runtime_duration.txt"))
-writeLines(capture.output(sessionInfo()), here::here("results", "output", "rf_session_info.txt"))
-
-message("⏱️ Runtime and session info saved.")
-
-```
-
-```{r}
 #Slim version already loads rf_metrics, so no need to reload
@@
 
-from this we see that the best model based on the lowest RMSE is the first one listed (Preprocessor1_Model6) with all 4 predictors used (mtry = 4), 15 observations per leaf in the tree ( min_n = 15), and RMSE = 115.22.
+The optimal Random Forest model was selected with mtry = 4 and min_n = 15, yielding the lowest cross-validated RMSE (115.2).
+Performance differences between parameter combinations were minor when mtry = 4, but models with fewer predictors showed substantially higher RMSE values.
 
@@
 
+The lowest cross-validated RMSE (115.2) was observed for the model with mtry = 4 and min_n = 15.
+RMSE decreased substantially with increasing mtry, while changes in min_n had minimal impact, as seen in the tuning plot.
+
+### 4.2.2 Model Performance
+
 ```{r}
@@
 
-  message("Final model fit and results saved.")
-}
+  message("Final model fit and results saved.")}
 
@@
     Model = "Random Forest",
-    Notes = "Tuned with 4 predictors (mtry = 4), min_n = 15"
-  ) %>%
+    Notes = "Tuned with 4 predictors (mtry = 4), min_n = 15") %>%
   select(Model, RMSE = rmse, R2 = rsq, MAE = mae, Notes)
@@
 ```
+
+The final random forest model achieved an RMSE of 115.3, MAE of 66.9, and R² of 0.748 on the test set. These metrics indicate strong predictive accuracy and model fit.
 
@@
     x = "Observed Mortality",
-    y = "Predicted Mortality"
-  ) +
+    y = "Predicted Mortality") +
   theme_minimal()
@@
 ```
+
+The predicted vs. observed plot shows close alignment with the 45-degree line for most observations, though some spread is visible at higher mortality levels.
+
+### 4.2.3 Variable Importance
+
+Variable importance was assessed using impurity-based measures from the Random Forest model to identify top predictors of county-level mortality.
 
@@
 
+Variable importance analysis indicated that age_group_Ages.65.years was by far the most influential predictor in the random forest model.
+Other variables with moderate contributions included race_ethnicity_Black and race_ethnicity_Hispanic.
+
 A Random Forest model was tuned and evaluated to predict county-level hypertension-related cardiovascular mortality.
@@
 This model will serve as a benchmark against which we compare other algorithms, including regularized regression (e.g., LASSO).
-
-####LASSO model
+While random forest provided strong predictive performance, we next evaluated a regularized linear model (LASSO) to assess interpretability and feature selection trade-offs.
+
+### 4.3 LASSO Regression
+
+We next fit a LASSO regression model using the `glmnet` engine. This model applies regularization to reduce overfitting and perform variable selection.
 
 Define the LASSO model specification
+
 ```{r}
@@
 
+### 4.3.1 Tuning and Coefficients
+
+We tune the penalty parameter (`lambda`) to identify the optimal level of shrinkage. Coefficients from the final model are extracted to assess variable importance.
+
 Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.
@@
 ## ---- lasso_tune_and_fit_if_guarded, message=TRUE, warning=FALSE -------------------
-
-library(tidymodels)
-library(vip)
-library(here)
 
@@
 
-  message("✅ All LASSO files found. Loading saved versions...")
+  message("All LASSO files found. Loading saved versions...")
 
@@
 } else {
-  message("🔁 Fitting LASSO model from scratch...")
+  message(" Fitting LASSO model from scratch...")
 
@@
     list(model = final_lasso_model, metrics = lasso_metrics, coefficients = lasso_coeffs, vip = lasso_vip),
-    file = slim_lasso_path
-  )
-
-  message("✅ Final LASSO model fitted and slim version saved.")}
+    file = slim_lasso_path)
+
+  message("Final LASSO model fitted and slim version saved.")}
 ```
@@
 The full RDS file is saved in the large-files folder which will not automatically upload to github.
+
 ```{r}
@@
 
-message("✅ LASSO RMSE summary with CI estimates saved.")
+message("LASSO RMSE summary with CI estimates saved.")
 ```
@@
 
-```{r}
-## ---- temporary_lasso_runtime_log, message=TRUE, warning=FALSE -------------------
-
-start_time <- Sys.time()
-Sys.sleep(1)  # Simulated delay
-end_time <- Sys.time()
-
-writeLines(as.character(end_time - start_time), here::here("results", "output", "lasso_runtime_duration.txt"))
-writeLines(capture.output(sessionInfo()), here::here("results", "output", "lasso_session_info.txt"))
-
-message("⏱️ LASSO runtime and session info saved.")
-
-```
-
+LASSO tuning selected a penalty value of approximately 0.0001, minimizing RMSE at ~119.63.
+Coefficient shrinkage was applied, resulting in the exclusion of several predictors and aiding model interpretability.
+
+### 4.3.2 Model Performance
+
+The final LASSO model is evaluated on the test set and compared to the Random Forest model in terms of predictive accuracy and fit.
 ```{r}
@@
   saveRDS(lasso_test_predictions, lasso_pred_path)
-  saveRDS(lasso_test_metrics, lasso_metrics_path)
-}
+  saveRDS(lasso_test_metrics, lasso_metrics_path)}
 
@@
 ```
+
+The LASSO model achieved an RMSE of 119.54, MAE of 70.66, and R² of 0.727 on the test set.
+While performance was slightly below that of the random forest model, LASSO retained strong explanatory power and offers the advantage of interpretability through variable selection.
 
@@
 ```
+
+LASSO is a linear model with regularization — it struggles to capture high-order nonlinear patterns.
+The horizontal “bands” in predicted values suggest that LASSO may be underfitting in high or low mortality counties.
+While LASSO remains interpretable, random forest outperforms it in predictive power and flexibility, especially in a dataset with many categorical and potentially non-linear features.
 
@@
 
-The LASSO regression model achieved a Root Mean Squared Error (RMSE) of ~119.63, with an R-squared of ~0.726, indicating a reasonably strong fit for predicting county-level hypertension-related mortality.
-Compared to the random forest model (RMSE ~115.30, R² ~0.748), LASSO performs slightly worse, but still retains good explanatory power.
+LASSO regression achieved slightly lower performance than random forest, with an RMSE of ~119.6 and R² of ~0.726indicating a reasonably strong fit for predicting county-level hypertension-related mortality.
+Compared to the random forest model, LASSO performs slightly worse, but still retains good explanatory power.
+While LASSO is less flexible than random forest, it offers interpretability through regularization and feature selection.
+These trade-offs motivate the direct comparison presented in the next section.
 
@@
 
-Plot Predicted vs Observed Mortality
-
-```{r}
-ggplot(lasso_test_predictions, aes(x = mortalities, y = .pred)) +
-  geom_point(alpha = 0.4, color = "steelblue") +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
-  labs(
-    title = "LASSO Regression: Predicted vs Observed Mortality",
-    x = "Observed Mortality",
-    y = "Predicted Mortality"
-  ) +
-  theme_minimal(base_size = 13)
-
-```
-
-LASSO is a linear model with regularization — it struggles to capture high-order nonlinear patterns.
-The horizontal “bands” in predicted values suggest that LASSO may be underfitting in high or low mortality counties.
-While LASSO remains interpretable, random forest outperforms it in predictive power and flexibility, especially in a dataset with many categorical and potentially non-linear features.
-
-###Comparing models
+### 4.4 Model Comparison
+
+To directly compare model performance, we generated side-by-side summaries of RMSE, R², and MAE for both random forest and LASSO regression using the test dataset.
+Confidence intervals are provided where available.
 
@@
   upper <- round(mean + 1.96 * se, 2)
-  paste0(round(mean, 2), " [", lower, ", ", upper, "]")
-}
+  paste0(round(mean, 2), " [", lower, ", ", upper, "]")}
 
@@
     select(model, .metric, ci) %>%
-    pivot_wider(names_from = .metric, values_from = ci)
-}
+    pivot_wider(names_from = .metric, values_from = ci)}
 
@@
     title = "Model Performance Summary",
-    subtitle = "Test Set Evaluation for Random Forest and LASSO"
-  ) %>%
+    subtitle = "Test Set Evaluation for Random Forest and LASSO") %>%
   cols_label(
@@
     rsq = "R²",
-    mae = "MAE"
-  ) %>%
+    mae = "MAE") %>%
   fmt_markdown(columns = everything()) %>%
@@
 
-Aim 4 Summary
+Random forest outperformed LASSO on all metrics, with lower RMSE (115.3 vs.
+119.5), lower MAE (66.9 vs.
+70.7), and higher R² (0.748 vs.
+0.727).
+While both models demonstrated good predictive accuracy, random forest offered stronger overall performance.
+
+Aim 4 Summary:
 We compared two predictive models for county-level hypertension-related CVD mortality: a random forest model and a LASSO regression model.

### 2025-04-19 products/manuscript/Manuscript.qmd
Part 5 Additions

--- 2025-03-17 products/manuscript/Manuscript.qmd
@@
 
-* First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678) 
-* Second Author$^{3, *}$
-* Third Autor$^{3, \dagger}$
-* Last Author$^{1, 2, \land}$
+-   First Author: Shaun van den Hurk$^{1,2,*}$
 
@@
 
-$\land$ Corresponding author: some@email.com
+$\land$ Corresponding author: shaun.vandenhurk\@uga.edu
 
@@
 # Summary/Abstract
-Lifestyle diseases such as cardiovascular disease make up the highest causes of death in developed nations such as the United States.
-This project focuses on hypertension-related cardiovascular disease and hopes to help provide a greater understandign of the dynamics and patterns that may exist with hypertension and cardiovascualr diseased related mortality in the United States.
-The project seeks to evaluate data from the CDC on the rates of hypertension-related cardiovascular disease mortality and the associated demographic factors for this collected data, including county, age group, race/ethnicity, and sex.
-The data represents information collected from 2000 to 2019.
-This will be used to see if there are any potential patterns or correlation between the rate of mortality and these other demographic factors within the data.
+
+Cardiovascular disease (CVD) remains a leading cause of mortality in the United States, with hypertension recognized as a major modifiable contributor.
+This study investigates patterns and predictors of hypertension-related CVD mortality among U.S.
+adults aged 35 and older using county-level data from 2000 to 2019.
+We explored temporal, demographic, and geographic disparities and evaluated the predictive performance of demographic and temporal factors using machine learning models.
+Descriptive analyses revealed persistent and pronounced disparities in mortality across race, sex, and age groups.
+Black and American Indian/Alaska Native populations experienced the highest mortality rates, particularly among adults aged 65 and older.
+Temporal modeling showed diverging trends across subgroups, with Black adults seeing the steepest declines, while mortality increased among males and American Indian/Alaska Native populations.
+Spatial clustering analyses identified modest but statistically significant autocorrelation, with localized high-mortality clusters in the Southeastern U.S.
+Finally, predictive modeling using Random Forest and LASSO regression demonstrated that demographic and temporal variables could explain a substantial proportion of county-level variation in mortality.
+The Random Forest model achieved an R² of 0.748 and identified race/ethnicity and year as the most influential predictors.
+These findings highlight the enduring influence of demographic and geographic factors on cardiovascular health outcomes and support the use of predictive models to inform future public health strategies.
 
@@
 
-Lifestyle diseases such as cardiovascular disease make up the highest causes of death in developed nations such as the United States.
-This data analysis hopes to evaluate the rates of hypertension-related cardiovascular disease mortality in adults in the United States and trends that might be associated with demographic and geographic factors.
+Lifestyle diseases such as cardiovascular disease (CVD) represent some
+of the most significant causes of death globally, both in low-income
+regions and high-income countries such as the United States of America
+(USA). Cardiovascular diseases are a broad umbrella term including
+different clinical manifestations such as hypertension, stroke, and
+ischemic heart disease. Cardiovascular diseases were cited as the
+leading cause of death globally in 2022 (..1..), in the USA they were
+responsible for one in every five deaths nationally (...2.).
+Hypertension is commonly identified as a significant risk factor for
+multiple cardiovascular conditions, and one that is modifiable, and thus
+represents a good metric for potential public health interventions
+(...3.). However, not all demographics appear to be equally affected by
+hypertension and CVD, with differences in rates of CVD and hypertension
+in people from different races, sexes, age groups, geographic location,
+income levels, and other groups in previous studies (..4..). A deeper
+understanding of these factors helps us to get a deeper understanding of
+the multifactorial nature of CVD and also helps guide effective public
+health interventions. Additionally, it is important to understand if
+there have been any differences in these rates of CVD and hypertension
+over time, and if there are particular geographic hotspots. This is
+essential as it allows public health officials to measure the impact of
+previous intervention strategies and to ensure that resources
+strategically target areas of concern.
 
@@
 
-The dataset was collected from the CDC website and the information was sourced from the National Vital Statistics System.
-The dataset contains rates and trends in the hypertension-related cardiovascular disease (CVD) death rates.
-The study period is from 2000 to 2019, with two main intervals displayed in the data 2000-2009, and 2010-2019.
-The data is further distributed by county, age group (36-64 years and 65 years and older), race/ethnicity, and sex.
+The dataset was collected from the CDC website and the information was
+sourced from the National Vital Statistics System (..5...). The dataset
+contains rates and trends in the hypertension-related cardiovascular
+disease (CVD) death rates. The study period is from 2000 to 2019, with
+two main intervals displayed in the data 2000-2009, and 2010-2019. The
+data is further distributed by county, age group (35-64 years and 65
+years and older), race/ethnicity, and sex.
+
+(1= )
+
+(2= Virani SS, et al. Heart Disease and Stroke Statistics—2020 Update: A
+Report From the American Heart Association. *Circulation*.
+2020;141:e139–e596.)
+
+(3= Whelton PK, et al. 2017 ACC/AHA Guideline for the Prevention,
+Detection, Evaluation, and Management of High Blood Pressure in Adults.
+*Hypertension*. 2018;71:e13–e115)
+
+(Mahmood SS, et al. The Framingham Heart Study and the epidemiology of
+cardiovascular disease: a historical perspective. *Lancet*.
+2014;383(9921):999–1008.)
+
+(4= Yusuf S, et al. Modifiable risk factors, cardiovascular disease, and
+mortality in 155,722 individuals from 21 high-income, middle-income, and
+low-income countries (PURE): a prospective cohort study. *Lancet*.
+2020;395(10226):795–808.)
+
+(undefined
+
+(5= at the following url:
+https://data.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Hypertension-related-Cardiovas/uc9k-vc2j/about_data.
+)
 
@@
 
-- Hypertension-related cardiovascular disease mortalities are higher in select demographics (age groups, race/ethnicities and sex), and thus these demographics show correlated with mortalities rates.
+This data analysis aimed to evaluate the rates of hypertension-related
+cardiovascular disease mortality in adults in the United States based on
+county-level data and to identify trends that might be associated with
+demographic and geographic factors. Furthermore, we aim to identify
+factors that can be used in predictive models to estimate mortality
+trends.
 
-- There are clear regional/geographic distributions associated with hypertension-related cardiovascular disease mortalities.
+#### Study Aims:
+
+\- Aim 1: Assess national temporal trends in hypertension-related CVD
+mortality, stratified by age group.
+
+\- Aim 2: Identify spatial clustering of county-level mortality.
+
+\- Aim 3: Evaluate disparities in mortality trends by race/ethnicity and
+sex, stratified by age group.
+
+\- Aim 4: Build predictive models using demographic and geographic
+features to estimate mortality trends.
+
+#### Hypotheses:
+
+\- Mortality trends vary significantly over time and across age groups.
+
+\- County-level mortality exhibits spatial clustering, reflecting
+geographic disparities.
+
+\- Temporal trends differ significantly across race and sex, with
+widening inequities over time.
+
+\- Predictive models can identify key demographic and spatial factors
+associated with higher mortality trends.
 
@@
 
-This data was downloaded from the CDC online database at the following url: https://data.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Hypertension-related-Cardiovas/uc9k-vc2j/about_data.
-A csv file was downloaded which was almost 300MB in size.
-The initial document/dataset contained 1103872 observations across 24 variables.
-There were some initial processing steps performed to evaluate and delete some of the unnecessary columns and to save this data to a .rds document to reduce the size.
-This was all outlined in the "processingcode" document, which will not run automatically because the Excel (csv) file will not be available due to size concerns.
+We obtained the dataset from the CDC’s online portal, sourced from the National Vital Statistics System [5].
+The raw CSV file was approximately 300MB in size and contained 1,103,872 rows and 24 columns.
+Initial preprocessing was conducted in R to remove redundant variables, reduce file size, and prepare the dataset for analysis.There were some initial processing steps performed to evaluate and delete some of the unnecessary columns and to save this data to an RDS
+document to reduce the size. This was all outlined in the "processingcode" document (and README file), which will not run
+automatically because the Excel (csv) file will not be available due to
+size concerns with Github uploads (available on request). This process is fully documented in the processingcode.R and processingfile-v1.qmd files within the project repository.
 
-The intial exploration and cleaning and removal of the redundant columns (columns with repeated single values without any additional information).
-Some of the other columns were renamed to have the name of the column reflect what these single value columns had contained instead.
-The columns were checked to see what they contained and to see how many unique entries there were and based upon this columns were deleted and others renamed.
+Non-informative or repetitive columns (e.g., columns with a single repeated value or metadata such as DataSource, Topic, Geographic_Level) were removed.
+Columns describing demographic stratification were renamed for clarity: Stratification1 was renamed to age_group, Stratification2 to race_ethnicity, and Stratification3 to sex.
+The primary outcome variable, Data_Value, was renamed mortalities.
 
-Deleted columns were: DataSource, Topic, Class, StratificationCategory1/2/3, Data_Value_Footnote, Data_Value_Footnote_Symbol, Geographic_Level, TopicID
+All rows with missing values in any of the key variables (mortality rate, year, age group, race/ethnicity, sex, and geographic coordinates) were removed.
+The resulting cleaned dataset contained 572,572 rows and 14 variables.
+Summary statistics and exploratory inspection (e.g., using the skimr package) confirmed the data's readiness for analysis.
+The dataset included both numeric variables (e.g., mortality rates, year, coordinates) and categorical variables (e.g., sex, race/ethnicity, age group, county).
 
-Renamed columns: 
-Stratification1 column to age_group, Stratification2 column to race_ethnicity, Stratification3 column to sex, Data_Value to mortalities.
+Age groups were categorized into two broad strata: 35–64 years and 65 years and older.
+While this limits deeper evaluation, it facilitates comparisons across the two broad groups and supports stratified modeling approaches.
+The cleaned dataset was saved in .rds format for improved performance and integration with downstream analysis scripts.
 
-After the initial cleaning the data was evaluated again and some further small cleaning and processing steps were performed (code is visible in the processingfile-v1.qmd file) to reduce the number of rows by removing NA values.
-The resulting dataframe contains 572572 observations and 14 variables.
+All cleaning and preprocessing steps are reproducible and are logged in the repository.
 
-Our summaries and the skimr function help to provide a good overview for further evaluation of the data. 
-We can see that our dataset contains a combination of numeric and character data. 
-The numeric data includes the actual incidence rates of the hypertension-related cardiovascular disease death rates, as well as the confidence intervals that were used, and geographic coordinates.
-Our character data includes the race, sex, age group, location data and the year.
+## Statistical Analysis
 
-This character data will be used as predictors of the mortality data as we break down these different variables and look for potential patterns and associations.
-The primary outcome of interest is the mortality rates but the important potential predictors are the location, year, race/ethnicity, and sex.
-AGe groups only provide two broad categories, namely 35 to 65 years old and 65 years and older, which is a bit limited but will allow for comparisons between these two groups.
+Statistical analysis was structured according to the study’s four primary aims and was conducted using R (version 4.3.1).
+All code is available in the project repository within the statistical-analysis-v2-slim.qmd script.
+Analyses incorporated regression modeling, spatial statistics, and supervised machine learning to assess temporal trends, demographic disparities, spatial clustering, and predictive performance.
 
-## Statistical analysis
+To evaluate national temporal trends in hypertension-related cardiovascular disease (CVD) mortality from 2000 to 2019 (Aim 1), we began by fitting a linear regression model to estimate the overall change in mortality over time.
+We then used generalized linear models (GLMs) to assess stratified time trends by race/ethnicity, sex, and age group.
+Interaction terms such as Year × race/ethnicity and Year × age group were included to evaluate whether trends varied between subgroups.
+We estimated annual mortality slopes for each demographic category using the emtrends() function from the emmeans package and conducted pairwise comparisons using the contrast() function to formally test for differences between groups.
+
+For spatial analysis (Aim 2), we examined county-level clustering of mortality rates using spatial statistics.
+County geometries were created by converting longitude and latitude coordinates into spatial features using the sf package.
+Spatial weights were computed using 5-nearest-neighbor relationships.
+Global spatial autocorrelation was assessed using Moran’s I statistic.
+Local clustering patterns were evaluated using Local Indicators of Spatial Association (LISA) via the spdep package.
+Local spatial relationships were visualized using standardized mortality z-scores and classified into high–high, low–low, or spatial outlier clusters based on the combination of mortality values and neighboring spatial lags.
+
+To evaluate demographic disparities in temporal trends (Aim 3), we extended the GLMs used in Aim 1 by including three-way interactions between year, race/ethnicity, and sex, and stratified analyses by age group.
+Estimated marginal trends were obtained separately for adults aged 35–64 and those aged 65 and older.
+Annual slope estimates and their confidence intervals were extracted and compared across race–sex combinations.
+Pairwise comparisons of slopes were again conducted using emmeans, with the goal of quantifying and visualizing disparities in mortality trends across key demographic intersections.
+
+In the final aim (Aim 4), we developed predictive models to estimate county-level mortality rates using demographic and temporal features.
+Two modeling approaches were implemented: a random forest model using the ranger engine and a LASSO regression model using the glmnet engine.
+The modeling dataset included year, age group, race/ethnicity, and sex as predictors.
+Categorical variables were dummy-encoded using the recipes package, and numeric predictors were normalized; zero-variance predictors were removed prior to modeling.
+Data were split into training (80%) and test (20%) sets, stratified by mortality outcome.
+Within the training set, 10-fold cross-validation was used to tune model hyperparameters.
+For random forests, tuning was performed over a regular grid of mtry and min_n values.
+LASSO regression was tuned using a log-scaled penalty grid to identify optimal shrinkage parameters.
+Final model performance was evaluated on the test set using RMSE (root mean squared error), MAE (mean absolute error), and R² (coefficient of determination).
+Confidence intervals for these metrics were not computed due to computational constraints.
+Model interpretability was further assessed using variable importance plots for the random forest model and non-zero coefficient selection in LASSO.
+
+This multi-step analytical framework enabled us to assess trends, identify disparities, detect geographic clustering, and test predictive capacity using a consistent set of demographic and temporal variables across all analytic aims.
 
@@
 
-## Exploratory/Descriptive analysis
+## Descriptive and Exploratory Analysis
+We began by examining overall patterns in hypertension-related cardiovascular disease (CVD) mortality among U.S.
+adults aged 35 and older from 2000 to 2019 (Aim 1; Hypothesis 1).
+Summary statistics revealed large disparities in mortality rates across racial/ethnic and age groups.
+For adults aged 35–64, Black individuals had the highest mean county-level mortality rate (92.2 per 100,000), followed by American Indian/Alaska Native (46.2), and White (32.9).
+In the 65+ age group, Black adults again had the highest mean mortality (662.6), with American Indian/Alaska Native (451.9) and White (398.3) also experiencing elevated rates.
 
-A summary table was produced using the skimr function which is printed below.
+Boxplots of mortality by race/ethnicity and sex (Figures below) confirmed these disparities, showing right-skewed distributions with particularly high upper ranges for Black and American Indian/Alaska Native adults.
+Males had consistently higher mortality distributions than females across all groups.
 
-```{r}
-skim_summary_hypertens_filtered <- readRDS(here::here("data","processed-data", "skim_summary_hypertens_filtered.rds"))
-# Print the skim summary
-print(skim_summary_hypertens_filtered)
+Temporal visualizations indicated divergent patterns between subgroups.
+For example, while some populations experienced declines in mortality rates, others showed upward or stable trends.
+These descriptive findings supported formal modeling to assess temporal and demographic patterns.
 
-```
-
-Our summary data shows that the mean of the hypertension cardiovascular-diseases mortalities is around 203, with a standard deviation of 223.
-
-A histogram was made of the mortalities columns since this is the main (only) column of interest with numerical data.
-
-```{r}
-revised_mortalities_histogram <- readRDS(here::here("data","processed-data", "revised_mortalities_histogram.rds"))
-# Print the mortalities histogram
-plot(revised_mortalities_histogram)
-```
-
-Our histogram depicts the hypertension-related cardiovasculat disease mortalities in the dataset, showing the number of mortalities and the frequency (count) of sprecific mortality values.
-The mortality counts are clustered towards the lower end of the scale, with a strong right-skew (long tail to the right/higher values).
-Peak mortalities are below 100, and the total values likely account for the second peak.
-
-Boxplots were created to evaluate the mortalities by race and by sex.
-
-```{r}
+```{r, echo=FALSE, message=FALSE}
 mortality_race_boxplot <- image_read(here::here("products", "manuscript", "supplement", "mortality_by_race.png"))
@@
 
-## Key Findings and Next Steps
-
-- The distribution of mortality rates is right-skewed, suggesting potential transformations before modeling.
-- Mortality rates vary markedly by race and sex, justifying stratified analysis.
-- A proof-of-concept temporal trend was created for a single county (Calhoun, GA) using Location ID, ensuring unique selection.
-- Temporal trend indicate a general increase the selected county, which should be modeled.
-- All processed data has been saved into a single RDS file (`final_processed_data.rds`) for use in statistical analysis.
-
 ## Basic statistical analysis
 
-Spatial plots of mortalities per geographic regions were generated for each race and compared on one plot.
-Two generalized linear models were evaluated, the first examined how time, race/ethnicity, and sex (plot below) influence mortality rates, and the second tested whether the effect of time varied by race/ethnicity.
+### Temporal Trends in Mortality 
+At the national level, a linear regression of overall mortality against year showed a small but statistically significant increase in mortality from 2000 to 2019 (slope = 1.40, p < 0.001).
+However, stratified models revealed considerable variation between groups.
 
-Overall, black individuals had the highest predcted mortality rates from the generalized linear models, followed by white and native american populations.
-Black individuals also had the highest hypertension-related mortalities on the boxplot follwoed by white individuals.
-Females appeared to have lower mortality rates than males across all races.
+A generalized linear model (GLM) incorporating race/ethnicity and sex predictors (Aim 1.2; Hypotheses 1 and 3) confirmed that Black individuals had the highest predicted mortality rates, followed by White and American Indian/Alaska Native populations.
+Males had significantly higher predicted rates than females across all racial/ethnic groups.
 
-Race, sex and time all seem to be predictors of mortality. We see differences in trends over time for different racial groups.
+An interaction model between year and race (Aim 1.3) revealed statistically significant divergence in trends.
+Mortality among Black individuals declined sharply over time (slope = –7.06, p < 0.001), while American Indian/Alaska Native (slope = –3.41) and Hispanic populations showed shallower or increasing trends.
 
-Our spatial clustering has revealed geographic clustering of high-mortality regions, with high densities in the Eastern regions of the country.
-There also appears to be a higher density of mortalities from people in the race/ethnic group Black in certain regions.
-Asian and Pacific Islander groups had relatively lower mortality rates and geographic clustering.
+Interaction between year and age group (Aim 1.4) demonstrated that adults aged 65+ experienced a significantly steeper increase in mortality over time compared to those aged 35–64 (interaction slope = +0.22, p < 0.001).
+Estimated marginal trends (Aim 1.5) further confirmed this: the slope among older adults was 1.26 per year (95% CI: 1.18–1.34), versus 1.04 (95% CI: 0.96–1.12) in the younger group (p = 0.0002).
 
-```{r}
+```{r, echo=FALSE, message=FALSE}
 #Print the glm plot
@@
 
-```{r}
-#Print the combined spatial plot
+### Geographic Distribution and Clustering 
+Spatial visualizations demonstrated considerable geographic variation in mortality by race/ethnicity (Aim 2.1–2.2).
+Elevated mortality was most prominent among Black individuals, particularly in counties across the Southeast and Midwest.
+Asian and Pacific Islander populations displayed lower and more geographically diffuse mortality rates.
+
+```{r, echo=FALSE, message=FALSE}
 combined_spatial_plot <- image_read(here::here("products", "manuscript", "supplement", "combined_spatial_plot.png"))
@@
 
-## Full analysis
+Global spatial clustering was modest but statistically significant for the national dataset (Moran’s I = 0.0435, p < 0.001), with similar low levels of spatial autocorrelation in state-level analyses: Georgia (I = 0.0297), Mississippi (0.0289), Texas (0.0168), and Kentucky (0.0227), all p < 0.001 (Aim 2.3).
 
-_
+Local spatial clustering (Aim 2.4) was visualized using Local Indicators of Spatial Association (LISA).
+Statistically significant high–high clusters were identified in counties across parts of the Southeast, but the majority of counties showed no local spatial clustering.
+This suggests that while some mortality hotspots exist, broad demographic and structural drivers likely contribute more to the observed geographic distribution than tightly bounded spatial clusters.
+
+### Demographic Disparities in Temporal Trends
+We assessed demographic disparities in mortality trends across race/ethnicity and sex within age groups (Aim 3.1–3.3; Hypothesis 3).
+Among adults aged 35–64, Black individuals experienced the steepest annual decline in mortality (–3.68), while American Indian/Alaska Native individuals showed the largest increases (+3.38).
+Overall, males experienced steeper increases than females.
+
+Pairwise comparisons showed the greatest differences between Overall Male and Black Overall groups (+7.00 per year), followed by contrasts involving American Indian/Alaska Native, White, and Asian/Pacific Islander groups.
+These results indicate persistent and widening disparities in mortality trends, particularly among marginalized populations.
+
+```{r, echo=FALSE, message=FALSE}
+readRDS(here::here("results", "tables", "slope_table_agegroup_fmt.rds"))
+readRDS(here::here("results", "tables", "slope_table_race_sex_fmt.rds"))
+```
+
+Additional visualizations from Aim 3.4 reinforce these findings:
+
+```{r, echo=FALSE, message=FALSE}
+slope_plot_35_64 <- image_read(here::here("results", "figures", "slope_plot_35_64.png"))
+print(slope_plot_35_64)
+
+slope_plot_65_plus <- image_read(here::here("results", "figures", "slope_plot_65_plus.png"))
+print(slope_plot_65_plus)
+
+```
+
+### Predictive Modeling of Mortality Rates
+
+To evaluate the ability of demographic and temporal factors to predict hypertension-related CVD mortality at the county level, we developed and compared two supervised machine learning models: a Random Forest and a LASSO regression model (Aim 4; Hypothesis 4).
+Both models were trained on 80% of the dataset and evaluated on a held-out 20% test set, with 10-fold cross-validation used during training to optimize hyperparameters and reduce overfitting.
+
+The Random Forest model demonstrated strong predictive accuracy.
+The best-performing model used all four available predictors (year, age group, race/ethnicity, and sex) with a minimum node size of 15.
+On the test set, this model achieved a root mean squared error (RMSE) of 115.3, a mean absolute error (MAE) of 66.9, and an R² of 0.748, indicating that approximately 75% of the variance in county-level mortality was explained by the model.
+Variable importance scores revealed that race/ethnicity and year were the most influential predictors, followed by age group and sex.
+The predicted versus observed plot showed generally good alignment, though some dispersion at the extremes suggested modest underfitting in counties with very high or low mortality.
+
+```{r, echo=FALSE, message=FALSE}
+model_performance_summary_table <- readRDS(here::here("results", "tables", "model_performance_summary.rds"))
+model_performance_summary_table
+
+```
+
+```{r, echo=FALSE, message=FALSE}
+rf_pred_plot <- image_read(here::here("results", "figures", "rf_pred_vs_obs.png"))
+print(rf_pred_plot)
+
+rf_vip_plot <- image_read(here::here("results", "figures", "rf_variable_importance.png"))
+print(rf_vip_plot)
+
+```
+
+The LASSO regression model provided a more interpretable, though slightly less accurate, alternative.
+On the same test set, the LASSO model yielded an RMSE of 119.5, an MAE of 70.7, and an R² of 0.726.
+This model performed coefficient shrinkage and variable selection, shrinking several dummy-coded predictors toward zero.
+While this feature enhances interpretability and highlights influential variables, it also limits the model’s flexibility to capture complex non-linear patterns.
+The predicted versus observed plot for the LASSO model revealed horizontal banding patterns.
+This reflects the impact of dummy-encoded categorical variables — many counties shared identical predictor combinations, resulting in repeated predictions and horizontal clustering in plots.
+These effects were more pronounced in LASSO than in the Random Forest, which is better able to handle categorical interactions.
+
+```{r, echo=FALSE, message=FALSE}
+lasso_pred_plot <- image_read(here::here("results", "figures", "lasso_pred_vs_obs.png"))
+print(lasso_pred_plot)
+
+```
+
+Despite these differences, both models demonstrated solid performance in capturing patterns in the data.
+The Random Forest provided better predictive accuracy, while the LASSO model offered greater clarity in identifying relevant features.
+Taken together, these results indicate that demographic and temporal factors are strong predictors of mortality variation at the county level, supporting Hypothesis 4 and offering tools for future public health risk modeling.
+
+\_
 
@@
 
+We examined patterns and predictors of hypertension-related cardiovascular disease (CVD) mortality among U.S.
+adults aged 35 and older between 2000 and 2019.
+Through a combination of exploratory visualization, statistical modeling, spatial clustering, and machine learning, we found consistent evidence of disparities in mortality across time, geography, and demographic subgroups.
+These findings underscore the ongoing burden of hypertension-related CVD in the United States and highlight areas for targeted intervention.
+
+National trends revealed that overall mortality rates remained relatively stable over the study period.
+However, stratified models showed considerable variation between groups.
+Black individuals experienced the highest mortality rates throughout, while American Indian/Alaska Native populations showed a concerning upward trend in recent years.
+Adults aged 65 and older had significantly steeper increases in mortality rates than younger adults.
+These findings are consistent with previous national surveillance reports that have identified persistent disparities in CVD burden across racial/ethnic and age groups [2,4].
+
+Spatial analyses detected statistically significant but modest spatial clustering of mortality at the national and state levels.
+While global Moran’s I statistics confirmed the presence of spatial autocorrelation, the magnitude was small (I = 0.0435 nationally).
+Local spatial clustering (LISA) identified limited high-mortality hotspots, primarily concentrated in the Southeast.
+These results suggest that broader structural and demographic factors—rather than localized geographic effects—may be the primary drivers of hypertension-related CVD mortality.
+This is in line with previous studies indicating that health disparities are often influenced more by social determinants and systemic inequities than by geographic proximity alone [4,6].
+
+Demographic subgroup analyses revealed widening inequities in mortality trends.
+Among adults aged 35–64, Black individuals experienced the steepest annual declines, suggesting some gains in this group.
+However, these improvements occurred in parallel with rising mortality among American Indian/Alaska Native individuals and males, who showed the largest annual increases.
+Among adults aged 65 and older, mortality increased for nearly all groups.
+These divergent trends highlight the importance of considering intersecting axes of inequality—such as race, sex, and age—in public health surveillance.
+They also support prior research showing that demographic-specific interventions are necessary to effectively reduce disparities in chronic disease outcomes [3,4,7].
+
+In the predictive modeling component, we assessed whether demographic and temporal variables could accurately predict county-level mortality.
+The Random Forest model performed best, explaining approximately 75% of the variation in mortality rates (R² = 0.748), with an RMSE of 115.3.
+Race/ethnicity and year were the most influential predictors.
+The LASSO regression model, while slightly less accurate (R² = 0.726), offered greater interpretability by selecting a reduced subset of predictors and shrinking non-influential coefficients to zero.
+These results suggest that relatively simple demographic and temporal features can provide strong predictive power, and that different modeling approaches offer complementary strengths.
+The Random Forest’s flexibility allowed for more accurate predictions, while LASSO highlighted core factors driving variability.
+
+Notably, both models showed horizontal banding in predicted vs.
+observed plots—an artifact of dummy-encoded categorical variables and repeated predictor combinations across counties.
+This highlights a broader methodological challenge in population-level modeling using grouped categorical features.
+Although machine learning approaches can improve predictive performance, careful attention must be paid to encoding strategies and interpretability.
+
+These findings contribute to a growing body of evidence emphasizing the importance of demographic stratification and predictive modeling in public health research.
+In particular, our analysis supports previous work demonstrating the utility of ensemble and regularized regression models in epidemiologic prediction tasks [8,9].
+Importantly, our results reinforce that even well-known risk factors—such as age, sex, race/ethnicity, and time—continue to explain a large proportion of the variance in hypertension-related CVD mortality and may be sufficient for initial risk stratification at the population level.
+
 ## Strengths and Limitations
+This study benefits from the use of a comprehensive, nationwide dataset covering nearly two decades of mortality records.
+The integration of statistical and machine learning methods allowed us to not only describe mortality patterns but also assess their predictability.
+However, several limitations should be noted.
+First, the analysis was limited to county-level aggregated data, which may obscure within-county heterogeneity and limit causal inference.
+Second, although we used age-standardized mortality rates, residual confounding by comorbidities, access to care, socioeconomic status, or other contextual factors cannot be excluded.
+Lastly, while our models achieved high predictive accuracy, they were not designed to support individual-level predictions or establish causal relationships.
+
+## Implications and Future Directions
+Our findings highlight important and persistent disparities in hypertension-related CVD mortality that should inform ongoing public health surveillance and intervention efforts.
+Geographic clustering, although weak, may still guide resource allocation to high-burden areas.
+Predictive models developed here can support proactive health planning, particularly when integrated with local surveillance systems.
+Future research should explore integrating additional variables—such as healthcare access, environmental exposures, and social determinants of health—to further refine predictions and address upstream drivers of CVD disparities.
 
@@
 
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
+This study provides a comprehensive analysis of hypertension-related cardiovascular disease (CVD) mortality among U.S.
+adults aged 35 and older from 2000 to 2019, using national county-level data.
+Our findings reveal that while overall mortality trends remained relatively stable at the national level, significant disparities persist across demographic and geographic groups.
+Black and American Indian/Alaska Native populations consistently experienced higher mortality rates, with the steepest increases observed among older adults and males.
+Spatial analysis identified weak but statistically significant clustering, particularly in the Southeastern United States, though most variation was driven by broader structural and demographic patterns rather than tightly localized effects.
+
+Temporal modeling demonstrated that subgroup-specific trends diverged substantially, suggesting that recent gains in cardiovascular health have not been equitably distributed.
+Our predictive modeling further supports this, with demographic features—particularly race/ethnicity and age—emerging as dominant predictors of county-level mortality risk.
+Random Forest models outperformed LASSO regression in accuracy, but both approaches reinforced the critical role of social and demographic context in shaping health outcomes.
+
+Taken together, these findings underscore the urgent need for targeted public health strategies that account for longstanding inequities in cardiovascular risk.
+Future research should integrate social determinants of health, healthcare access, and policy-level factors to better understand the drivers of these disparities and to design interventions that are both equitable and effective.
+
+Note that this cited reference will show up at the end of the document,
+the reference formatting is determined by the CSL file specified in the
+YAML header. Many more style files for almost any journal [are
+available](https://www.zotero.org/styles). You also specify the location
+of your bibtex reference file in the YAML. You can call your reference
+file anything you like.
 

### 2025-04-25 project-review-template-cann-van-den-Hurk.qmd
Project Review

---
title: "Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States" Review 
author: Natalie. Cann
date: April 24, 2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States

Name of project author(s): Shaun van den Hurk

Name of project reviewer: Natalie Cann

# Instructions

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

Your background is great! I like how you listed out several diseases that fall under the cardiovascular disease umbrella.

I really liked how you shared that cardiovascular disease was the global leading cause of death in 2022 - this really conveys the importance of researching cardiovascular disease to the reader.

Some statements about previous work is included, which is a good start. I think it would be a good idea to add some of the findings from this reference in your background.

There was some redunancy with wording - take a look at things again at some point before submitting and see if you can add some variation. For example:
"A deeper understanding of these factors helps us to get a deeper understanding of..." - I would recommend not repeating the phrase "a deeper understanding" twice in this sentence.

### Summary assessment (PICK ONE, DELETE THE OTHERS)
* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

Same here with the abbreviation for USA/U.S. - I think you can just refer to the United States as USA since you introduced it as that previously. I could be wrong though - I was never an expert at english :)

Maybe for aim 4 you could get a bit more specific - I would recommend stating what you're predicting here. Same with the hypothesis for this aim.

Overall, I really like the way that you structured your aims and then had the hypotheses right below!

### Summary assessment
* question/hypotheses somewhat explained

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

In the second sentence, you have spelled out CVD as "cardiovascular disease" and included its abbreviation in parenthesis. I think since you introduced this abbreviation in your background, you can just use the abbrevation from here on out.

### Summary assessment
* source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

I was not able to render either of your files (processingcode.R and processingfile-v1.qmd). I see that you included instructions in your readme.md file, which is really helpful. However, when I tried to run the processingfile-v1.qmd file, I ran into issues (data/raw-data/cleaned_hypertens_adult_mortality.rds', probable reason 'No such file or directory') because I couldn't get the cleaned data from the first file. I do not see this file in the raw data folder.

Your data cleaning looks good - you did a great job of explaining your code! I like how you explained it both above/below code chunks in your qmd file AND within the code chunks using comments (#). I do this too :)

This section of manuscript looks great - it reads very well!

I also was not able to render the eda.qmd file for a similar reason (/data/processed-data/final_processed_data.rds', probable reason 'No such file or directory') - I don't seen a final_processed_data.rds file in the processed data folder.

Overall, your eda looks good and was very thorough from what I can see! I really like how you included a small summary section at the end of your file - this is a great idea.

### Summary assessment
* essentially no weaknesses in wrangling and exploratory component
(only weakness I can think of is the reproducibility - but this is assessed in a different part of the review)

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

I couldn't render the statistical-analysis-v2-slim.qmd file due to the data file not being present (/data/processed-data/eda_outputs.rds', probable reason 'No such file or directory').

Your analysis looks good! You did a great job at interpreting things below code chunks. It was so neat how you included code to fill in maps - I've always wondered how people have done this! This is very impressive!

I think you did a great job walking the reader through your analysis in your manuscript!

### Summary assessment
* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

Your graphs look awesome - however, I would make the tables using the gt() package to make them look a bit more profressional. Then, you can save them and add them to your manuscript in the same way you added your graphs.

### Summary assessment
* results are presented ok, with room for improvement

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

Discussion of results in manuscript is very impressive!!

Several strengths and limitations are addressed.

### Summary assessment
* strong, complete and clear discussion

## Further comments

1. Add instructions to your main README file to make it clear in what order the code should be run. For example, processing code > eda code > analysis code > manuscript code > supplementary code

2. Add references in form of citation manager like Zotero (I struggled with this and found that this video helped a lot: https://www.youtube.com/watch?v=zuuOYjE8m98

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

Project is very organized and the readme folders are helpful/appropriate. I would recommend removing the 2020-mckay files from the references folder.

### Summary assessment
* well structured

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

As explained above, you did a great job explaining your code both in code chunks and above and below your code chunks.

### Summary assessment
* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

I discussed the issues above with reproducibility. The data files appear to be missing. I'm sure once you find a way to add them everything will run smoothly though!

Or it is entirely possible this is an issue on my end!

### Summary assessment
* major parts not reproducible

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

Your project was very detailed and interesting! I'm glad I got assigned to review yours. Well done!

### Summary assessment
* strong level of thorougness

## Further comments

Very impressed with your project and all of the hard work you have put into it!

The biggest thing that needs work in my opinion is the reproducibility. I have included the errors that R's console was giving me when I tried to run your code. Hopefully this is helpful!

### 2025-05-02 code/analysis-code/statistical-analysis-v2-slim.qmd
Changes for final product

@@
 ---
+
+This should be the third and final file to run when reproducing the full analysis. Thereafter you will get results in the manuscript/supplementary.
 
@@
 library(forcats)
+library(webshot2)  # for gt image export
+library(ggpubr)
 
@@
 print(lm_results_year)
-saveRDS(lm_results_year, file = here::here("products", "manuscript", "supplement", "lm_results_year.rds"))
+invisible(saveRDS(lm_results_year, file = here::here("results", "output", "statistical-analysis", "lm_results_year.rds")))
 
@@
 # Save as table
-saveRDS(model_fit_lm, here::here("results", "tables", "model_fit_national.rds"))
+invisible(saveRDS(model_fit_lm, here::here("results", "tables", "model_fit_national.rds")))
 
@@
 print(lm_plot_year)
-ggsave(here::here("products", "manuscript", "supplement", "lm_plot_year.png"), plot = lm_plot_year)
+invisible(ggsave(here::here("results", "figures", "lm_plot_year.png"), plot = lm_plot_year))
 
@@
 
-The linear regression model estimated a modest but statistically significant annual increase in national hypertension-related CVD mortality rates between 2000 and 2019 (slope = 1.40, p < 2.2e-16).
+The linear regression model estimated a modest but statistically significant annual increase in national hypertension-related CVD mortality rates between 2000 and 2019 (slope = 1.40, p \< 2.2e-16).
 However, as shown in the plot, the fitted trend line is visually flat due to the scale and large variation in county-level rates.
@@
 print(glm_results)
-saveRDS(glm_results, file = here::here("products", "manuscript", "supplement", "glm_results.rds"))
+invisible(saveRDS(glm_results, file = here::here("results", "output", "statistical-analysis", "glm_results.rds")))
 
@@
   print(glm_race_sex_plot)
-  ggsave(here::here("products", "manuscript", "supplement", "glm_race_sex_plot.png"), plot = glm_race_sex_plot)
+  invisible(ggsave(here::here("results", "figures", "glm_race_sex_plot.png"), plot = glm_race_sex_plot))
+
+```
+
+```{r}
+# Examining multiple predictors
+glm_model <- glm(mortalities ~ Year + race_ethnicity + sex, data = county_race_sex, family = gaussian())
+glm_results <- broom::tidy(glm_model)
+
+# Display and save the GLM results
+print(glm_results)
+invisible(saveRDS(glm_results, file = here::here("results", "output", "statistical-analysis", "glm_results.rds")))
+
+# Add predicted values
+county_race_sex$predicted_mortalities <- predict(glm_model, type = "response")
+
+# Filter out "Overall" race_ethnicity values before plotting
+plot_data <- county_race_sex %>%
+  filter(race_ethnicity != "Overall")
+
+# Create polished plot
+glm_race_sex_plot <- ggplot(plot_data, aes(x = Year, y = predicted_mortalities, color = race_ethnicity)) +
+  geom_line(size = 1.2) +
+  theme_minimal(base_size = 12) +
+  labs(
+    title = "GLM Predictions: Mortality Rates by Year and Race/Ethnicity",
+    x = "Year",
+    y = "Predicted Mortality Rate (per 100,000)",
+    color = "Race/Ethnicity"
+  ) +
+  scale_x_continuous(
+    breaks = seq(min(plot_data$Year, na.rm = TRUE), max(plot_data$Year, na.rm = TRUE), by = 5)
+  ) +
+  scale_color_discrete(labels = function(x) str_wrap(x, width = 20)) +
+  theme(
+    legend.position = "right",
+    legend.text = element_text(size = 10),
+    legend.title = element_text(size = 11),
+    legend.key.size = unit(0.4, "cm"),
+    legend.spacing.y = unit(0.2, "cm"),
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    plot.title = element_text(face = "bold", size = 13, hjust = 0.5))
+
+# Display plot
+print(glm_race_sex_plot)
+
+# Save figure
+ggsave(
+  here::here("results", "figures", "glm_race_sex_plot.png"),
+  plot = glm_race_sex_plot,
+  width = 7.5,
+  height = 5.5,
+  dpi = 300)
+
+```
+
+```{r}
+
+# Fit GLM model (no interaction)
+glm_model <- glm(mortalities ~ Year + race_ethnicity + sex, data = county_race_sex, family = gaussian())
+
+# Save tidy model output
+glm_results <- broom::tidy(glm_model)
+print(glm_results)
+invisible(saveRDS(glm_results, file = here::here("results", "output", "statistical-analysis", "glm_results.rds")))
+
+# Predict mortality values from model
+county_race_sex$predicted_mortalities <- predict(glm_model, type = "response")
+
+# Remove "Overall" for plotting
+filtered_data <- county_race_sex %>%
+  filter(race_ethnicity != "Overall")
+
+# Summarize observed means
+observed_summary <- filtered_data %>%
+  group_by(Year, race_ethnicity) %>%
+  summarise(
+    observed_mean = mean(mortalities, na.rm = TRUE),
+    .groups = "drop")
+
+# Summarize predicted means
+predicted_summary <- filtered_data %>%
+  group_by(Year, race_ethnicity) %>%
+  summarise(
+    predicted_mean = mean(predicted_mortalities, na.rm = TRUE),
+    .groups = "drop")
+
+# Join observed and predicted for plotting
+plot_data <- left_join(observed_summary, predicted_summary, by = c("Year", "race_ethnicity"))
+
+# Create overlay plot
+glm_race_sex_plot <- ggplot(plot_data, aes(x = Year, color = race_ethnicity)) +
+  geom_point(aes(y = observed_mean), size = 1.5, alpha = 0.6) +
+  geom_line(aes(y = predicted_mean), size = 1.2) +
+  theme_minimal(base_size = 12) +
+  labs(
+    title = "Observed and Predicted Mortality Rates by Race/Ethnicity (2000–2019)",
+    x = "Year",
+    y = "Mortality Rate (per 100,000)",
+    color = "Race/Ethnicity"
+  ) +
+  scale_x_continuous(
+    breaks = seq(min(plot_data$Year, na.rm = TRUE), max(plot_data$Year, na.rm = TRUE), by = 5)
+  ) +
+  scale_color_discrete(labels = function(x) str_wrap(x, width = 20)) +
+  theme(
+    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
+    legend.position = "right",
+    legend.title = element_text(size = 11),
+    legend.text = element_text(size = 10),
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    legend.key.size = unit(0.4, "cm"),
+    legend.spacing.y = unit(0.2, "cm"))
+
+# Display
+print(glm_race_sex_plot)
+
+# Save plot (overwrite previous)
+ggsave(
+  here::here("results", "figures", "glm_race_sex_plot.png"),
+  plot = glm_race_sex_plot,
+  width = 7.5,
+  height = 5.5,
+  dpi = 300)
 
@@
 The GLM including race/ethnicity and sex as predictors highlights large disparities in predicted mortality rates.
-Black populations had the highest predicted rates, significantly higher than all other groups (e.g., β = 205.8, p < 2e-16).
+Black populations had the highest predicted rates, significantly higher than all other groups (e.g., β = 205.8, p \< 2e-16).
 Hispanic and Asian/Pacific Islander groups had lower mortality, though the Asian effect was not significant (p = 0.35).
-Males also showed significantly higher mortality than females (β = 20.3, p < 2e-16).
+Males also showed significantly higher mortality than females (β = 20.3, p \< 2e-16).
 Notably, the “Overall” category behaves inconsistently in plots due to it being modeled as a separate group rather than a true average, likely affected by unmodeled age variation.
@@
 
-saveRDS(glm_results_interaction, file = here::here("products", "manuscript", "supplement", "glm_results_interaction.rds"))
+invisible(saveRDS(glm_results_interaction, file = here::here("results", "output", "statistical-analysis", "glm_results_interaction.rds")))
 
@@
   scale_x_continuous(breaks = seq(min(county_race_sex$Year, na.rm = TRUE), max(county_race_sex$Year, na.rm = TRUE), by = 5)) +
-  theme(legend.position = "right", legend.text = element_text(size = 10), legend.key.size = unit(0.4, "cm"), legend.spacing.y = unit(0.2, "cm"), axis.text.x = element_text(angle = 45, hjust = 1))
+  theme(legend.position = "right", legend.text = element_text(size = 11), legend.key.size = unit(0.4, "cm"), legend.spacing.y = unit(0.2, "cm"), axis.text.x = element_text(angle = 45, hjust = 1))
 
 print(glm_interaction_plot)
-ggsave(here::here("products", "manuscript", "supplement", "glm_interaction_plot.png"), plot = glm_interaction_plot)
-
-saveRDS(county_race_sex, file = here::here("results", "output", "county_race_sex_with_preds.rds"))
+invisible(ggsave(here::here("results", "figures", "glm_interaction_plot.png"), plot = glm_interaction_plot))
+
+invisible(saveRDS(county_race_sex, file = here::here("results", "output", "statistical-analysis", "county_race_sex_with_preds.rds")))
+```
+
+```{r}
+# Explore whether interactions exist between year and race/sex
+glm_model_interaction <- glm(
+  mortalities ~ Year * race_ethnicity + sex,
+  data = county_race_sex,
+  family = gaussian())
+
+glm_results_interaction <- broom::tidy(glm_model_interaction)
+print(glm_results_interaction)
+
+invisible(saveRDS(
+  glm_results_interaction,
+  file = here::here("results", "output", "statistical-analysis", "glm_results_interaction.rds")))
+
+# Add interaction predictions
+county_race_sex$predicted_mortalities_interaction <- predict(glm_model_interaction, type = "response")
+
+# Filter out "Overall" before plotting
+plot_data <- county_race_sex %>%
+  filter(race_ethnicity != "Overall")
+
+# Create clean interaction plot
+glm_interaction_plot <- ggplot(plot_data, aes(x = Year, y = predicted_mortalities_interaction, color = race_ethnicity)) +
+  geom_line(size = 1.2) +
+  theme_minimal(base_size = 12) +
+  labs(
+    title = "GLM Interaction Model: Mortality Trends by Year and Race/Ethnicity",
+    x = "Year",
+    y = "Predicted Mortality Rate (per 100,000)",
+    color = "Race/Ethnicity"
+  ) +
+  scale_x_continuous(
+    breaks = seq(min(plot_data$Year, na.rm = TRUE), max(plot_data$Year, na.rm = TRUE), by = 5)
+  ) +
+  scale_color_discrete(labels = function(x) str_wrap(x, width = 20)) +
+  theme(
+    legend.position = "right",
+    legend.text = element_text(size = 10),
+    legend.title = element_text(size = 11),
+    legend.key.size = unit(0.4, "cm"),
+    legend.spacing.y = unit(0.2, "cm"),
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    plot.title = element_text(size = 13, face = "bold", hjust = 0.5))
+
+# Display plot
+print(glm_interaction_plot)
+
+# Save cleaned image
+ggsave(
+  here::here("results", "figures", "glm_interaction_plot.png"),
+  plot = glm_interaction_plot,
+  width = 7.5,
+  height = 5.5,
+  dpi = 300)
+
+# Save updated data
+invisible(saveRDS(county_race_sex,file = here::here("results", "output", "statistical-analysis", "county_race_sex_with_preds.rds")))
+
+```
+
+```{r}
+# Fit interaction model
+glm_model_interaction <- glm(
+  mortalities ~ Year * race_ethnicity + sex,
+  data = county_race_sex,
+  family = gaussian())
+
+# Save model results
+glm_results_interaction <- broom::tidy(glm_model_interaction)
+print(glm_results_interaction)
+invisible(saveRDS(
+  glm_results_interaction,
+  file = here::here("results", "output", "statistical-analysis", "glm_results_interaction.rds")))
+
+# Add predicted values to dataset
+county_race_sex$predicted_mortalities_interaction <- predict(glm_model_interaction, type = "response")
+
+# Filter out "Overall" before plotting
+filtered_data <- county_race_sex %>%
+  filter(race_ethnicity != "Overall")
+
+# Summarize observed data by group and year
+observed_summary <- filtered_data %>%
+  group_by(Year, race_ethnicity) %>%
+  summarise(observed_mean = mean(mortalities, na.rm = TRUE), .groups = "drop")
+
+# Summarize predicted means
+predicted_summary <- filtered_data %>%
+  group_by(Year, race_ethnicity) %>%
+  summarise(predicted_mean = mean(predicted_mortalities_interaction, na.rm = TRUE), .groups = "drop")
+
+# Combine both summaries
+plot_data <- left_join(observed_summary, predicted_summary, by = c("Year", "race_ethnicity"))
+
+# Create overlay plot
+glm_interaction_plot <- ggplot(plot_data, aes(x = Year, color = race_ethnicity)) +
+  geom_point(aes(y = observed_mean), size = 1.5, alpha = 0.6) +  # Observed
+  geom_line(aes(y = predicted_mean), size = 1.2) +               # Predicted
+  theme_minimal(base_size = 12) +
+  labs(
+    title = "Observed and Predicted Mortality Rates by Race/Ethnicity (2000–2019)",
+    x = "Year",
+    y = "Mortality Rate (per 100,000)",
+    color = "Race/Ethnicity"
+  ) +
+  scale_x_continuous(
+    breaks = seq(min(plot_data$Year, na.rm = TRUE), max(plot_data$Year, na.rm = TRUE), by = 5)
+  ) +
+  scale_color_discrete(labels = function(x) str_wrap(x, width = 20)) +
+  theme(
+    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
+    legend.position = "right",
+    legend.title = element_text(size = 11),
+    legend.text = element_text(size = 10),
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    legend.key.size = unit(0.4, "cm"),
+    legend.spacing.y = unit(0.2, "cm"))
+
+# Display updated plot
+print(glm_interaction_plot)
+
+# Save image
+ggsave(
+  here::here("results", "figures", "glm_interaction_plot.png"),
+  plot = glm_interaction_plot,
+  width = 7.5,
+  height = 5.5,
+  dpi = 300)
+
+# Save updated dataset
+invisible(saveRDS(
+  county_race_sex,
+  file = here::here("results", "output", "statistical-analysis", "county_race_sex_with_preds.rds")))
+
 ```
@@
 Including an interaction between year and race revealed significantly divergent temporal trends.
-Mortality rates among Black individuals showed the steepest annual decline (β = –7.06, p < 0.001), while American Indian and Alaska Native (AI/AN) populations exhibited the largest annual increase (β for year = 3.38, p < 0.001, with positive main effect and significant negative interactions for other groups).
+Mortality rates among Black individuals showed the steepest annual decline (β = –7.06, p \< 0.001), while American Indian and Alaska Native (AI/AN) populations exhibited the largest annual increase (β for year = 3.38, p \< 0.001, with positive main effect and significant negative interactions for other groups).
 Asian and Pacific Islander, Hispanic, and White populations also showed declining trends, though to a lesser extent.
@@
 # Save results
-saveRDS(glm_age_group_results, file = here::here("products", "manuscript", "supplement", "glm_age_group_results.rds"))
+invisible(saveRDS(glm_age_group_results, file = here::here("results", "output", "statistical-analysis", "glm_age_group_results.rds")))
 
@@
 # Save as table
-saveRDS(model_fit_agegroup, here::here("results", "tables", "model_fit_agegroup.rds"))
+invisible(saveRDS(model_fit_agegroup, here::here("results", "tables", "model_fit_agegroup.rds")))
 
@@
 # Save plot
-ggsave(here::here("products", "manuscript", "supplement", "glm_age_plot.png"), 
-       plot = glm_age_plot, width = 9, height = 6)
+invisible(ggsave(here::here("results", "figures", "glm_age_plot.png"), 
+       plot = glm_age_plot, width = 9, height = 6))
 
@@
 Including an interaction between year and age group revealed that mortality rates are substantially higher in the 65+ age group across all years.
-While both age groups show increasing mortality over time, the slope is significantly steeper for older adults (interaction term β = 0.224, p < 0.001), suggesting accelerated increases in mortality among those aged 65 and older.
+While both age groups show increasing mortality over time, the slope is significantly steeper for older adults (interaction term β = 0.224, p \< 0.001), suggesting accelerated increases in mortality among those aged 65 and older.
 
@@
 # Save as manuscript-ready table
-saveRDS(slope_table_age_fmt, here::here("results", "tables", "slope_table_agegroup_fmt.rds"))
-
+invisible(saveRDS(slope_table_age_fmt, here::here("results", "tables", "slope_table_agegroup_fmt.rds")))
+
+# Create and save polished PNG table for manuscript
+slope_table_agegroup_gt <- slope_table_age_fmt %>%
+  gt() %>%
+  tab_header(
+    title = "Estimated Annual Increase in Mortality by Age Group",
+    subtitle = "Derived from GLM with Year × Age Group Interaction"
+  ) %>%
+  fmt_number(columns = c(Estimate, SE), decimals = 3) %>%
+  cols_label(
+    `Age Group` = "Age Group",
+    Estimate = "Slope",
+    SE = "Standard Error",
+    CI = "95% CI"
+  ) %>%
+  opt_table_font(font = google_font("Roboto")) %>%
+  tab_options(table.width = pct(80),             # shrink width
+  table.font.size = "12",         # shrink font
+  data_row.padding = px(2),          # reduce row height
+  column_labels.font.weight = "bold")
+
+gtsave(
+  slope_table_agegroup_gt,
+  filename = here::here("results", "tables", "slope_table_agegroup.png"))
+
+slope_table_agegroup_gt
+
+```
+
+```{r}
 # compare trends between groups (pairwise contrast)
@@
 # Save
-saveRDS(em_trends_age, file = here::here("products", "manuscript", "supplement", "emtrends_age_group.rds"))
-saveRDS(slope_contrast, file = here::here("products", "manuscript", "supplement", "slope_contrast_age_group.rds"))
+invisible(saveRDS(em_trends_age, file = here::here("results", "output", "statistical-analysis", "emtrends_age_group.rds")))
+invisible(saveRDS(slope_contrast, file = here::here("results", "output", "statistical-analysis", "slope_contrast_age_group.rds")))
 
@@
 # Save formatted contrast table
-saveRDS(slope_contrast_age_fmt, here::here("results", "tables", "slope_contrast_agegroup_fmt.rds"))
+invisible(saveRDS(slope_contrast_age_fmt, here::here("results", "tables", "slope_contrast_agegroup_fmt.rds")))
 
@@
 # Save figure to manuscript folder
-ggsave(
-  filename = here::here("results", "figures", "slope_plot_age_group.png"),
+invisible(ggsave(filename = here::here("results", "figures", "slope_plot_age_group.png"),
   plot = slope_plot_agegroup,
-  width = 9, height = 6, dpi = 300)
+  width = 9, height = 6, dpi = 300))
 
 # Save RDS for reproducibility
-saveRDS(slope_plot_agegroup, file = here::here("results", "figures", "slope_plot_age_group.rds"))
+invisible(saveRDS(slope_plot_agegroup, file = here::here("results", "output", "statistical-analysis", "slope_plot_age_group.rds")))
 
@@
 Estimated yearly changes in mortality rates were significantly higher for adults aged 65+ (slope = 1.26, 95% CI: 1.18–1.34) compared to those aged 35–64 (slope = 1.04, 95% CI: 0.96–1.12).
-The difference in slope between age groups was statistically significant (difference = -0.224, p < 0.001), indicating a steeper rise in mortality among older adults.
+The difference in slope between age groups was statistically significant (difference = -0.224, p \< 0.001), indicating a steeper rise in mortality among older adults.
 
@@
 Converting county data to spatial object using longitude/latitude.
-Creating static spatial maps of county-level mortality rates by race/ethnicity, visualized using spatial coordinates. These provide an overview of geographic variation across U.S. counties.
+Creating static spatial maps of county-level mortality rates by race/ethnicity, visualized using spatial coordinates.
+These provide an overview of geographic variation across U.S.
+counties.
 
@@
 print(spatial_plot_all)
-ggsave(here::here("products", "manuscript", "supplement", "spatial_plot_all.png"), plot = spatial_plot_all)
+invisible(ggsave(here::here("results", "figures", "spatial_plot_all.png"), plot = spatial_plot_all))
 
@@
 print(black_spatial_plot)
-ggsave(here::here("products", "manuscript", "supplement", "black_spatial_plot.png"), plot = black_spatial_plot)
+invisible(ggsave(here::here("results", "figures", "black_spatial_plot.png"), plot = black_spatial_plot))
 
@@
 print(white_spatial_plot)
-ggsave(here::here("products", "manuscript", "supplement", "white_spatial_plot.png"), plot = white_spatial_plot)
+invisible(ggsave(here::here("results", "figures", "white_spatial_plot.png"), plot = white_spatial_plot))
 
@@
 print(hispanic_spatial_plot)
-ggsave(here::here("products", "manuscript", "supplement", "hispanic_spatial_plot.png"), plot = hispanic_spatial_plot)
+invisible(ggsave(here::here("results", "figures", "hispanic_spatial_plot.png"), plot = hispanic_spatial_plot))
 
@@
 print(native_spatial_plot)
-ggsave(here::here("products", "manuscript", "supplement", "native_spatial_plot.png"), plot = native_spatial_plot)
+invisible(ggsave(here::here("results", "figures", "native_spatial_plot.png"), plot = native_spatial_plot))
 
@@
 print(asian_spatial_plot)
-ggsave(here::here("products", "manuscript", "supplement", "asian_spatial_plot.png"), plot = asian_spatial_plot)
+invisible(ggsave(here::here("results", "figures", "asian_spatial_plot.png"), plot = asian_spatial_plot))
+
+```
+
+```{r}
+# Convert to sf object
+county_race_sex_sf <- county_race_sex %>%
+  st_as_sf(coords = c("X_long", "Y_lat"), crs = 4326, remove = FALSE)
+
+# Create a reusable theme
+spatial_theme <- theme_minimal(base_size = 12) +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 13, face = "bold"),
+    axis.text = element_text(size = 10),
+    axis.title = element_text(size = 11),
+    legend.text = element_text(size = 10),
+    legend.title = element_text(size = 10)
+  )
+
+# 1. All races
+spatial_plot_all <- ggplot() +
+  geom_sf(data = county_race_sex_sf, aes(color = mortalities), size = 0.7) +
+  scale_color_viridis_c() +
+  labs(title = "Spatial Distribution of Mortality Rates – All Races", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
+  spatial_theme
+ggsave(here::here("results", "figures", "spatial_plot_all.png"), spatial_plot_all, width = 7, height = 5, dpi = 320)
+
+spatial_plot_all
+
+# 2. Black
+black_race_sf <- county_race_sex_sf %>% filter(race_ethnicity == "Black")
+black_spatial_plot <- ggplot() +
+  geom_sf(data = black_race_sf, aes(color = mortalities), size = 0.7) +
+  scale_color_viridis_c() +
+  labs(title = "Spatial Distribution of Mortality Rates – Black", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
+  spatial_theme
+ggsave(here::here("results", "figures", "black_spatial_plot.png"), black_spatial_plot, width = 7, height = 5, dpi = 320)
+
+black_spatial_plot
+
+# 3. White
+white_race_sf <- county_race_sex_sf %>% filter(race_ethnicity == "White")
+white_spatial_plot <- ggplot() +
+  geom_sf(data = white_race_sf, aes(color = mortalities), size = 0.7) +
+  scale_color_viridis_c() +
+  labs(title = "Spatial Distribution of Mortality Rates – White", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
+  spatial_theme
+ggsave(here::here("results", "figures", "white_spatial_plot.png"), white_spatial_plot, width = 7, height = 5, dpi = 320)
+
+white_spatial_plot
+
+# 4. Hispanic
+hispanic_sf <- county_race_sex_sf %>% filter(race_ethnicity == "Hispanic")
+hispanic_spatial_plot <- ggplot() +
+  geom_sf(data = hispanic_sf, aes(color = mortalities), size = 0.7) +
+  scale_color_viridis_c() +
+  labs(title = "Spatial Distribution of Mortality Rates – Hispanic", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
+  spatial_theme
+ggsave(here::here("results", "figures", "hispanic_spatial_plot.png"), hispanic_spatial_plot, width = 7, height = 5, dpi = 320)
+
+hispanic_spatial_plot
+
+# 5. American Indian and Alaska Native
+native_sf <- county_race_sex_sf %>% filter(race_ethnicity == "American Indian and Alaska Native")
+native_spatial_plot <- ggplot() +
+  geom_sf(data = native_sf, aes(color = mortalities), size = 0.7) +
+  scale_color_viridis_c() +
+  labs(title = "Spatial Distribution of Mortality Rates – American Indian and Alaska Native", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
+  spatial_theme
+ggsave(here::here("results", "figures", "native_spatial_plot.png"), native_spatial_plot, width = 7, height = 5, dpi = 320)
+
+native_spatial_plot
+
+# 6. Asian and Pacific Islander
+asian_sf <- county_race_sex_sf %>% filter(race_ethnicity == "Asian and Pacific Islander")
+asian_spatial_plot <- ggplot() +
+  geom_sf(data = asian_sf, aes(color = mortalities), size = 0.7) +
+  scale_color_viridis_c() +
+  labs(title = "Spatial Distribution of Mortality Rates – Asian and Pacific Islander", x = "Longitude", y = "Latitude", color = "Mortality Rate") +
+  spatial_theme
+ggsave(here::here("results", "figures", "asian_spatial_plot.png"), asian_spatial_plot, width = 10, height = 5, dpi = 320)
+
+asian_spatial_plot
 
@@
 ```{r}
-
-# Combine all of the spatial charts using patchwork
-combined_spatial_plot <- (spatial_plot_all | black_spatial_plot | white_spatial_plot) /
-  (hispanic_spatial_plot | native_spatial_plot |
-  asian_spatial_plot)
-
-# Print and save the combined spatial plot
+# Explicit spacing and margins
+combined_spatial_plot <- (
+  spatial_plot_all + theme(plot.margin = margin(10, 10, 10, 10)) |
+  black_spatial_plot + theme(plot.margin = margin(10, 10, 10, 10))
+) /
+(
+  white_spatial_plot + theme(plot.margin = margin(10, 10, 10, 10)) |
+  hispanic_spatial_plot + theme(plot.margin = margin(10, 10, 10, 10))
+) /
+(
+  native_spatial_plot + theme(plot.margin = margin(10, 10, 10, 10)) |
+  asian_spatial_plot + theme(plot.margin = margin(10, 10, 10, 10))
+)
+
+ggsave(
+  here::here("results", "figures", "combined_spatial_plot.png"),
+  plot = combined_spatial_plot,
+  width = 21,   # large print-friendly layout
+  height = 15,
+  dpi = 320
+)
+
 print(combined_spatial_plot)
-ggsave(here::here("products", "manuscript", "supplement", "combined_spatial_plot.png"), 
-       plot = combined_spatial_plot, width = 14, height = 10)
-
-```
-
-These plots are best viewed as images outside of R due to size.
+```
+
+These plots are best viewed as images outside of R due to size limits in R.
 
@@
 ### Evaluating spatial clustering of Mortality Rates.
+
 ### 2.3: Global Moran’s I – All Races Combined
@@
 # Save to results/tables
-saveRDS(moran_table, here::here("results", "tables", "global_morans_i_summary.rds"))
-saveRDS(moran_result, file = moran_path)
-
-```
-
-The Global Moran’s I statistic (I = 0.0435, p < 0.001) indicates weak but statistically significant positive spatial autocorrelation in hypertension-related CVD mortality at the county level.
+invisible(saveRDS(moran_table, here::here("results", "tables", "global_morans_i_summary.rds")))
+invisible(saveRDS(moran_result, file = moran_path))
+
+```
+
+The Global Moran’s I statistic (I = 0.0435, p \< 0.001) indicates weak but statistically significant positive spatial autocorrelation in hypertension-related CVD mortality at the county level.
 There is weak-to-moderate positive spatial autocorrelation in mortality rates across U.S.
@@
 # Save plot
-ggsave(filename = here::here("products", "manuscript", "supplement", "lisa_significance_map_centroids.png"),
+invisible(ggsave(filename = here::here("results", "figures", "lisa_significance_map_centroids.png"),
   plot = lisa_point_plot,
   width = 10,
-  height = 6)
+  height = 6))
 
@@
 
-ggsave(
-  here::here("products", "manuscript", "supplement", "lisa_cluster_type_map.png"),
+invisible(ggsave(here::here("results", "figures", "lisa_cluster_type_map.png"),
   plot = lisa_cluster_plot,
-  width = 10, height = 6)
+  width = 10, height = 6))
 
@@
 #### Focusing at a State level Due to the size of the dataset it is challenging to work with the full dataset.
+
 For convenience and because the SouthEast is historically known to have a high prevalence of cardiovasccular mortalities we will focus on Georgia for this next step.
@@
 
-ggsave(
-  here::here("products", "manuscript", "supplement", "ga_lisa_cluster_plot.png"),
+invisible(ggsave(here::here("results", "figures", "ga_lisa_cluster_plot.png"),
   plot = ga_lisa_plot,
   width = 8,
-  height = 6)
+  height = 6))
 ```
@@
 
-ggsave(filename = here::here("products", "manuscript", "supplement", "ms_lisa_cluster_plot.png"),
+invisible(ggsave(filename = here::here("results", "figures", "ms_lisa_cluster_plot.png"),
   plot = ms_lisa_plot,
   width = 8,
-  height = 6)
+  height = 6))
 
@@
 # Save plot
-ggsave(
-  filename = here::here("products", "manuscript", "supplement", "tx_lisa_cluster_plot.png"),
+invisible(ggsave(filename = here::here("results", "figures", "tx_lisa_cluster_plot.png"),
   plot = tx_lisa_plot,
-  width = 9, height = 6)
+  width = 9, height = 6))
 ```
@@
 ```{r}
-
 # Paths for saving spatial weights
@@
 # Save
-ggsave(
-  here::here("products", "manuscript", "supplement", "lisa_cluster_kentucky.png"),
-  plot = lisa_ky_plot, width = 9, height = 6)
+invisible(ggsave(here::here("results", "figures", "lisa_cluster_kentucky.png"),
+  plot = lisa_ky_plot, width = 9, height = 6))
 
@@
 # Save the result
-saveRDS(ky_moran_result, here::here("results", "output", "ky_global_morans_i.rds"))
+invisible(saveRDS(ky_moran_result, here::here("results", "output", "ky_global_morans_i.rds")))
 
@@
 counties.
-Nationally, over 40% of counties showed statistically significant clustering at p < 0.05.
+Nationally, over 40% of counties showed statistically significant clustering at p \< 0.05.
 However, most of these represented spatial outliers—counties with high (or low) mortality rates surrounded by contrasting neighbors—rather than concentrated hotspots.
@@
 # Save table
-saveRDS(moran_gt_tbl, here::here("results", "tables", "moran_global_summary.rds"))
+invisible(saveRDS(moran_gt_tbl, here::here("results", "tables", "moran_global_summary.rds")))
 
@@
 Global Moran’s I statistics were statistically significant across all regions analyzed.
-The USA showed weak but significant positive spatial autocorrelation (I = 0.0435, p < 0.001).
-Georgia, Mississippi, Texas, and Kentucky also had low Moran’s I values (0.0168–0.0297), but all were statistically significant (p < 0.001), confirming that mortality rates were not randomly distributed in space.
+The USA showed weak but significant positive spatial autocorrelation (I = 0.0435, p \< 0.001).
+Georgia, Mississippi, Texas, and Kentucky also had low Moran’s I values (0.0168–0.0297), but all were statistically significant (p \< 0.001), confirming that mortality rates were not randomly distributed in space.
 These findings support further investigation of local spatial clustering using LISA.
@@
 # Save for results
-saveRDS(model_fit_race_sex, here::here("results", "tables", "model_fit_race_sex.rds"))
+invisible(saveRDS(model_fit_race_sex, here::here("results", "tables", "model_fit_race_sex.rds")))
 
@@
 # Save nicely formatted table
-saveRDS(slope_table_race_sex_fmt, here::here("results", "tables", "slope_table_race_sex_fmt.rds"))
+invisible(saveRDS(slope_table_race_sex_fmt, here::here("results", "tables", "slope_table_race_sex_fmt.rds")))
 
@@
 
+# Create high-quality PNG table for manuscript
+slope_table_race_sex_gt <- slope_table_race_sex_fmt %>%
+  gt() %>%
+  tab_header(
+    title = "Estimated Annual Mortality Slopes by Race/Ethnicity and Sex",
+    subtitle = "From GLM with Year × Race × Sex Interaction"
+  ) %>%
+  fmt_number(columns = c(Estimate, SE), decimals = 3) %>%
+  cols_label(
+    Race = "Race/Ethnicity",
+    Sex = "Sex",
+    Estimate = "Slope",
+    SE = "Standard Error",
+    CI = "95% CI"
+  ) %>%
+  opt_table_font(font = google_font("Roboto")) %>%
+  tab_options(table.font.size = 10)
+
+# Save as image for manuscript inclusion
+gtsave(
+  slope_table_race_sex_gt,
+  filename = here::here("results", "tables", "slope_table_race_sex.png")
+)
+
+slope_table_race_sex_gt
+
+```
+
+```{r}
 # Compare slopes between race/sex groups
@@
 # Save slope estimates and pairwise contrasts
-saveRDS(em_trends_race_sex, file = here::here("results", "output", "emtrends_race_sex.rds"))
-saveRDS(slope_contrast_race_sex, file = here::here("results", "output", "slope_contrast_race_sex.rds"))
+invisible(saveRDS(em_trends_race_sex, file = here::here("results", "output", "emtrends_race_sex.rds")))
+invisible(saveRDS(slope_contrast_race_sex, file = here::here("results", "output", "slope_contrast_race_sex.rds")))
 
@@
 # Save and display the plot
-ggsave(here::here("results", "output", "slope_plot_race_sex.png"),
-       plot = slope_plot_race_sex, width = 10, height = 6)
+invisible(ggsave(here::here("results", "figures", "slope_plot_race_sex.png"),
+       plot = slope_plot_race_sex, width = 10, height = 6))
 print(slope_plot_race_sex)
@@
 # Save slopes
-saveRDS(em_trends_race_sex_age, here::here("results", "output", "emtrends_race_sex_by_age.rds"))
+invisible(saveRDS(em_trends_race_sex_age, here::here("results", "output", "emtrends_race_sex_by_age.rds")))
 
@@
 # Save
-saveRDS(slope_contrast_race_sex_age, here::here("results", "output", "slope_contrast_race_sex_by_age.rds"))
+invisible(saveRDS(slope_contrast_race_sex_age, here::here("results", "output", "slope_contrast_race_sex_by_age.rds")))
 
@@
 # Save output
-saveRDS(slope_contrast_race_sex_age_sorted,
-        here::here("results", "output", "slope_contrast_race_sex_by_age_sorted.rds"))
+invisible(saveRDS(slope_contrast_race_sex_age_sorted, here::here("results", "output", "slope_contrast_race_sex_by_age_sorted.rds")))
 
@@
 # Save and display
-ggsave(here::here("results", "output", "pairwise_slope_contrasts_filtered.png"),
-  plot = contrast_plot, width = 11, height = 8)
+invisible(ggsave(here::here("results", "figures", "pairwise_slope_contrasts_filtered.png"),
+  plot = contrast_plot, width = 11, height = 8))
 print(contrast_plot)
@@
 ```{r}
-# Convert estimated emtrends object to a data frame for plotting
-# Create a group label that includes race, sex, and age group
-slope_df <- as.data.frame(em_trends_race_sex_age) %>%
-  mutate(group = paste(race_ethnicity, sex),
-         group = paste(group, age_group, sep = "___"))
-
 # Helper function to clean group labels (removes age group text)
@@
       group = str_remove(group, "___Ages 65\\+ years"),
-      group = str_replace(group, "Asian and Pacific Islander", "Asian & Pacific\nIslander"),
-      group = str_replace(group, "American Indian and Alaska Native", "American Indian\n& Alaska Native"),
-      group = str_replace(group, "Overall Female", "Female"),
-      group = str_replace(group, "Overall Male", "Male"),
-      group = str_replace(group, "Overall Overall", "All Groups"),
-      group = str_replace(group, "White Overall", "White"),
-      group = str_replace(group, "Black Overall", "Black"),
-      group = str_replace(group, "Hispanic Overall", "Hispanic"),
-      group = str_replace(group, "Asian & Pacific\nIslander Overall", "Asian & Pacific\nIslander"),
-      group = str_replace(group, "American Indian\n& Alaska Native Overall", "American Indian\n& Alaska Native"),
-      group = str_trim(group))}
-
+      group = str_replace_all(group, "Asian and Pacific Islander", "Asian & Pacific Islander"),
+      group = str_replace_all(group, "American Indian and Alaska Native", "American Indian & Alaska Native"),
+      group = case_when(
+        str_detect(group, "^Overall Female") ~ "Overall Female (All Races)",
+        str_detect(group, "^Overall Male") ~ "Overall Male (All Races)",
+        str_detect(group, "^Overall Overall") ~ "Overall (All Races and Sexes)",
+        TRUE ~ group
+      ),
+      group = str_squish(group),
+      group = str_wrap(group, width = 18))}
+
+# Convert estimated emtrends object to a data frame for plotting
+# Create a group label that includes race, sex, and age group
 # Apply and split
+slope_df <- as.data.frame(em_trends_race_sex_age) %>%
+  mutate(group = paste(race_ethnicity, sex),
+         group = paste(group, age_group, sep = "___"))
+
 slope_df_35_64 <- slope_df %>% filter(age_group == "Ages 35-64 years") %>% clean_labels()
@@
 
-# Common ggplot function
-make_slope_plot <- function(data, title, fill_color) {
-  data <- data %>% 
-    filter(!is.na(group), !is.na(Year.trend)) %>%   # remove rows with missing group or slope
-    mutate(group = as.factor(group))                # make sure group is a factor
-
-  ggplot(data, aes(x = Year.trend, y = fct_reorder(group, Year.trend))) +
-    geom_col(fill = fill_color) +
-    geom_errorbarh(aes(xmin = Year.trend - 1.96 * SE,
-                       xmax = Year.trend + 1.96 * SE),
-                   height = 0.3, color = "black", alpha = 0.7) +
-    labs(title = title,
-      x = "Estimated Annual Change in Mortality Rate",
-      y = "Race and Ethnicity") +
-    theme_minimal(base_size = 13) +
+# Final slope plot function
+make_slope_plot <- function(data, title_text, fill_color) {
+  data %>%
+    filter(!is.na(group), !is.na(Year.trend)) %>%
+    mutate(group = fct_reorder(group, Year.trend)) %>%
+    ggplot(aes(x = Year.trend, y = group)) +
+    geom_col(fill = fill_color, width = 0.7, alpha = 0.9) +
+    geom_errorbarh(
+      aes(xmin = Year.trend - 1.96 * SE, xmax = Year.trend + 1.96 * SE),
+      height = 0.3, color = "black", alpha = 0.7
+    ) +
+    labs(
+      title = title_text,
+      subtitle = "Estimated slope with 95% CI from GLM (Year × Race × Sex)",
+      x = "Estimated Annual Change in Mortality Rate (per 100,000)",
+      y = "Race/Sex Group"
+    ) +
+    theme_minimal(base_size = 14) +
     theme(
-      plot.title = element_text(hjust = 0.5),
-      axis.text.y = element_text(size = 11),
-      axis.text.x = element_text(size = 11))}
-
-# Generate plots
-slope_plot_35_64 <- make_slope_plot(slope_df_35_64, "Estimated Annual Mortality Change (Ages 35–64)", "#F8766D")
-slope_plot_65_plus <- make_slope_plot(slope_df_65_plus, "Estimated Annual Mortality Change (Ages 65+)", "#00BFC4")
-
-# Save plots
-ggsave(here("results", "figures", "slope_plot_35_64.png"),
-       slope_plot_35_64, width = 11, height = 7, dpi = 300)
-
-ggsave(here("results", "figures", "slope_plot_65_plus.png"),
-       slope_plot_65_plus, width = 11, height = 7, dpi = 300)
-
-# Save RDS objects (optional)
-saveRDS(slope_plot_35_64, here("results", "figures", "slope_plot_35_64.rds"))
-saveRDS(slope_plot_65_plus, here("results", "figures", "slope_plot_65_plus.rds"))
+      plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
+      plot.subtitle = element_text(size = 16, hjust = 0.5),
+      axis.title = element_text(size = 15),
+      axis.text.y = element_text(size = 15),
+      axis.text.x = element_text(size = 15)
+    )
+}
+
+# Create final plots
+slope_plot_35_64 <- make_slope_plot(slope_df_35_64, "Estimated Annual Mortality Change (Ages 35–64)", "#4C78A8")
+slope_plot_65_plus <- make_slope_plot(slope_df_65_plus, "Estimated Annual Mortality Change (Ages 65+)", "#A2C7E5")
+
+# Save both
+ggsave(here("results", "figures", "slope_plot_35_64.png"), slope_plot_35_64, width = 10.5, height = 7, dpi = 300)
+ggsave(here("results", "figures", "slope_plot_65_plus.png"), slope_plot_65_plus, width = 10.5, height = 7, dpi = 300)
+
+#Save slimmed RDS for manuscript or reuse
+saveRDS(slope_plot_35_64, here("results", "output", "statistical-analysis", "slope_plot_35_64.rds"))
+saveRDS(slope_plot_65_plus, here("results", "output", "statistical-analysis", "slope_plot_65_plus.rds"))
 
@@
 slope_plot_65_plus
+
 ```
@@
 # Save for manuscript
-saveRDS(mortality_summary_table, here::here("results", "tables", "mortality_summary_table.rds"))
+invisible(saveRDS(mortality_summary_table, here::here("results", "tables", "mortality_summary_table.rds")))
 
@@
 
-To optimize the number of predictors (`mtry`) and minimum node size (`min_n`), we perform 10-fold cross-validation over a regular grid of parameter values.
-# Define path to save tuning results
+To optimize the number of predictors (`mtry`) and minimum node size (`min_n`), we perform 10-fold cross-validation over a regular grid of parameter values. \# Define path to save tuning results
 
@@
 Performance differences between parameter combinations were minor when mtry = 4, but models with fewer predictors showed substantially higher RMSE values.
+
+```{r}
+# Build summary with 95% CI
+rf_cv_summary <- rf_metrics %>%
+  filter(.metric %in% c("rmse", "rsq")) %>%
+  mutate(
+    lower_95 = mean - 1.96 * std_err,
+    upper_95 = mean + 1.96 * std_err,
+    ci_range = paste0(sprintf("%.2f", mean), " [", 
+                      sprintf("%.2f", lower_95), ", ", 
+                      sprintf("%.2f", upper_95), "]")
+  ) %>%
+  select(.metric, ci_range, mtry, min_n) %>%
+  arrange(.metric, mtry)
+
+# Get row with best RMSE
+rf_cv_summary_table <- rf_cv_summary %>%
+  group_by(.metric) %>%
+  slice_min(mtry, n = 1) %>%
+  ungroup() %>%
+  tidyr::pivot_wider(names_from = .metric, values_from = ci_range) %>%
+  rename(
+    `mtry` = mtry,
+    `min_n` = min_n,
+    `RMSE (95% CI)` = rmse,
+    `R² (95% CI)` = rsq
+  )
+
+# Create GT table
+rf_cv_gt <- rf_cv_summary_table %>%
+  gt() %>%
+  tab_header(
+    title = md("**Random Forest CV Performance Summary**"),
+    subtitle = "Mean performance across 10-fold CV with 95% confidence intervals"
+  ) %>%
+  tab_options(
+    table.font.size = "small",
+    table.width = pct(70),
+    column_labels.font.weight = "bold",
+    data_row.padding = px(3)
+  )
+
+# Save
+saveRDS(rf_cv_gt, here("results", "output", "statistical-analysis", "rf_cv_performance_table.rds"))
+gtsave(rf_cv_gt, here("results", "tables", "rf_cv_performance_table.png"), expand = 3)
+
+rf_cv_gt
+
+```
 
@@
 # Save the summary table
-saveRDS(rf_summary_table, here::here("results", "tables", "model_performance_summary.rds"))
+invisible(saveRDS(rf_summary_table, here::here("results", "tables", "rf_performance_summary.rds")))
 
@@
 ```{r}
-## Random Forest: Predicted vs Observed Plot
-
-# Load predictions (if not already in memory)
-rf_predictions <- readRDS(here::here("results", "output", "rf_test_predictions.rds"))
-
-# Create the predicted vs observed plot
+# Load predictions
+rf_predictions <- readRDS(here("results", "output", "rf_test_predictions.rds"))
+
+# Calculate R² on test set (optional overlay)
+r_squared <- cor(rf_predictions$mortalities, rf_predictions$.pred)^2
+
+# Generate the plot
 rf_pred_plot <- ggplot(rf_predictions, aes(x = mortalities, y = .pred)) +
-  geom_point(alpha = 0.4, color = "darkgreen") +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
+  geom_point(alpha = 0.5, color = "#4C78A8", size = 1.2) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
   labs(
-    title = "Random Forest: Predicted vs Observed Mortality",
-    x = "Observed Mortality",
-    y = "Predicted Mortality") +
-  theme_minimal()
-
-# Save the plot to results/figures
+    title = "Random Forest: Predicted vs. Observed Mortality",
+    subtitle = paste0("Test set R² = ", round(r_squared, 3)),
+    x = "Observed Mortality Rate (per 100,000)",
+    y = "Predicted Mortality Rate (per 100,000)"
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    plot.title = element_text(face = "bold", hjust = 0.5),
+    plot.subtitle = element_text(hjust = 0.5)
+  )
+
+# Save the figure
 ggsave(
-  filename = here::here("results", "figures", "rf_pred_vs_obs.png"),
-  plot = rf_pred_plot,
-  width = 7, height = 5, dpi = 300)
+  here("results", "figures", "rf_pred_vs_obs.png"),
+  rf_pred_plot, width = 8, height = 6, dpi = 300)
 
 rf_pred_plot
+
 ```
@@
 ```{r}
-## Random Forest: Variable Importance Plot
-
-# Load fitted model (if not already in memory)
-final_rf_fit <- readRDS(here::here("results", "output", "final_rf_fit.rds"))
-
-# Create VIP plot
-rf_vip_plot <- final_rf_fit %>%
+# Load the final fitted model
+final_rf_fit <- readRDS(here("results", "output", "final_rf_fit.rds"))
+
+# Extract and clean variable importance
+rf_vip_data <- final_rf_fit %>%
   extract_fit_parsnip() %>%
-  vip::vip(num_features = 10) +
-  labs(title = "Top 10 Predictors (Random Forest)") +
-  theme_minimal()
-
-# Save the plot to results/figures
+  vi() %>%
+  arrange(desc(Importance)) %>%
+  slice(1:10) %>%
+  mutate(
+    CleanLabel = Variable %>%
+      str_replace_all("_", " ") %>%
+      str_replace_all("\\.", " ") %>%
+      str_replace("age group Ages 65  years", "Age Group: 65+ years") %>%
+      str_replace("sex Male", "Sex: Male") %>%
+      str_replace("sex Female", "Sex: Female") %>%
+      str_replace("sex Overall", "Sex: Overall") %>%
+      str_replace("race ethnicity Black", "Race: Black") %>%
+      str_replace("race ethnicity White", "Race: White") %>%
+      str_replace("race ethnicity Hispanic", "Race: Hispanic") %>%
+      str_replace("race ethnicity Asian and Pacific Islander", "Race: Asian & Pacific Islander") %>%
+      str_replace("race ethnicity American Indian and Alaska Native", "Race: American Indian & Alaska Native") %>%
+      str_replace("race ethnicity Overall", "Race: Overall") %>%
+      str_squish()
+  )
+
+# Create the VIP plot
+rf_vip_plot <- ggplot(rf_vip_data, aes(x = Importance, y = fct_reorder(CleanLabel, Importance))) +
+  geom_col(fill = "#4C78A8", width = 0.75) +
+  labs(
+    title = "Random Forest: Variable Importance",
+    subtitle = "Top 10 predictors ranked by permutation importance",
+    x = "Importance Score",
+    y = NULL
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
+    plot.subtitle = element_text(size = 12, hjust = 0.5),
+    axis.text.y = element_text(size = 12)
+  )
+
+# Save the plot
 ggsave(
-  filename = here::here("results", "figures", "rf_variable_importance.png"),
+  filename = here("results", "figures", "rf_variable_importance.png"),
   plot = rf_vip_plot,
-  width = 7)
+  width = 8,
+  height = 7,
+  dpi = 300
+)
 
@@
 ```
+
+```{r}
+# Define output path
+lasso_cv_rds_path <- here("results", "tables", "lasso_cv_performance_table.rds")
+
+if (file.exists(lasso_cv_rds_path)) {
+  message("LASSO CV performance table already exists. Skipping evaluation...")
+  lasso_cv_gt_ci <- readRDS(lasso_cv_rds_path)
+
+} else {
+  message("Generating LASSO CV performance table from large file...")
+
+  # Load full tuned object
+  lasso_tuned <- readRDS(here("results", "large-files", "lasso_tuned.rds"))
+
+  # 1. Collect fold-level predictions and rename for yardstick
+  lasso_cv_pred <- collect_predictions(lasso_tuned) %>%
+    rename(truth = mortalities)
+
+  # 2. Identify best penalty based on average RMSE across folds
+  lasso_cv_best_penalty <- lasso_cv_pred %>%
+    group_by(penalty) %>%
+    yardstick::rmse(truth = truth, estimate = .pred) %>%
+    slice_min(.estimate) %>%
+    pull(penalty)
+  lasso_cv_best_penalty <- lasso_cv_best_penalty[1]  # ensure scalar
+
+  # 3. Filter only the rows with the selected penalty
+  lasso_cv_filtered <- lasso_cv_pred %>%
+    filter(penalty == lasso_cv_best_penalty)
+
+  # 4. Loop through folds to compute RMSE and R²
+  fold_ids <- unique(lasso_cv_filtered$id)
+
+  lasso_cv_fold_metrics <- lapply(fold_ids, function(fold) {
+    fold_data <- lasso_cv_filtered %>% filter(id == fold)
+
+    tibble(
+      id = fold,
+      rmse = yardstick::rmse(fold_data, truth = truth, estimate = .pred)[[".estimate"]],
+      rsq  = yardstick::rsq(fold_data, truth = truth, estimate = .pred)[[".estimate"]]
+    )
+  }) %>%
+    bind_rows()
+
+  # 5. Create 95% CI summary table
+  lasso_cv_ci_table <- tibble(
+    `Selected Penalty` = lasso_cv_best_penalty,
+    `RMSE (95% CI)` = sprintf("%.2f [%.2f, %.2f]",
+                              mean(lasso_cv_fold_metrics$rmse),
+                              mean(lasso_cv_fold_metrics$rmse) - 1.96 * sd(lasso_cv_fold_metrics$rmse),
+                              mean(lasso_cv_fold_metrics$rmse) + 1.96 * sd(lasso_cv_fold_metrics$rmse)),
+    `R² (95% CI)` = sprintf("%.2f [%.2f, %.2f]",
+                            mean(lasso_cv_fold_metrics$rsq),
+                            mean(lasso_cv_fold_metrics$rsq) - 1.96 * sd(lasso_cv_fold_metrics$rsq),
+                            mean(lasso_cv_fold_metrics$rsq) + 1.96 * sd(lasso_cv_fold_metrics$rsq))
+  )
+
+  # 6. Format with gt
+  lasso_cv_gt_ci <- lasso_cv_ci_table %>%
+    gt() %>%
+    tab_header(
+      title = md("**LASSO CV Performance Summary**"),
+      subtitle = "Mean RMSE and R² with 95% CI across CV folds"
+    ) %>%
+    tab_options(
+      table.font.size = "small",
+      table.width = pct(70),
+      column_labels.font.weight = "bold",
+      data_row.padding = px(3)
+    )
+
+  # 7. Save outputs
+  saveRDS(lasso_cv_gt_ci, lasso_cv_rds_path)
+  gtsave(lasso_cv_gt_ci,
+         filename = here("results", "tables", "lasso_cv_performance_table.png"),
+         expand = 3)
+}
+
+# View (always)
+lasso_cv_gt_ci
+
+```
+
+Cross-validation results for the LASSO regression model demonstrated consistent performance across folds.
+The model achieved a mean RMSE of 119.63 (95% CI: 117.24, 122.02) and an R² of 0.73 (95% CI: 0.72, 0.73), indicating high stability in predictive accuracy and explained variance.
+These narrow confidence intervals suggest minimal variability across folds, further supporting the robustness and generalizability of the LASSO model when applied to county-level mortality data.
 
@@
 # Save to results/figures
-ggsave(filename = here::here("results", "figures", "lasso_tuning_plot.png"),
+invisible(ggsave(filename = here::here("results", "figures", "lasso_tuning_plot.png"),
   plot = lasso_tune_plot,
-  width = 7, height = 5, dpi = 300)
+  width = 7, height = 5, dpi = 300))
 
@@
 
-LASSO tuning selected a penalty value of approximately 0.0001, minimizing RMSE at ~119.63.
+LASSO tuning selected a penalty value of approximately 0.0001, minimizing RMSE at \~119.63.
 Coefficient shrinkage was applied, resulting in the exclusion of several predictors and aiding model interpretability.
@@
 The final LASSO model is evaluated on the test set and compared to the Random Forest model in terms of predictive accuracy and fit.
+
 ```{r}
@@
 ## LASSO: Predicted vs Observed Plot
-
-# Load predictions (if not already in memory)
-lasso_test_predictions <- readRDS(here::here("results", "output", "lasso_test_predictions.rds"))
-
-lasso_pred_plot <- ggplot(lasso_test_predictions, aes(x = mortalities, y = .pred)) +
-  geom_point(alpha = 0.4, color = "steelblue") +
-  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
+# Load test predictions
+lasso_predictions <- readRDS(here("results", "output", "lasso_test_predictions.rds"))
+
+# Compute R² on test set
+lasso_r2 <- cor(lasso_predictions$mortalities, lasso_predictions$.pred)^2
+
+# Create plot
+lasso_pred_plot <- ggplot(lasso_predictions, aes(x = mortalities, y = .pred)) +
+  geom_point(alpha = 0.4, color = "#4C78A8", size = 1.2) +
+  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray40") +
   labs(
-    title = "LASSO Regression: Predicted vs Observed Mortality",
-    x = "Observed Mortality",
-    y = "Predicted Mortality") +
-  theme_minimal()
-
-# Save to file
+    title = "LASSO Regression: Predicted vs. Observed Mortality",
+    subtitle = paste0("Test set R² = ", round(lasso_r2, 3)),
+    x = "Observed Mortality Rate (per 100,000)",
+    y = "Predicted Mortality Rate (per 100,000)"
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    plot.title = element_text(face = "bold", hjust = 0.5),
+    plot.subtitle = element_text(hjust = 0.5)
+  )
+
+# Save as PNG
 ggsave(
-  filename = here::here("results", "figures", "lasso_pred_vs_obs.png"),
-  plot = lasso_pred_plot,
-  width = 7, height = 5, dpi = 300)
+  here("results", "figures", "lasso_pred_vs_obs.png"),
+  lasso_pred_plot,
+  width = 8, height = 6, dpi = 300
+)
 
@@
 
-saveRDS(lasso_summary_table, here::here("results", "tables", "lasso_model_summary.rds"))
+invisible(saveRDS(lasso_summary_table, here::here("results", "tables", "lasso_model_summary.rds")))
 
@@
 
-LASSO regression achieved slightly lower performance than random forest, with an RMSE of ~119.6 and R² of ~0.726indicating a reasonably strong fit for predicting county-level hypertension-related mortality.
+LASSO regression achieved slightly lower performance than random forest, with an RMSE of \~119.6 and R² of \~0.726indicating a reasonably strong fit for predicting county-level hypertension-related mortality.
 Compared to the random forest model, LASSO performs slightly worse, but still retains good explanatory power.
@@
 
-### 4.4 Model Comparison
-
-To directly compare model performance, we generated side-by-side summaries of RMSE, R², and MAE for both random forest and LASSO regression using the test dataset.
-Confidence intervals are provided where available.
-
-Generating a model comparison table:
-
-```{r}
-
-## Generate model comparison table with 95% confidence intervals
-
-# Load test metrics
-rf_test_metrics <- readRDS(here::here("results", "output", "rf_test_metrics.rds"))
-lasso_metrics <- readRDS(here::here("results", "output", "lasso_test_metrics.rds"))
-
-# Add model names
-rf_test_metrics$model <- "Random Forest"
-lasso_metrics$model <- "LASSO Regression"
+### 4.4 Train vs. Test Set Performance
+
+```{r}
+# Predict on train data — Random Forest
+rf_train_predictions <- predict(final_rf_fit, new_data = train_data) %>%
+  bind_cols(train_data %>% select(mortalities))
+
+# Predict on test data — already exists
+rf_test_predictions <- readRDS(here("results", "output", "rf_test_predictions.rds"))
+
+# Evaluate
+rf_train_metrics <- rf_train_predictions %>%
+  metrics(truth = mortalities, estimate = .pred) %>%
+  mutate(Model = "Random Forest", Dataset = "Train")
+
+rf_test_metrics <- rf_test_predictions %>%
+  metrics(truth = mortalities, estimate = .pred) %>%
+  mutate(Model = "Random Forest", Dataset = "Test")
+
+# Predict on train data — LASSO
+lasso_train_predictions <- predict(final_lasso_fit, new_data = train_data) %>%
+  bind_cols(train_data %>% select(mortalities))
+
+# Predict on test data — already exists
+lasso_test_predictions <- readRDS(here("results", "output", "lasso_test_predictions.rds"))
+
+# Evaluate
+lasso_train_metrics <- lasso_train_predictions %>%
+  metrics(truth = mortalities, estimate = .pred) %>%
+  mutate(Model = "LASSO Regression", Dataset = "Train")
+
+lasso_test_metrics <- lasso_test_predictions %>%
+  metrics(truth = mortalities, estimate = .pred) %>%
+  mutate(Model = "LASSO Regression", Dataset = "Test")
 
 # Combine
-all_metrics <- bind_rows(rf_test_metrics, lasso_metrics)
-
-# Check if std_err is available
-has_se <- "std_err" %in% names(all_metrics)
-
-# Define function to create CI string
-format_ci <- function(mean, se = NA) {
-  if (is.na(se) || se == 0) return(as.character(round(mean, 2)))
-  lower <- round(mean - 1.96 * se, 2)
-  upper <- round(mean + 1.96 * se, 2)
-  paste0(round(mean, 2), " [", lower, ", ", upper, "]")}
-
-# Create final table
-if (has_se) {
-  # Use CI formatting
-  all_metrics_ci <- all_metrics %>%
-    select(model, .metric, mean = .estimate, std_err) %>%
-    mutate(ci = mapply(format_ci, mean, std_err)) %>%
-    select(model, .metric, ci) %>%
-    pivot_wider(names_from = .metric, values_from = ci)
-} else {
-  # Fall back to just point estimates
-  all_metrics_ci <- all_metrics %>%
-    select(model, .metric, mean = .estimate) %>%
-    mutate(ci = round(mean, 2)) %>%
-    select(model, .metric, ci) %>%
-    pivot_wider(names_from = .metric, values_from = ci)}
+train_test_metrics_combined <- bind_rows(
+  rf_train_metrics, rf_test_metrics,
+  lasso_train_metrics, lasso_test_metrics)
+
+# Format wide
+train_test_wide <- train_test_metrics_combined %>%
+  select(Model, Dataset, .metric, .estimate) %>%
+  tidyr::pivot_wider(names_from = .metric, values_from = .estimate) %>%
+  mutate(
+    RMSE = sprintf("%.2f", rmse),
+    MAE = sprintf("%.2f", mae),
+    R2  = sprintf("%.2f", rsq)
+  ) %>%
+  select(Model, Dataset, RMSE, MAE, R2)
 
 # Build gt table
-model_perf_ci_table <- all_metrics_ci %>%
+train_test_performance_table <- train_test_wide %>%
   gt() %>%
   tab_header(
-    title = "Model Performance Summary",
-    subtitle = "Test Set Evaluation for Random Forest and LASSO") %>%
+    title = md("**Model Performance: Train vs. Test Set**"),
+    subtitle = "Evaluation metrics for Random Forest and LASSO"
+  ) %>%
   cols_label(
-    model = "Model",
-    rmse = "RMSE",
-    rsq = "R²",
-    mae = "MAE") %>%
-  fmt_markdown(columns = everything()) %>%
+    RMSE = "RMSE",
+    MAE = "MAE",
+    R2 = "R²"
+  ) %>%
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_body(columns = "Model")
+  ) %>%
   tab_options(
-    table.font.size = px(12),
-    heading.title.font.size = px(14),
-    heading.subtitle.font.size = px(12),
-    column_labels.font.weight = "bold")
-
-# Save table
-saveRDS(model_perf_ci_table, here::here("results", "tables", "model_performance_summary.rds"))
-
-# View it
-model_perf_ci_table
-
-```
-
-Additional comparison table without CIs:
-
-```{r}
-model_perf_numeric <- bind_rows(rf_test_metrics, lasso_metrics) %>%
-  select(model, .metric, .estimate) %>%
-  pivot_wider(names_from = .metric, values_from = .estimate)
-
-saveRDS(model_perf_numeric, here::here("results", "output", "model_performance_numeric.rds"))
-
-model_perf_numeric
-```
+    table.font.size = "small",
+    table.width = pct(100),
+    column_labels.font.weight = "bold"
+  )
+
+# Save
+saveRDS(train_test_performance_table, here("results", "tables", "model_train_test_comparison.rds"))
+
+gtsave(
+  train_test_performance_table,
+  filename = here("results", "tables", "model_train_test_comparison.png"),
+  expand = 5)
+
+# View inline
+train_test_performance_table
+
+```
+
+These results indicate that both the Random Forest and LASSO models perform consistently on the training and test datasets, with near-identical RMSE, MAE, and R² values across splits.
+This suggests that the models are well-tuned and exhibit no signs of overfitting, supporting their generalizability to unseen data.
+
+### 4.5 Model Comparison (Including Null Model)
+
+#### 4.5.1 Null Model (Baseline Performance)
+
+```{r}
+# Load train and test data
+# train_data and test_data should contain the outcome: mortalities
+
+# 1. Compute the mean outcome in the training data
+mean_train_mortality <- mean(train_data$mortalities)
+
+# 2. Create a prediction column in test data
+null_predictions <- test_data %>%
+  mutate(.pred = mean_train_mortality)
+
+# 3. Calculate performance metrics
+null_test_metrics <- null_predictions %>%
+  yardstick::metrics(truth = mortalities, estimate = .pred)
+
+# 4. Reformat for consistency with other model outputs
+null_summary <- null_test_metrics %>%
+  select(.metric, .estimate) %>%
+  pivot_wider(names_from = .metric, values_from = .estimate) %>%
+  mutate(
+    R2 = ifelse(is.na(rsq), NA_real_, rsq),
+    Model = "Null Model",
+    Notes = "Predicts mean of training set"
+  ) %>%
+  select(Model, RMSE = rmse, R2 = rsq, MAE = mae, Notes)
+
+# Save to use in table if desired
+saveRDS(null_summary, here::here("results", "output", "null_model_summary.rds"))
+
+# preview
+null_summary
+
+```
+
+Note: The null model does not produce a valid R² because it predicts a constant value, resulting in undefined correlation.
+
+#### 4.5.2 LASSO and RF Model Comparison
+
+Comnparing model performance metrics between LASSO and RF. Generating a model comparison table:
+
+```{r}
+# Load saved test metrics
+rf_test_metrics <- readRDS(here("results", "output", "rf_test_metrics.rds"))
+lasso_test_metrics <- readRDS(here("results", "output", "lasso_test_metrics.rds"))
+
+# Summarize RF
+rf_summary <- rf_test_metrics %>%
+  select(.metric, .estimate) %>%
+  pivot_wider(names_from = .metric, values_from = .estimate) %>%
+  mutate(
+    Model = "Random Forest",
+    Notes = "Tuned (mtry = 4, min_n = 15)"
+  ) %>%
+  select(Model, RMSE = rmse, R2 = rsq, MAE = mae, Notes)
+
+# Summarize LASSO
+lasso_summary <- lasso_test_metrics %>%
+  select(.metric, .estimate) %>%
+  pivot_wider(names_from = .metric, values_from = .estimate) %>%
+  mutate(
+    Model = "LASSO Regression",
+    Notes = "Penalty = 0.1"
+  ) %>%
+  select(Model, RMSE = rmse, R2 = rsq, MAE = mae, Notes)
+
+# Combine summaries
+model_summary <- bind_rows(rf_summary, lasso_summary)
+
+# Create polished gt table
+model_performance_summary_table <- model_summary %>%
+  gt() %>%
+  tab_header(
+    title = md("**Model Performance Summary**"),
+    subtitle = "Held-out test set evaluation for Random Forest and LASSO"
+  ) %>%
+  fmt_number(columns = c(RMSE, R2, MAE), decimals = 2) %>%
+  cols_label(
+    RMSE = "RMSE",
+    R2 = "R²",
+    MAE = "MAE",
+    Notes = "Model Details"
+  ) %>%
+  tab_options(
+    table.font.size = "small",
+    data_row.padding = px(4),
+    table.width = pct(100))
+
+# Save as RDS
+saveRDS(model_performance_summary_table, here("results", "tables", "model_performance_summary.rds"))
+
+# Save as PNG
+gtsave(data = model_performance_summary_table,
+  filename = here("results", "tables", "model_performance_summary.png"),
+  expand = 5)
+
+# Also save numeric summary (wide format)
+saveRDS(model_summary, here("results", "output", "model_performance_numeric.rds"))
+
+model_performance_summary_table
+```
+
+To directly compare model performance, we generated side-by-side summaries of RMSE, R², and MAE for both random forest and LASSO regression using the test dataset.
 
@@
 
-Aim 4 Summary:
-We compared two predictive models for county-level hypertension-related CVD mortality: a random forest model and a LASSO regression model.
+#### 4.4.3 Comparing the LASSO and RF models to the Null model
+
+```{r}
+# Format RF
+rf_metrics_for_null <- rf_test_metrics %>%
+  select(.metric, .estimate) %>%
+  pivot_wider(names_from = .metric, values_from = .estimate) %>%
+  mutate(Model = "Random Forest")
+
+# Format LASSO
+lasso_metrics_for_null <- lasso_test_metrics %>%
+  select(.metric, .estimate) %>%
+  pivot_wider(names_from = .metric, values_from = .estimate) %>%
+  mutate(Model = "LASSO Regression")
+
+# Format Null
+null_metrics_for_null <- null_summary %>%
+  mutate(Model = "Null Model") %>%
+  select(Model, RMSE, R2, MAE)
+
+# Combine with MAE before R2
+model_comparison_null_data <- bind_rows(
+  rf_metrics_for_null %>% select(Model, RMSE = rmse, MAE = mae, R2 = rsq),
+  lasso_metrics_for_null %>% select(Model, RMSE = rmse, MAE = mae, R2 = rsq),
+  null_metrics_for_null %>% select(Model, RMSE, MAE, R2)
+)
+
+# Format as strings for clean spacing
+model_comparison_null_table <- model_comparison_null_data %>%
+  mutate(
+    RMSE = sprintf("%.2f", RMSE),
+    MAE  = sprintf("%.2f", MAE),
+    R2   = ifelse(is.na(R2), "–", sprintf("%.2f", R2))
+  ) %>%
+  gt() %>%
+  tab_header(
+    title = md("**Model Comparison Including Null Model**"),
+    subtitle = "Performance on held-out test set"
+  ) %>%
+  cols_label(
+    RMSE = "RMSE",
+    MAE  = "MAE",
+    R2   = "R²"
+  ) %>%
+  tab_options(
+    table.font.size = "small",
+    data_row.padding = px(4),
+    table.width = pct(100),
+    column_labels.font.weight = "bold"
+  )
+
+# Save
+saveRDS(model_comparison_null_table, here("results", "output", "statistical-analysis", "model_null_comparison_table.rds"))
+
+gtsave(model_comparison_null_table,
+       filename = here("results", "tables", "model_null_comparison_table.png"),
+       expand = 5)
+
+model_comparison_null_table
+```
+
+Reproducing the table for the manuscript:
+
+```{r}
+# 1. RF Summary for null-inclusive table
+null_comparison_rf_summary <- rf_test_metrics %>%
+  select(.metric, .estimate) %>%
+  tidyr::pivot_wider(names_from = .metric, values_from = .estimate) %>%
+  mutate(
+    Model = "Random Forest",
+    Notes = "Tuned (mtry = 4, min_n = 15)"
+  ) %>%
+  select(Model, RMSE = rmse, R2 = rsq, MAE = mae, Notes)
+
+# 2. LASSO Summary for null-inclusive table
+null_comparison_lasso_summary <- lasso_test_metrics %>%
+  select(.metric, .estimate) %>%
+  tidyr::pivot_wider(names_from = .metric, values_from = .estimate) %>%
+  mutate(
+    Model = "LASSO Regression",
+    Notes = "Penalty = 0.1"
+  ) %>%
+  select(Model, RMSE = rmse, R2 = rsq, MAE = mae, Notes)
+
+# 3. Null model summary for table
+null_comparison_null_summary <- null_summary %>%
+  mutate(
+    Model = "Null Model",
+    R2 = NA_real_,
+    Notes = "Predicts mean of training set"
+  ) %>%
+  select(Model, RMSE, R2, MAE, Notes)
+
+# 4. Combine rows into single summary
+null_comparison_summary_all <- bind_rows(
+  null_comparison_rf_summary,
+  null_comparison_lasso_summary,
+  null_comparison_null_summary
+)
+
+# 5. Create gt table
+null_comparison_gt_table <- null_comparison_summary_all %>%
+  mutate(
+    RMSE = sprintf("%.2f", RMSE),
+    MAE  = sprintf("%.2f", MAE),
+    R2   = ifelse(is.na(R2), "–", sprintf("%.2f", R2))
+  ) %>%
+  select(Model, RMSE, MAE, R2, Notes) %>%
+  gt() %>%
+  tab_header(
+    title = md("**Model Performance Summary**"),
+    subtitle = "Held-out test set evaluation for Random Forest, LASSO, and Null Model"
+  ) %>%
+  cols_label(
+    RMSE = "RMSE",
+    MAE  = "MAE",
+    R2   = "R²",
+    Notes = "Model Details"
+  ) %>%
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_body(columns = "Model")
+  ) %>%
+  tab_options(
+    table.font.size = "14",
+    data_row.padding = px(3),
+    table.width = pct(65),
+    column_labels.font.weight = "bold"
+  )
+
+# 6. Save outputs
+saveRDS(null_comparison_gt_table, here("results", "output", "statistical-analysis", "final_model_performance_table.rds"))
+
+gtsave(
+  data = null_comparison_gt_table,
+  filename = here("results", "tables", "final_model_performance_table.png"),
+  expand = 3
+)
+
+# 7. Save raw data
+saveRDS(null_comparison_summary_all, here("results", "output", "statistical-analysis", "model_performance_with_null_numeric.rds"))
+
+# View table
+null_comparison_gt_table
+
+```
+
+4.4.3 Null Model Comparison To contextualize model performance, a null model was included that predicted the mean training set mortality rate for all test observations.
+This baseline yielded substantially higher error metrics, with an RMSE of 228.65 and a mean absolute error (MAE) of 192.28.
+As expected, R² was undefined for the null model due to its constant prediction, which results in zero variance.
+In contrast, both the Random Forest (RMSE = 115.30; R² = 0.75) and LASSO regression (RMSE = 119.54; R² = 0.73) models significantly outperformed the null model across all metrics.
+This confirms that even the more constrained linear model (LASSO) captured meaningful structure in the data beyond what could be explained by the overall average alone.
+
+Aim 4 Summary: We compared two predictive models for county-level hypertension-related CVD mortality: a random forest model and a LASSO regression model.
 The random forest achieved the best performance, with an RMSE of 115.3 and R² of 0.748 on the test set.

### 2025-05-02 code/eda-code/eda.qmd
Changes for final product

@@
 
+This should be the second file to run when reproducing the full analysis for the project.
+
 # Setup
@@
 #load needed packages. make sure they are installed.
+library(tidyverse) #for data processing/cleaning
+library(dplyr) #for data processing/cleaning
 library(tidyr) #for data processing/cleaning
@@
 library(gt) #For a tidy table
+library(stringr)
 
@@
 
-saveRDS(revised_mortalities_histogram, file = here::here("data","processed-data", "revised_mortalities_histogram.rds"))
-revised_mortalities_histogram <- readRDS(here::here("data","processed-data", "revised_mortalities_histogram.rds"))
+invisible(saveRDS(revised_mortalities_histogram, file = here::here("results","output", "eda", "revised_mortalities_histogram.rds")))
+revised_mortalities_histogram <- readRDS(here::here("results","output", "eda", "revised_mortalities_histogram.rds"))
 # Print the mortalities histogram
@@
 
-# Optional: Save these for later
+# Save these for later
 saveRDS(age_group_1, here::here("data", "processed-data", "age_group_1.rds"))
@@
 # Save new histo variable
-saveRDS(hist_age_group_1, here::here("data", "processed-data", "hist_age_group_1.rds"))
+invisible(saveRDS(hist_age_group_1, here::here("results", "output", "eda", "hist_age_group_1.rds")))
 #Print histogram
@@
 # Save and optionally reload
-saveRDS(hist_age_group_2, here::here("data", "processed-data", "hist_age_group_2.rds"))
+invisible(saveRDS(hist_age_group_2, here::here("results", "output", "eda", "hist_age_group_2.rds")))
 # Print histogram 2
@@
 
-saveRDS(boxplot_race_sex, file = here::here("data","processed-data", "boxplot_race_sex.rds"))
+invisible(saveRDS(boxplot_race_sex, file = here::here("results", "output", "eda", "boxplot_race_sex.rds")))
 print(boxplot_race_sex)
 
+```
+
+Summary Statistics of mean and median:
+
+```{r}
+# Summarize mean and median
+mortality_summary <- county_race %>%
+  group_by(race_ethnicity) %>%
+  summarise(
+    Mean = round(mean(mortalities, na.rm = TRUE), 1),
+    Median = round(median(mortalities, na.rm = TRUE), 1),
+    .groups = "drop"
+  ) %>%
+  rename(`Race/Ethnicity` = race_ethnicity)
+
+# Create gt table
+mortality_summary_gt <- mortality_summary %>%
+  gt() %>%tab_header(
+    title = "County-Level Mortality by Race/Ethnicity (2000–2019)",
+    subtitle = "Mean and median mortality rates per 100,000 population") %>%
+  cols_label(Mean = html("Mean Mortality<br>(per 100,000)"),
+    Median = html("Median Mortality<br>(per 100,000)")) %>%
+  cols_align(align = "left", columns = `Race/Ethnicity`) %>%
+  cols_align(align = "right", columns = c(Mean, Median)) %>%
+  fmt_number(columns = c(Mean, Median), decimals = 1) %>%
+  opt_table_font(font = google_font("Roboto")) %>%
+  tab_options(table.width = pct(100), column_labels.font.weight = "bold",
+    data_row.padding = px(3),heading.align = "center")
+
+# Save table
+gtsave(
+  mortality_summary_gt,
+  filename = here::here("results", "tables", "mortality_summary_by_race.png"))
+
+mortality_summary_gt
+```
+
+Boxpplot by sex:
+```{r}
 boxplot_sex <- ggplot(county_sex, aes(x = sex, y = mortalities)) +
-  geom_boxplot() +
-  theme_minimal() +
-  labs(title = "Mortality Rates by Sex", x = "Sex", y = "Mortality Rate")
+  geom_boxplot(fill = "#4C78A8", alpha = 0.7, outlier.size = 0.7, outlier.alpha = 0.5) +
+  labs(
+    title = "County-Level Mortality Rates by Sex",
+    x = "Sex",
+    y = "Mortality per 100,000"
+  ) +
+  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
+  theme_minimal(base_size = 12) +
+  theme(
+    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
+    axis.text.x = element_text(size = 10, hjust = 0.5))
 
-saveRDS(boxplot_sex, file = here::here("data","processed-data", "boxplot_sex.rds"))
-print(boxplot_sex)
+boxplot_sex
+
+# Save file as png
+ggsave(here::here("results", "figures", "mortality_by_sex.png"), plot = boxplot_sex,
+  width = 8, height = 6, dpi = 300)
 
@@
 Neatening up the boxplot for race without sex
+
 ```{r}
-# Relationship between mortality and race
-race_plot <- ggplot(county_race, aes(x = race_ethnicity, y = mortalities)) +
-  geom_boxplot() +
-  theme_minimal() +
-  labs(title = "Mortality Rates by Race/Ethnicity", x = "Race/Ethnicity", y = "Mortality Rate") +
-  theme(axis.text.x = element_text(size = 10)) +
-  scale_x_discrete(labels = function(x) stringr::str_wrap(x, width = 10))
+# Plot: Mortality by race (boxplot)
+mortality_by_race_plot <- ggplot(county_race, aes(x = race_ethnicity, y = mortalities)) +
+  geom_boxplot(fill = "#4C78A8", alpha = 0.7, outlier.size = 0.7, outlier.alpha = 0.5) +
+  stat_summary(fun = "mean", geom = "point", shape = 20, size = 2.8, color = "red") +
+  labs(
+    title = "County-Level Mortality Rates by Race/Ethnicity",
+    x = "Race/Ethnicity",
+    y = "Mortality per 100,000"
+  ) +
+  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
+  theme_minimal(base_size = 12) +
+  theme(
+    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
+    axis.title.x = element_text(size = 14),     
+    axis.title.y = element_text(size = 14),     
+    axis.text.x = element_text(size = 12, hjust = 0.5),
+    axis.text.y  = element_text(size = 12),)
 
-ggsave(here::here("results", "figures", "mortality_by_race.png"), plot = race_plot)
-print(race_plot)
+mortality_by_race_plot
+
+# Save file as png
+ggsave(here::here("results", "figures", "mortality_by_race.png"),
+  plot = mortality_by_race_plot, width = 8, height = 6, dpi = 300)
 
 ```
-This is the same but without sex and the colours for a neater output.
+
+This is the same but without sex and the colours for a neater output and inclusion of a dot to represent the mean values.
+
+Boxplot of county-level mortality rates by race/ethnicity in the U.S. (2000–2019).
+Red dots represent mean mortality. Although White adults had the highest median, Black adults experienced the highest mean mortality, reflecting greater spread and right-skew in their distribution.
 
@@
 
-saveRDS(national_age_trend, here::here("data", "processed-data", "national_trend_by_age.rds"))
+invisible(saveRDS(national_age_trend, here::here("results", "output", "eda", "national_trend_by_age.rds")))
 print(national_age_trend)
@@
 
-saveRDS(race_trend_65_plus, here::here("data", "processed-data", "race_trend_65_plus.rds"))
+invisible(saveRDS(race_trend_65_plus, here::here("results", "output", "eda", "race_trend_65_plus.rds")))
 print(race_trend_65_plus)
@@
 
-saveRDS(race_trend_35_64, here::here("data", "processed-data", "race_trend_35_64.rds"))
+invisible(saveRDS(race_trend_35_64, here::here("results", "output", "eda", "race_trend_35_64.rds")))
 print(race_trend_35_64)
@@
 
-saveRDS(sex_trend_65_plus, here::here("data", "processed-data", "sex_trend_65_plus.rds"))
+invisible(saveRDS(sex_trend_65_plus, here::here("results", "output", "eda", "sex_trend_65_plus.rds")))
 print(sex_trend_65_plus)
@@
 
-saveRDS(sex_trend_35_64, here::here("data", "processed-data", "sex_trend_35_64.rds"))
+invisible(saveRDS(sex_trend_35_64, here::here("results", "output", "eda", "sex_trend_35_64.rds")))
 print(sex_trend_35_64)
@@
 
-gtsave(grouped_summary_clean, here::here("results", "tables", "summary_mortality_table.png"))
+invisible(gtsave(grouped_summary_clean, here::here("results", "tables", "summary_mortality_table.png")))
 
@@
 
-saveRDS(yearly_means, here::here("data", "processed-data", "yearly_means.rds"))
+invisible(saveRDS(yearly_means, here::here("results", "output", "eda", "yearly_means.rds")))
 ```

### 2025-05-02 code/processing-code/processingfile-v1.qmd
Changes for final product

@@
 ---
+
+This should be the first file to run when reproducing the full analysis.
 
@@
 kable(skim_summary_hypertens_filtered)
-saveRDS(skim_summary_hypertens_filtered, file = here::here("data","processed-data", "skim_summary_hypertens_filtered.rds"))
+invisible(saveRDS(skim_summary_hypertens_filtered, file = here::here("results", "output", "processing", "skim_summary_hypertens_filtered.rds")))
 print(skim_summary_hypertens_filtered)
@@
 mortalities_histogram <- hist(hypertens_adult_filtered$mortalities)
-saveRDS(mortalities_histogram, file = here::here("data","processed-data", "mortalities_histogram.rds"))
-mortalities_histogram <- readRDS(here::here("data","processed-data", "mortalities_histogram.rds"))
+invisible(saveRDS(mortalities_histogram, file = here::here("results","output", "processing", "mortalities_histogram.rds")))
+mortalities_histogram <- readRDS(here::here("results", "output","processing", "mortalities_histogram.rds"))
 # Print the mortalities histogram

### 2025-05-02 products/manuscript/Manuscript.qmd
Changes for final product

@@
     highlight-style: github
-bibliography: ../../assets/dataanalysis-references.bib
+bibliography: ../../assets/references/CVD-project-references.bib
 csl: ../../assets/american-journal-of-epidemiology.csl
@@
 
-```{r, echo=FALSE, message=FALSE}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # load a few R packages
@@
 
--   First Author: Shaun van den Hurk$^{1,2,*}$
+-   First Author: Shaun van den Hurk$^{1*}$
 
@@
     USA.
-
-$*$ These authors contributed equally to this work.
 
@@
 adults aged 35 and older using county-level data from 2000 to 2019.
-We explored temporal, demographic, and geographic disparities and evaluated the predictive performance of demographic and temporal factors using machine learning models.
-Descriptive analyses revealed persistent and pronounced disparities in mortality across race, sex, and age groups.
-Black and American Indian/Alaska Native populations experienced the highest mortality rates, particularly among adults aged 65 and older.
-Temporal modeling showed diverging trends across subgroups, with Black adults seeing the steepest declines, while mortality increased among males and American Indian/Alaska Native populations.
-Spatial clustering analyses identified modest but statistically significant autocorrelation, with localized high-mortality clusters in the Southeastern U.S.
-Finally, predictive modeling using Random Forest and LASSO regression demonstrated that demographic and temporal variables could explain a substantial proportion of county-level variation in mortality.
-The Random Forest model achieved an R² of 0.748 and identified race/ethnicity and year as the most influential predictors.
-These findings highlight the enduring influence of demographic and geographic factors on cardiovascular health outcomes and support the use of predictive models to inform future public health strategies.
+We examined temporal, demographic, and geographic disparities to assess both descriptive trends and predictive capacity.
+
+Exploratory analyses revealed persistent disparities in mortality across race, sex, and age groups.
+Black and American Indian/Alaska Native adults consistently exhibited the highest mortality rates, especially among those aged 65 and older.
+Temporal models showed diverging trends across subgroups, with steep declines in Black adults and rising mortality among males and American Indian/Alaska Native populations.
+Spatial clustering was modest but statistically significant, with localized high-mortality areas concentrated in the Southeast.
+
+Predictive modeling using Random Forest and LASSO regression demonstrated that demographic and temporal variables explained a substantial proportion of county-level mortality variation.
+The Random Forest model achieved an R² of 0.748, with age group emerging as the most influential predictor, followed by race/ethnicity.
+These findings underscore the enduring impact of demographic and geographic factors on cardiovascular health outcomes and support the use of predictive models to inform targeted public health interventions.
 
@@
 ischemic heart disease. Cardiovascular diseases were cited as the
-leading cause of death globally in 2022 (..1..), in the USA they were
-responsible for one in every five deaths nationally (...2.).
+leading cause of death globally in 2022, in the USA they were
+responsible for one in every five deaths nationally [@mills_global_2020; @virani_heart_2020; @vaughan_countylevel_2022].
 Hypertension is commonly identified as a significant risk factor for
-multiple cardiovascular conditions, and one that is modifiable, and thus
-represents a good metric for potential public health interventions
-(...3.). However, not all demographics appear to be equally affected by
+multiple cardiovascular conditions, and one that is modifiable, making it an ideal target for public health interventions [@whelton_2017_2018; @mahmood_framingham_2014].
+However, not all demographics appear to be equally affected by
 hypertension and CVD, with differences in rates of CVD and hypertension
 in people from different races, sexes, age groups, geographic location,
-income levels, and other groups in previous studies (..4..). A deeper
-understanding of these factors helps us to get a deeper understanding of
-the multifactorial nature of CVD and also helps guide effective public
-health interventions. Additionally, it is important to understand if
+income levels, and other groups in previous studies [@mahmood_framingham_2014; @yusuf_modifiable_2020].
+
+A deeper understanding of these factors helps us to address the multifactorial nature of CVD and also helps guide effective public health interventions. 
+Additionally, it is important to understand if
 there have been any differences in these rates of CVD and hypertension
@@
 The dataset was collected from the CDC website and the information was
-sourced from the National Vital Statistics System (..5...). The dataset
+sourced from the National Vital Statistics System [@noauthor_rates_nodate]. The dataset
 contains rates and trends in the hypertension-related cardiovascular
@@
 years and older), race/ethnicity, and sex.
-
-(1= )
-
-(2= Virani SS, et al. Heart Disease and Stroke Statistics—2020 Update: A
-Report From the American Heart Association. *Circulation*.
-2020;141:e139–e596.)
-
-(3= Whelton PK, et al. 2017 ACC/AHA Guideline for the Prevention,
-Detection, Evaluation, and Management of High Blood Pressure in Adults.
-*Hypertension*. 2018;71:e13–e115)
-
-(Mahmood SS, et al. The Framingham Heart Study and the epidemiology of
-cardiovascular disease: a historical perspective. *Lancet*.
-2014;383(9921):999–1008.)
-
-(4= Yusuf S, et al. Modifiable risk factors, cardiovascular disease, and
-mortality in 155,722 individuals from 21 high-income, middle-income, and
-low-income countries (PURE): a prospective cohort study. *Lancet*.
-2020;395(10226):795–808.)
-
-(undefined
-
-(5= at the following url:
-https://data.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Hypertension-related-Cardiovas/uc9k-vc2j/about_data.
-)
 
@@
 
-\- Temporal trends differ significantly across race and sex, with
-widening inequities over time.
+\- Temporal trends differ across race and sex, with widening inequities over time.
 
-\- Predictive models can identify key demographic and spatial factors
-associated with higher mortality trends.
+\- Predictive models can identify key demographic and spatial features associated with elevated mortality risk.
 
@@
 
-We obtained the dataset from the CDC’s online portal, sourced from the National Vital Statistics System [5].
+We obtained the dataset from the CDC’s online portal, sourced from the National Vital Statistics System [@noauthor_rates_nodate].
 The raw CSV file was approximately 300MB in size and contained 1,103,872 rows and 24 columns.
-Initial preprocessing was conducted in R to remove redundant variables, reduce file size, and prepare the dataset for analysis.There were some initial processing steps performed to evaluate and delete some of the unnecessary columns and to save this data to an RDS
-document to reduce the size. This was all outlined in the "processingcode" document (and README file), which will not run
+Initial preprocessing was conducted in R to remove redundant variables, reduce file size, and prepare the dataset for analysis.
+This was all outlined in the "processingcode" document (and README file), which will not run
 automatically because the Excel (csv) file will not be available due to
-size concerns with Github uploads (available on request). This process is fully documented in the processingcode.R and processingfile-v1.qmd files within the project repository.
+size concerns with uploading documents (available on request). Full preprocessing steps are documented in the repository’s processingfile-v1.qmd and processingcode.R scripts.
 
@@
 The resulting cleaned dataset contained 572,572 rows and 14 variables.
-Summary statistics and exploratory inspection (e.g., using the skimr package) confirmed the data's readiness for analysis.
-The dataset included both numeric variables (e.g., mortality rates, year, coordinates) and categorical variables (e.g., sex, race/ethnicity, age group, county).
+Summary statistics and exploratory inspection (e.g.
+using the skimr package) confirmed the data's readiness for analysis.
+The dataset included both numeric variables (e.g.
+mortality rates, year, coordinates) and categorical variables (e.g.
+sex, race/ethnicity, age group, county).
 
@@
 The cleaned dataset was saved in .rds format for improved performance and integration with downstream analysis scripts.
-
-All cleaning and preprocessing steps are reproducible and are logged in the repository.
 
@@
 Local clustering patterns were evaluated using Local Indicators of Spatial Association (LISA) via the spdep package.
-Local spatial relationships were visualized using standardized mortality z-scores and classified into high–high, low–low, or spatial outlier clusters based on the combination of mortality values and neighboring spatial lags.
+We visualized local spatial relationships using standardized mortality z-scores, classifying counties as high–high, low–low, or spatial outliers depending on both their mortality values and those of neighboring counties.
 
-To evaluate demographic disparities in temporal trends (Aim 3), we extended the GLMs used in Aim 1 by including three-way interactions between year, race/ethnicity, and sex, and stratified analyses by age group.
+To evaluate demographic disparities in temporal trends (Aim 3), we extended the GLMs used in Aim 1 by including three-way interaction terms between year, race/ethnicity, and sex, and stratified analyses by age group.
 Estimated marginal trends were obtained separately for adults aged 35–64 and those aged 65 and older.
 Annual slope estimates and their confidence intervals were extracted and compared across race–sex combinations.
-Pairwise comparisons of slopes were again conducted using emmeans, with the goal of quantifying and visualizing disparities in mortality trends across key demographic intersections.
+Pairwise slope comparisons were again conducted using emmeans, allowing us to quantify and visualize disparities across key demographic intersections.
 
@@
 The modeling dataset included year, age group, race/ethnicity, and sex as predictors.
-Categorical variables were dummy-encoded using the recipes package, and numeric predictors were normalized; zero-variance predictors were removed prior to modeling.
+Categorical variables were dummy-encoded using the recipes package, and numeric predictors were normalized; zero-variance predictors were removed.
 Data were split into training (80%) and test (20%) sets, stratified by mortality outcome.
@@
 Final model performance was evaluated on the test set using RMSE (root mean squared error), MAE (mean absolute error), and R² (coefficient of determination).
-Confidence intervals for these metrics were not computed due to computational constraints.
+Confidence intervals were computed for cross-validation metrics but not for the final test set performance due to computational constraints.
 Model interpretability was further assessed using variable importance plots for the random forest model and non-zero coefficient selection in LASSO.
@@
 We began by examining overall patterns in hypertension-related cardiovascular disease (CVD) mortality among U.S.
-adults aged 35 and older from 2000 to 2019 (Aim 1; Hypothesis 1).
+adults aged 35 and older from 2000 to 2019.
 Summary statistics revealed large disparities in mortality rates across racial/ethnic and age groups.
@@
 In the 65+ age group, Black adults again had the highest mean mortality (662.6), with American Indian/Alaska Native (451.9) and White (398.3) also experiencing elevated rates.
+To further characterize these disparities, we summarized the mean and median mortality rates by race/ethnicity across all counties and years.
+Black adults (all ages) had the highest mean mortality (305 per 100,000), while White adults (all ages) had the highest median mortality (185 per 100,000), reflecting differing distributions and the influence of extreme values (see Supplementary Table S1).
+These differences are visualized in boxplots by race/ethnicity and sex (@fig-mortality-rate-by-race and @fig-mortality-rate-by-sex).
 
-Boxplots of mortality by race/ethnicity and sex (Figures below) confirmed these disparities, showing right-skewed distributions with particularly high upper ranges for Black and American Indian/Alaska Native adults.
-Males had consistently higher mortality distributions than females across all groups.
+Boxplots confirmed these disparities, showing right-skewed distributions with particularly high upper ranges for Black and American Indian/Alaska Native adults.
+Males consistently had higher mortality distributions than females across all groups.
+While some groups showed declines in mortality over time, others experienced stable or rising trends, supporting the need for formal modeling to assess temporal and demographic patterns
 
-Temporal visualizations indicated divergent patterns between subgroups.
+Temporal visualizations indicated divergent patterns between subgroups in @fig-glm-interaction.
 For example, while some populations experienced declines in mortality rates, others showed upward or stable trends.
@@
 ```{r, echo=FALSE, message=FALSE}
-mortality_race_boxplot <- image_read(here::here("products", "manuscript", "supplement", "mortality_by_race.png"))
-print(mortality_race_boxplot)
+#| label: fig-mortality-rate-by-race
+#| fig-cap: "County-Level Mortality Rates by Race/Ethnicity (2000–2019). Boxplot shows distribution across counties and years."
+#| echo: FALSE
+knitr::include_graphics(here::here("results", "figures", "mortality_by_race.png"))
 
-mortality_sex_boxplot <- image_read(here::here("products", "manuscript", "supplement", "mortality_by_sex.png"))
-print(mortality_sex_boxplot)
+```
+
+```{r, echo=FALSE, message=FALSE}
+#| label: fig-mortality-rate-by-sex
+#| fig-cap: "County-Level Mortality Rates by Sex (2000–2019). Boxplot shows distribution across counties and years."
+#| echo: FALSE
+knitr::include_graphics(here::here("results", "figures", "mortality_by_sex.png"))
 ```
@@
 At the national level, a linear regression of overall mortality against year showed a small but statistically significant increase in mortality from 2000 to 2019 (slope = 1.40, p < 0.001).
-However, stratified models revealed considerable variation between groups.
+However, stratified models revealed considerable variation between groups which may otherwise be masked by this overall increase.
 
-A generalized linear model (GLM) incorporating race/ethnicity and sex predictors (Aim 1.2; Hypotheses 1 and 3) confirmed that Black individuals had the highest predicted mortality rates, followed by White and American Indian/Alaska Native populations.
-Males had significantly higher predicted rates than females across all racial/ethnic groups.
+To better capture these differences, we fit a generalized linear model (GLM) including an interaction between year and race/ethnicity.
+The model revealed substantial divergence in mortality trends across groups.
+Black adults had the highest predicted mortality in 2000 but experienced a steep decline over time (slope = –7.06, p < 0.001), while mortality rates increased among American Indian/Alaska Native adults.
+White adults showed a modest rise in mortality over the twenty year period.
+These patterns and contrasts were not evident in a simpler main-effects-only model, which is presented in Supplementary Figure S1 for comparison.
 
-An interaction model between year and race (Aim 1.3) revealed statistically significant divergence in trends.
-Mortality among Black individuals declined sharply over time (slope = –7.06, p < 0.001), while American Indian/Alaska Native (slope = –3.41) and Hispanic populations showed shallower or increasing trends.
-
-Interaction between year and age group (Aim 1.4) demonstrated that adults aged 65+ experienced a significantly steeper increase in mortality over time compared to those aged 35–64 (interaction slope = +0.22, p < 0.001).
-Estimated marginal trends (Aim 1.5) further confirmed this: the slope among older adults was 1.26 per year (95% CI: 1.18–1.34), versus 1.04 (95% CI: 0.96–1.12) in the younger group (p = 0.0002).
+@fig-glm-interaction shows observed county-level mean mortality rates overlaid with GLM-predicted trends by group.
 
 ```{r, echo=FALSE, message=FALSE}
-#Print the glm plot
-glm_interaction_plot <- image_read(here::here("products", "manuscript", "supplement", "glm_interaction_plot.png"))
-print(glm_interaction_plot)
+#| label: fig-glm-interaction
+#| fig-cap: "Observed and model-predicted mortality rates by race/ethnicity (2000–2019), based on a generalized linear model with interaction between year and race/ethnicity.
+Dots represent annual observed means; lines show GLM predictions adjusting for sex."
+#| echo: FALSE
+knitr::include_graphics(here::here("results", "figures", "glm_interaction_plot.png"))
 ```
 
-### Geographic Distribution and Clustering 
-Spatial visualizations demonstrated considerable geographic variation in mortality by race/ethnicity (Aim 2.1–2.2).
-Elevated mortality was most prominent among Black individuals, particularly in counties across the Southeast and Midwest.
-Asian and Pacific Islander populations displayed lower and more geographically diffuse mortality rates.
+Building on the race/ethnicity model, we next assessed differences in mortality trends by age group using an interaction between year and age.
+This analysis revealed that adults aged 65 and older experienced a significantly steeper increase in mortality over time than those aged 35–64 (interaction slope = +0.22, p < 0.001).
+Estimated marginal trends from the model further supported this finding: the slope for older adults was 1.26 deaths per 100,000 per year (95% CI: 1.18–1.34), compared to 1.04 (95% CI: 0.96–1.12) among younger adults, a difference that was statistically significant (p = 0.0002).
+These results are visualized in Figure @fig-slope-agegroup, which presents the estimated annual mortality increase for each age group with 95% confidence intervals.
 
 ```{r, echo=FALSE, message=FALSE}
-combined_spatial_plot <- image_read(here::here("products", "manuscript", "supplement", "combined_spatial_plot.png"))
-print(combined_spatial_plot)
+#| label: fig-slope-agegroup
+#| fig-cap: "Estimated annual increase in hypertension-related mortality by age group.
+Slope values represent the yearly increase in predicted mortality rates per 100,000, based on a GLM with interaction between year and age group.
+Error bars reflect 95% confidence intervals."
+#| echo: FALSE
+knitr::include_graphics(here::here("results", "figures", "slope_plot_age_group.png"))
 ```
 
-Global spatial clustering was modest but statistically significant for the national dataset (Moran’s I = 0.0435, p < 0.001), with similar low levels of spatial autocorrelation in state-level analyses: Georgia (I = 0.0297), Mississippi (0.0289), Texas (0.0168), and Kentucky (0.0227), all p < 0.001 (Aim 2.3).
+A corresponding figure from the GLM without interaction terms is available in the Supplement (Supplementary Figure S1).
+It assumes parallel trends across groups and does not capture the observed divergence over time.
 
-Local spatial clustering (Aim 2.4) was visualized using Local Indicators of Spatial Association (LISA).
-Statistically significant high–high clusters were identified in counties across parts of the Southeast, but the majority of counties showed no local spatial clustering.
-This suggests that while some mortality hotspots exist, broad demographic and structural drivers likely contribute more to the observed geographic distribution than tightly bounded spatial clusters.
+### Geographic Distribution and Clustering 
+@fig-spatial-all, @fig-spatial-black, and @fig-spatial-asian show spatial variation in hypertension-related mortality across all U.S.
+counties, both overall and for selected racial/ethnic groups.
+Elevated mortality was most prominent among Black individuals, particularly in counties across the Southeast and Midwest.
+Asian and Pacific Islander populations displayed lower and more geographically diffuse mortality patterns.
+Additional subgroup maps are included in the Supplement (Figures S2 - S6), and the combined spatial overview plot (Supplementary Figure S3) provides a side-by-side comparison of geographic patterns by group.
+
+```{r, echo=FALSE, message=FALSE}
+#| label: fig-spatial-all
+#| fig-cap: "Spatial distribution of hypertension-related mortality – All races."
+#| fig-width: 6.5
+#| fig-height: 4.5
+#| echo: FALSE
+knitr::include_graphics(here::here("results", "figures", "spatial_plot_all.png"))
+```
+
+```{r, echo=FALSE, message=FALSE}
+#| label: fig-spatial-black
+#| fig-cap: "Spatial distribution of hypertension-related mortality – Black population."
+#| fig-width: 6.5
+#| fig-height: 4.5
+#| echo: FALSE
+knitr::include_graphics(here::here("results", "figures", "black_spatial_plot.png"))
+```
+
+```{r, echo=FALSE, message=FALSE}
+#| label: fig-spatial-asian
+#| fig-cap: "Spatial distribution of hypertension-related mortality – Asian and Pacific Islander population."
+#| fig-width: 6.5
+#| fig-height: 4.5
+#| echo: FALSE
+knitr::include_graphics(here::here("results", "figures", "asian_spatial_plot.png"))
+```
+
+Global spatial clustering of hypertension-related CVD mortality was modest but statistically significant at the national level, with Moran’s I = 0.0435 (p < 0.001).
+Similar low levels of spatial autocorrelation were observed in state-level analyses: Georgia (I = 0.0297), Mississippi (0.0289), Texas (0.0168), and Kentucky (0.0227), all p < 0.001.
+To further explore local patterns, we conducted Local Indicators of Spatial Association (LISA) analyses.
+Supplementary Figure S7 displays the LISA cluster map for all U.S.
+counties, revealing several high–high clusters concentrated in the Southeastern United States, particularly in regions with high Black mortality.
+Supplementary Figures S8 and S9 provide state-level LISA maps for Georgia and Kentucky, which further illustrate the localized nature of spatial clustering.
+Although some statistically significant clusters were observed, the majority of counties did not exhibit strong local spatial dependence.
+These findings suggest that while certain mortality hotspots exist, broader structural and demographic factors likely account for the majority of geographic variation, rather than tightly bounded spatial processes.
 
 ### Demographic Disparities in Temporal Trends
-We assessed demographic disparities in mortality trends across race/ethnicity and sex within age groups (Aim 3.1–3.3; Hypothesis 3).
-Among adults aged 35–64, Black individuals experienced the steepest annual decline in mortality (–3.68), while American Indian/Alaska Native individuals showed the largest increases (+3.38).
-Overall, males experienced steeper increases than females.
+We assessed demographic disparities in mortality trends by analyzing race/ethnicity and sex within each age group using GLMs with Year × Race × Sex interaction terms.
+Among adults aged 35–64, Black individuals experienced the steepest annual decline in mortality (–3.68 per 100,000), while American Indian/Alaska Native individuals showed the greatest increases (+3.38).
+Similar patterns were observed in the 65+ group, though the absolute changes were larger.
+Overall, males experienced steeper annual increases in mortality compared to females.
+Pairwise slope comparisons revealed the greatest disparity between Overall Male and Black Overall groups (+7.00 per year), followed by contrasts involving American Indian/Alaska Native, White, and Asian/Pacific Islander populations.
+These trends are summarized in @tbl-slope-race-sex and visualized in @fig-slope-35-64 and @fig-slope-65-plus.
+These results indicate persistent and widening disparities in hypertension-related mortality trends, particularly among historically marginalized populations, underscoring the need for targeted public health strategies.
 
-Pairwise comparisons showed the greatest differences between Overall Male and Black Overall groups (+7.00 per year), followed by contrasts involving American Indian/Alaska Native, White, and Asian/Pacific Islander groups.
-These results indicate persistent and widening disparities in mortality trends, particularly among marginalized populations.
-
-```{r, echo=FALSE, message=FALSE}
-readRDS(here::here("results", "tables", "slope_table_agegroup_fmt.rds"))
-readRDS(here::here("results", "tables", "slope_table_race_sex_fmt.rds"))
-```
-
-Additional visualizations from Aim 3.4 reinforce these findings:
-
-```{r, echo=FALSE, message=FALSE}
-slope_plot_35_64 <- image_read(here::here("results", "figures", "slope_plot_35_64.png"))
-print(slope_plot_35_64)
-
-slope_plot_65_plus <- image_read(here::here("results", "figures", "slope_plot_65_plus.png"))
-print(slope_plot_65_plus)
+```{r}
+#| label: tbl-slope-agegroup
+#| tbl-cap: "Estimated annual increase in hypertension-related mortality by age group.
+Values are based on a generalized linear model with an interaction between year and age group.
+Slopes represent predicted yearly change in mortality rates per 100,000."
+#| echo: FALSE
+knitr::include_graphics(here::here("results", "tables", "slope_table_agegroup.png"))
 
@@
 
-### Predictive Modeling of Mortality Rates
-
-To evaluate the ability of demographic and temporal factors to predict hypertension-related CVD mortality at the county level, we developed and compared two supervised machine learning models: a Random Forest and a LASSO regression model (Aim 4; Hypothesis 4).
-Both models were trained on 80% of the dataset and evaluated on a held-out 20% test set, with 10-fold cross-validation used during training to optimize hyperparameters and reduce overfitting.
-
-The Random Forest model demonstrated strong predictive accuracy.
-The best-performing model used all four available predictors (year, age group, race/ethnicity, and sex) with a minimum node size of 15.
-On the test set, this model achieved a root mean squared error (RMSE) of 115.3, a mean absolute error (MAE) of 66.9, and an R² of 0.748, indicating that approximately 75% of the variance in county-level mortality was explained by the model.
-Variable importance scores revealed that race/ethnicity and year were the most influential predictors, followed by age group and sex.
-The predicted versus observed plot showed generally good alignment, though some dispersion at the extremes suggested modest underfitting in counties with very high or low mortality.
-
-```{r, echo=FALSE, message=FALSE}
-model_performance_summary_table <- readRDS(here::here("results", "tables", "model_performance_summary.rds"))
-model_performance_summary_table
+```{r}
+#| label: tbl-slope-race-sex
+#| tbl-cap: "Estimated annual mortality trends by race/ethnicity and sex.
+Slopes were derived from a generalized linear model with Year × Race × Sex interaction terms and represent predicted yearly change in mortality per 100,000.
+Confidence intervals are shown in the final column."
+#| echo: FALSE
+knitr::include_graphics(here::here("results", "tables", "slope_table_race_sex.png"))
 
@@
 
+Additional visualizations from Aim 3.4 reinforce these findings (@fig-slope-35-64 and @fig-slope-65-plus):
+
 ```{r, echo=FALSE, message=FALSE}
-rf_pred_plot <- image_read(here::here("results", "figures", "rf_pred_vs_obs.png"))
-print(rf_pred_plot)
+#| label: fig-slope-35-64
+#| fig-cap: "Estimated annual change in hypertension-related mortality for adults aged 35–64, by race and sex.
+Values represent GLM-estimated slopes with 95% confidence intervals.
+'Overall Male' and 'Overall Female' refer to aggregate trends across all racial/ethnic groups within the respective sex category."
+#| echo: false
+knitr::include_graphics(here::here("results", "figures", "slope_plot_35_64.png"))
+```
 
-rf_vip_plot <- image_read(here::here("results", "figures", "rf_variable_importance.png"))
-print(rf_vip_plot)
+```{r}
+#| label: fig-slope-65-plus
+#| fig-cap: "Estimated annual change in hypertension-related mortality for adults aged 65 and older, by race and sex.
+Values represent GLM-estimated slopes with 95% confidence intervals.
+'Overall Male' and 'Overall Female' refer to aggregate trends across all racial/ethnic groups within the respective sex category."
+#| echo: false
+knitr::include_graphics(here::here("results", "figures", "slope_plot_65_plus.png"))
+```
+
+### Predictive Modeling of Mortality Rates 
+To evaluate the predictive value of demographic and temporal factors for hypertension-related CVD mortality at the county level, we developed two supervised machine learning models: a Random Forest and a LASSO regression model.
+Each model was trained on 80% of the data and evaluated on a 20% held-out test set, with 10-fold cross-validation used to optimize hyperparameters and minimize overfitting.
+
+The Random Forest model achieved the highest predictive accuracy.
+The optimal model used all four predictors—year, age group, race/ethnicity, and sex—with a minimum node size of 15 and mtry = 4.
+On the test set, it produced an RMSE of 115.3, MAE of 66.9, and R² of 0.748, explaining approximately 75% of the variance in county-level mortality (see @tbl-model-performance).
+Variable importance scores, calculated using permutation-based methods, highlighted age group (particularly 65+) as the most influential predictor, followed by race/ethnicity.
+In contrast, sex and year had lower relative importance (@fig-rf-vip).
+The predicted versus observed plot (@fig-rf-pred) showed strong alignment, although slight underprediction was observed in counties with extreme mortality rates, indicating potential underfitting at the distribution tails.
+
+```{r}
+#| label: tbl-model-performance
+#| tbl-cap: "Test set performance metrics for Random Forest, LASSO regression, and a null model.
+RMSE = root mean squared error, MAE = mean absolute error, and R² = coefficient of determination.
+The null model predicts the mean mortality rate and serves as a baseline for comparison."
+#| echo: false
+#| #| fig-width: 9
+#| fig-align: center
+knitr::include_graphics(here::here("results", "tables", "final_model_performance_table.png"))
+```
+
+```{r}
+#| label: fig-rf-pred
+#| fig-cap: "Predicted versus observed mortality rates for the Random Forest model on the test set. Dashed line represents perfect prediction. Test set R² = 0.748."
+#| echo: false
+knitr::include_graphics(here::here("results", "figures", "rf_pred_vs_obs.png"))
 
@@
 
-The LASSO regression model provided a more interpretable, though slightly less accurate, alternative.
-On the same test set, the LASSO model yielded an RMSE of 119.5, an MAE of 70.7, and an R² of 0.726.
-This model performed coefficient shrinkage and variable selection, shrinking several dummy-coded predictors toward zero.
-While this feature enhances interpretability and highlights influential variables, it also limits the model’s flexibility to capture complex non-linear patterns.
-The predicted versus observed plot for the LASSO model revealed horizontal banding patterns.
-This reflects the impact of dummy-encoded categorical variables — many counties shared identical predictor combinations, resulting in repeated predictions and horizontal clustering in plots.
-These effects were more pronounced in LASSO than in the Random Forest, which is better able to handle categorical interactions.
-
-```{r, echo=FALSE, message=FALSE}
-lasso_pred_plot <- image_read(here::here("results", "figures", "lasso_pred_vs_obs.png"))
-print(lasso_pred_plot)
+```{r}
+#| label: fig-rf-vip
+#| fig-cap: "Variable importance from the Random Forest model. Importance was calculated using permutation-based methods. Results are shown for the top 20 predictors."
+#| echo: false
+knitr::include_graphics(here::here("results", "figures", "rf_variable_importance.png"))
 
@@
 
-Despite these differences, both models demonstrated solid performance in capturing patterns in the data.
-The Random Forest provided better predictive accuracy, while the LASSO model offered greater clarity in identifying relevant features.
-Taken together, these results indicate that demographic and temporal factors are strong predictors of mortality variation at the county level, supporting Hypothesis 4 and offering tools for future public health risk modeling.
+The LASSO regression model served as a more interpretable, though slightly less accurate, alternative.
+On the test set, it achieved an RMSE of 119.5, MAE of 70.7, and R² of 0.726 (see @tbl-model-performance).
+The model applied coefficient shrinkage and automatic variable selection using a penalty parameter (λ = 1e–4), which was selected via cross-validation to minimize RMSE.
+This regularization excluded several dummy-coded predictors, improving interpretability by focusing on the most informative features.
+However, LASSO’s linear structure and shrinkage approach limited its flexibility in capturing non-linear interactions.
+The predicted versus observed plot (@fig-lasso-pred) displayed clear horizontal banding patterns, a result of many counties sharing identical categorical predictor combinations.
+These banding effects were more pronounced than in the Random Forest model, which better captured complex interactions through hierarchical tree splits.
+Supplementary results confirmed strong generalization in LASSO model performance across folds, with a mean RMSE of 119.63 (95% CI: 117.24–122.02) and R² of 0.73 (95% CI: 0.72–0.73) across 10-fold CV resamples (see Supplementary Table S4).
+
+```{r}
+#| label: fig-lasso-pred
+#| fig-cap: "Predicted versus observed mortality rates for the LASSO regression model on the test set.
+Dashed line represents perfect prediction.
+Horizontal banding reflects repeated predictions due to dummy-encoded categorical variables."
+#| echo: false
+knitr::include_graphics(here::here("results", "figures", "lasso_pred_vs_obs.png"))
+
+```
+
+To contextualize model performance, we also evaluated a null model that predicted the mean training set mortality rate for all test observations.
+As expected, this baseline yielded substantially worse performance: RMSE = 228.65, MAE = 192.28, and an undefined R² due to zero variance in predictions (see Table 3).
+In contrast, both the Random Forest and LASSO models performed significantly better, confirming their ability to capture meaningful structure in the data.
+These results underscore that even with a limited set of demographic and temporal features, machine learning models can explain a substantial portion of the variation in county-level mortality.
+
+To assess generalization, we compared model performance across the training and test sets.
+As shown in Supplementary Table S2, both models yielded nearly identical metrics across splits.
+For example, the Random Forest achieved RMSE = 115.31 (train) vs.
+115.30 (test), and LASSO showed similarly consistent results.
+This close alignment suggests minimal overfitting.
+Cross-validation results further supported model stability.
+The Random Forest model demonstrated mean CV RMSE values ranging from 175.01 to 177.53, with R² values consistently around 0.72–0.73 across hyperparameter settings (see Supplementary Table S3).
+The LASSO model showed equally strong consistency, with a mean RMSE of 119.63 (95% CI: 117.24–122.02) and R² of 0.73 (95% CI: 0.72–0.73) across the 10 CV folds (Supplementary Table S4).
+These findings reinforce that both models are well-calibrated and generalize effectively, capturing robust mortality patterns without overfitting.
+
+Despite their differences, both models demonstrated strong performance in capturing meaningful mortality patterns.
+The Random Forest offered greater predictive accuracy, while the LASSO model provided clearer interpretability through variable selection and shrinkage.
+Together, these results highlight that demographic and temporal variables, particularly age, race/ethnicity, and year, are robust predictors of county-level hypertension-related mortality.
+This supports our hypothesis that predictive models can effectively identify demographic patterns associated with elevated mortality risk.
+As such, these approaches represent promising tools for population-level risk estimation and public health planning.
 
@@
 ## Summary and Interpretation
-
 We examined patterns and predictors of hypertension-related cardiovascular disease (CVD) mortality among U.S.
 adults aged 35 and older between 2000 and 2019.
-Through a combination of exploratory visualization, statistical modeling, spatial clustering, and machine learning, we found consistent evidence of disparities in mortality across time, geography, and demographic subgroups.
-These findings underscore the ongoing burden of hypertension-related CVD in the United States and highlight areas for targeted intervention.
+Through exploratory visualizations, regression modeling, spatial clustering, and machine learning, we identified consistent disparities in mortality across time, geography, and demographic groups.
+These findings highlight the persistent burden of hypertension-related CVD and point to critical areas for targeted intervention.
 
-National trends revealed that overall mortality rates remained relatively stable over the study period.
-However, stratified models showed considerable variation between groups.
-Black individuals experienced the highest mortality rates throughout, while American Indian/Alaska Native populations showed a concerning upward trend in recent years.
-Adults aged 65 and older had significantly steeper increases in mortality rates than younger adults.
-These findings are consistent with previous national surveillance reports that have identified persistent disparities in CVD burden across racial/ethnic and age groups [2,4].
+Nationally, overall mortality rates remained relatively stable across the study period.
+However, stratified analyses revealed substantial variation between subgroups.
+Black individuals consistently experienced the highest mortality rates, while American Indian/Alaska Native populations showed a notable upward trend in recent years.
+Adults aged 65 and older exhibited steeper annual increases than younger adults.
+These patterns align with prior surveillance studies [@vaughan_countylevel_2022; @mahmood_framingham_2014] and underscore the importance of modeling subgroup-specific trends.
+Aggregated models with shared slopes risk masking meaningful divergences—such as the sharp declines among Black adults versus rising mortality in American Indian/Alaska Native populations.
 
-Spatial analyses detected statistically significant but modest spatial clustering of mortality at the national and state levels.
-While global Moran’s I statistics confirmed the presence of spatial autocorrelation, the magnitude was small (I = 0.0435 nationally).
-Local spatial clustering (LISA) identified limited high-mortality hotspots, primarily concentrated in the Southeast.
-These results suggest that broader structural and demographic factors—rather than localized geographic effects—may be the primary drivers of hypertension-related CVD mortality.
-This is in line with previous studies indicating that health disparities are often influenced more by social determinants and systemic inequities than by geographic proximity alone [4,6].
+Spatial analysis revealed modest but statistically significant clustering of mortality at national and state levels.
+Global Moran’s I was low (I = 0.0435), and LISA analyses identified limited high–high clusters, primarily in the Southeast.
+These findings suggest that broad demographic and structural factors, rather than localized spatial effects, may better explain mortality distribution—echoing prior work on the role of social determinants in health disparities [@mahmood_framingham_2014; @yusuf_modifiable_2020].
 
-Demographic subgroup analyses revealed widening inequities in mortality trends.
-Among adults aged 35–64, Black individuals experienced the steepest annual declines, suggesting some gains in this group.
-However, these improvements occurred in parallel with rising mortality among American Indian/Alaska Native individuals and males, who showed the largest annual increases.
-Among adults aged 65 and older, mortality increased for nearly all groups.
-These divergent trends highlight the importance of considering intersecting axes of inequality—such as race, sex, and age—in public health surveillance.
-They also support prior research showing that demographic-specific interventions are necessary to effectively reduce disparities in chronic disease outcomes [3,4,7].
+Demographic subgroup analyses showed diverging mortality trends.
+Among adults aged 35–64, Black individuals experienced the largest declines, while mortality increased sharply among American Indian/Alaska Native adults and males.
+In the 65+ group, mortality rose across nearly all categories.
+These results reinforce the importance of considering intersecting dimensions of inequality, such as race, sex, and age, in surveillance and intervention planning.
+They also support the need for demographic-specific public health responses to reduce disparities in chronic disease outcomes [@whelton_2017_2018; @mahmood_framingham_2014].These stratified insights are further strengthened by the predictive modeling results, which confirmed age, race/ethnicity, and year as the most influential predictors of county-level mortality.
 
@@
 These findings contribute to a growing body of evidence emphasizing the importance of demographic stratification and predictive modeling in public health research.
-In particular, our analysis supports previous work demonstrating the utility of ensemble and regularized regression models in epidemiologic prediction tasks [8,9].
 Importantly, our results reinforce that even well-known risk factors—such as age, sex, race/ethnicity, and time—continue to explain a large proportion of the variance in hypertension-related CVD mortality and may be sufficient for initial risk stratification at the population level.
@@
 ## Strengths and Limitations
-This study benefits from the use of a comprehensive, nationwide dataset covering nearly two decades of mortality records.
-The integration of statistical and machine learning methods allowed us to not only describe mortality patterns but also assess their predictability.
+This study benefits from the use of a comprehensive, nationwide dataset sourced from the CDC’s National Vital Statistics System, covering nearly two decades of hypertension-related cardiovascular disease mortality records.
+The integration of traditional statistical models and machine learning approaches allowed us to describe mortality patterns and assess their predictability using demographic and temporal inputs.
 However, several limitations should be noted.
-First, the analysis was limited to county-level aggregated data, which may obscure within-county heterogeneity and limit causal inference.
-Second, although we used age-standardized mortality rates, residual confounding by comorbidities, access to care, socioeconomic status, or other contextual factors cannot be excluded.
-Lastly, while our models achieved high predictive accuracy, they were not designed to support individual-level predictions or establish causal relationships.
+First, the analysis relied on county-level aggregated data, which may obscure within-county heterogeneity and limits the ability to draw individual-level conclusions.
+Second, although we used age-standardized mortality rates, key contextual variables such as socioeconomic status, healthcare access, or comorbidities were not included, potentially introducing residual confounding.
+Lastly, while our models achieved high predictive accuracy for county-level outcomes, they were not intended to establish causal relationships or support individual-level risk prediction.
 
 ## Implications and Future Directions
-Our findings highlight important and persistent disparities in hypertension-related CVD mortality that should inform ongoing public health surveillance and intervention efforts.
-Geographic clustering, although weak, may still guide resource allocation to high-burden areas.
-Predictive models developed here can support proactive health planning, particularly when integrated with local surveillance systems.
-Future research should explore integrating additional variables—such as healthcare access, environmental exposures, and social determinants of health—to further refine predictions and address upstream drivers of CVD disparities.
+Our findings highlight persistent and inequitable patterns in hypertension-related CVD mortality that should inform public health surveillance, policy development, and resource allocation strategies.
+While geographic clustering was modest, identifying high-burden areas remains important for targeting interventions.
+The predictive models developed here demonstrate that even simple demographic and temporal variables can capture a substantial proportion of county-level variation, making them useful for settings where more granular data are unavailable.
+These models can support proactive health planning and serve as screening tools when integrated with local or regional surveillance systems.
+
+Future research should incorporate additional contextual predictors—such as healthcare access, environmental risk factors, and social determinants of health—to improve model precision and equity relevance.
+Efforts to enhance interpretability and generalizability through multi-level modeling, causal frameworks, or longitudinal datasets may also help inform upstream interventions to reduce CVD disparities.
 
@@
 
-This study provides a comprehensive analysis of hypertension-related cardiovascular disease (CVD) mortality among U.S.
-adults aged 35 and older from 2000 to 2019, using national county-level data.
-Our findings reveal that while overall mortality trends remained relatively stable at the national level, significant disparities persist across demographic and geographic groups.
-Black and American Indian/Alaska Native populations consistently experienced higher mortality rates, with the steepest increases observed among older adults and males.
-Spatial analysis identified weak but statistically significant clustering, particularly in the Southeastern United States, though most variation was driven by broader structural and demographic patterns rather than tightly localized effects.
+This study presents a comprehensive analysis of hypertension-related cardiovascular disease (CVD) mortality among U.S.
+adults aged 35 and older from 2000 to 2019, leveraging national county-level data.
+While national mortality trends remained relatively stable, we identified significant disparities across race, age, sex, and geography.
+Black and American Indian/Alaska Native populations consistently bore higher mortality burdens, with the most pronounced increases observed among older adults and males.
+Spatial analysis revealed modest but statistically significant clustering, particularly in the Southeastern U.S., with broader demographic and structural factors driving most of the observed variation.
 
-Temporal modeling demonstrated that subgroup-specific trends diverged substantially, suggesting that recent gains in cardiovascular health have not been equitably distributed.
-Our predictive modeling further supports this, with demographic features—particularly race/ethnicity and age—emerging as dominant predictors of county-level mortality risk.
-Random Forest models outperformed LASSO regression in accuracy, but both approaches reinforced the critical role of social and demographic context in shaping health outcomes.
+Temporal modeling showed that subgroup-specific mortality trends diverged considerably over time, suggesting that recent public health gains have not been shared equally.
+Our predictive modeling confirmed the strong influence of demographic variables—especially age group, which emerged as the most important predictor, followed by race/ethnicity and year.
+These variables alone explained a substantial portion of the variance in county-level mortality rates.
+While Random Forest models achieved higher predictive accuracy, LASSO models enhanced interpretability, illustrating complementary strengths for public health modeling.
 
-Taken together, these findings underscore the urgent need for targeted public health strategies that account for longstanding inequities in cardiovascular risk.
-Future research should integrate social determinants of health, healthcare access, and policy-level factors to better understand the drivers of these disparities and to design interventions that are both equitable and effective.
-
-Note that this cited reference will show up at the end of the document,
-the reference formatting is determined by the CSL file specified in the
-YAML header. Many more style files for almost any journal [are
-available](https://www.zotero.org/styles). You also specify the location
-of your bibtex reference file in the YAML. You can call your reference
-file anything you like.
+Collectively, these findings reinforce the need for targeted, equity-focused public health strategies.
+Future work should incorporate structural and social determinants of health, including healthcare access and policy context, to better understand and address the root causes of persistent CVD disparities.
+Robust, interpretable models like those presented here can support evidence-based decision-making, guide resource prioritization, and ultimately contribute to more equitable cardiovascular outcomes.
 

### 2025-05-02 products/manuscript/supplement/Supplement-CVD.qmd
Changes for final product

---
title: "Supplement to Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States."
format:
  docx:
    toc: false
    number-sections: False
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

## Supplementary figures

```{r}
#| echo: FALSE
#| results: 'asis'

knitr::include_graphics(here::here("results", "tables", "mortality_summary_by_race.png"))
cat("**Supplementary Table S1.** Mean and median county-level mortality rates per 100,000 by race/ethnicity (2000–2019).\n\n")
```

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'
knitr::include_graphics(here::here("results", "figures", "glm_race_sex_plot.png"))
cat("**Supplementary Figure S1.** Observed and predicted mortality rates by race/ethnicity (2000–2019), based on a generalized linear model without interaction terms. Assumes a shared temporal trend across all groups.\n\n")

```

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'

knitr::include_graphics(here::here("results", "figures", "spatial_plot_all.png"))
cat("**Supplementary Figure S2.** Spatial distribution of hypertension-related mortality – All Races.\n\n")
```

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'
#| fig.width: 12
#| fig.height: 9
knitr::include_graphics(here::here("results", "figures", "combined_spatial_plot.png"))
cat("**Supplementary Figure S3.** Spatial distribution of hypertension-related mortality – Combined plot of all race/ethnicity groups.\n\n")

```

The combined plot (Supplement Figure 3) provides an overview of mortality patterns across groups. While harder to read at the small scale once rendered in Word, it was still included as it offers helpful visual context when reviewed alongside the full-size individual maps.

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'

knitr::include_graphics(here::here("results", "figures", "white_spatial_plot.png"))
cat("**Supplementary Figure S4.** Spatial distribution of hypertension-related mortality – White population.\n\n")
```

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'

knitr::include_graphics(here::here("results", "figures", "hispanic_spatial_plot.png"))
cat("**Supplementary Figure S5.** Spatial distribution of hypertension-related mortality – Hispanic population.\n\n")
```

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'

knitr::include_graphics(here::here("results", "figures", "native_spatial_plot.png"))
cat("**Supplementary Figure S6.** Spatial distribution of hypertension-related mortality – American Indian and Alaska Native population.\n\n")
```

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'

knitr::include_graphics(here::here("results", "figures", "lisa_cluster_type_map.png"))
cat("**Supplementary Figure S7.** Local Indicators of Spatial Association (LISA) cluster map for the contiguous United States. The map identifies county-level clusters of high or low mortality rates relative to their neighbors. Statistically significant clusters include high–high (mortality hotspots), low–low (coldspots), and spatial outliers (high–low or low–high), with color coding based on the spatial relationship and statistical significance.\n\n")
```

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'

knitr::include_graphics(here::here("results", "figures", "ga_lisa_cluster_plot.png"))
cat("**Supplementary Figure S8.** Local Indicators of Spatial Association (LISA) cluster map for Georgia. This map visualizes statistically significant local spatial autocorrelation in hypertension-related mortality rates across counties in Georgia. High–high clusters indicate counties with high mortality rates surrounded by similarly high neighbors, while low–low clusters represent the opposite. Other categories denote spatial outliers and areas with no significant local clustering.\n\n")
```

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'

knitr::include_graphics(here::here("results", "figures", "lisa_cluster_kentucky.png"))
cat("**Supplementary Figure S9.** Local Indicators of Spatial Association (LISA) cluster map for Kentucky. This figure shows localized spatial clusters of hypertension-related mortality, highlighting counties with significantly elevated or reduced mortality compared to their immediate neighbors. High–high clusters represent potential hotspots for targeted public health intervention.\n\n")
```

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'

knitr::include_graphics(here::here("results", "tables", "model_train_test_comparison.png"))
cat("**Supplementary Table S2.** Train and test set performance metrics for Random Forest and LASSO models. Metrics include RMSE, MAE, and R². Similar values across splits indicate strong generalization and minimal overfitting.\n\n")
```

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'

knitr::include_graphics(here::here("results", "tables", "rf_cv_performance_table.png"))
cat("**Supplementary Table S3.** Ten-fold cross-validation performance metrics for the Random Forest model. Values represent mean RMSE and R² across folds with 95% confidence intervals calculated using the standard error for each hyperparameter combination.\n\n")
```

Cross-validation performance metrics for Random Forest model. Mean RMSE and R² values are shown across 10-fold cross-validation for selected hyperparameter combinations (mtry and min_n). Values reflect consistent predictive performance across folds, indicating high model stability.

\`\`\`

```{r}
#| echo: FALSE
#| results: 'asis'

knitr::include_graphics(here::here("results", "tables", "lasso_cv_performance_table.png"))
cat("**Supplementary Table S4.** Ten-fold cross-validation performance metrics for the LASSO regression model. Values represent mean RMSE and R² with 95% confidence intervals based on variability across folds using the selected penalty parameter.\n\n")
```

Cross-validation performance metrics for LASSO regression model. Mean RMSE and R² values with 95% confidence intervals are reported across 10-fold cross-validation folds using the optimal penalty value. The narrow confidence intervals reflect low variability and high consistency in model performance.

\`\`\`

## Reproducibility and Software

### Steps for reproducing the project

To fully reproduce all steps of the project, the code should be run in the following order:

1. processing-code/processing-code.qmd 
2. eda-code/eda.qmd
3. analysis-code/statistical-analysis-v2-slim.qmd* 
4. products/manuscript/Manuscript.qmd
5. products/manuscript/supplement/Supplement-CVD.qmd

*Note: This is a large script and will take some time and processing power to run. The script uses loops to load saved data to reduce computation but expect it to take about 15 minutes to run the full script even with those steps.

Note: Because all data, figures, and tables are saved in the relevant subfolders, the manuscript 
can be produced by simply running the Manuscript.qmd file.

Note: processing-code/processing-code.qmd - This file will not run as the Excel CSV file is not in the repository due to size limitations. It was included purely to demonstrate processing steps. You can contact the author to get it or download the file directly from the CDC website if you wish to run this yourself.

### Software packages

The analysis was conducted using R version 4.4.0 (2024-04-24 ucrt) on a Windows 11 x64 system via RStudio 2024.12.0+467. The Quarto engine, R packages, and their versions used in this analysis are listed below to support full reproducibility.

**R version:**\
R version 4.4.0 (2024-04-24 ucrt)

**Operating system:**\
Windows 11 x64 (build 26100)

**RStudio version:**\
2024.12.0+467 Kousa Dogwood

**Session packages:**\
*(includes attached and namespace-loaded packages)*

-   **tidymodels** 1.3.0
-   **tidyverse** 2.0.0
-   **gt** 0.11.1
-   **yardstick** 1.3.2
-   **rsample** 1.3.0
-   **parsnip** 1.3.0
-   **tune** 1.3.0
-   **recipes** 1.1.1
-   **workflows** 1.2.0
-   **workflowsets** 1.1.0
-   **vip** 0.4.1
-   **spdep** 1.3-10
-   **spatialreg** 1.3-6
-   **sf** 1.0-19
-   **spData** 2.3.4
-   **skimr** 2.1.5
-   **dplyr** 1.1.4
-   **ggplot2** 3.5.1
-   **forcats** 1.0.0
-   **purrr** 1.0.4
-   **readr** 2.1.5
-   **stringr** 1.5.1
-   **tibble** 3.2.1
-   **lubridate** 1.9.3
-   **gtsummary** (if used)
-   **webshot2** 0.1.1
-   **here** 1.0.1
-   **magick** 2.8.5
-   **sessioninfo** 1.2.3
-   Other dependencies (e.g., cli, glue, broom, etc.) were loaded automatically via dependencies of the core packages listed above.

**Note:** A full session output is available upon request. All analyses were conducted using a reproducible pipeline, and Quarto was used for document generation and reporting.

### 2025-05-05 _peer-reviews-temp/project-review-vandenhurk.qmd
updates for reproducibility and readmes

---
title: Project Review Template 
author: Pooja Gokhale
date: date-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States.

Name of project author(s): Shaun van den Hurk

Name of project reviewer: Pooja Gokhale

# Specific project content evaluation
## Background, Context and Motivation

### Feedback and Comments
The background information is comprehensive. There are a few grammatical mistakes and the general writing could be slightly improved.
The aims and hypotheses are well stated.
The references are not cited appropriately - (..1..) - they appear in this way in the manuscript.
Additionally, in the data source, the citations seem to have been added manually. These should be removed.

### Summary assessment
* strong contextualization and motivation

## Question description
### Feedback and Comments
The question and hypotheses are stated clearly. The data source is directly related to the question.

### Summary assessment
* question/hypotheses fully clear

## Data description
### Feedback and Comments
The data source is provided clearly. There is no codebook. It would be helpful to mention all the variables available in the dataset.

### Summary assessment
* source and overall structure of data somewhat explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments
The data cleaning process has been mentioned in adequate detail. 
The steps are reasonable and well explained.
Alternative methods have not been discussed, however the methods used are appropriate.
The exploratory figures have also been given.
The metadata for the figures does not seem necessary and could be removed.

### Summary assessment
* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
### Feedback and Comments
The analysis methods were appropriate. The analysis was done correctly.

### Summary assessment
* strong and reasonable analysis

## Presentation
### Feedback and Comments
The figures and tables are generally good quality. However, the figures have meta-data included.

### Summary assessment
* results are presented ok, with room for improvement

## Discussion/Conclusions
### Feedback and Comments
The discussion is very well written. Reference numbers have been given but there are no references.

### Summary assessment
* minor parts wrong, missing or unclear

## Further comments

Make sure the references are cited correctly. Remove meta-data from the figures.

# Overall project content evaluation

## Structure
### Feedback and Comments
The project is well-structured. The reference file has not been updated. There are a few irrelevant files in the assets folder.

### Summary assessment
* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation 
### Feedback and Comments
The code files are very well documented and explained.

### Summary assessment
* fully and well documented

## Reproducibility
### Feedback and Comments
The code files run without issues. The code works when the code chunks are run, however does not render as an html document. This may be related to issues with computational power and storage.

### Summary assessment
* small parts not reproducible or required manual intervention

## Thoroughness
### Feedback and Comments
The overall study is extremely thorough. All parts have been discussed well. Most of the issues are related to documentation, not the actual analysis.

### Summary assessment
* strong level of thorougness

## Further comments
Here are my suggestions:
1. Add references.
2. Ensure that all code files render properly.
3. Remove meta-data from figures.
4. General writing improvement (minor).

### 2025-05-05 products/manuscript/supplement/Supplement-CVD.qmd
updates for reproducibility and readmes

@@
 
-1. processing-code/processing-code.qmd 
-2. eda-code/eda.qmd
-3. analysis-code/statistical-analysis-v2-slim.qmd* 
-4. products/manuscript/Manuscript.qmd
-5. products/manuscript/supplement/Supplement-CVD.qmd
+1.  processing-code/processing-code.qmd^^ (Not required to reproduce results - This file will not run - see below)
+2.  processing-code/processingfile-v1.qmd
+3.  eda-code/eda.qmd
+4.  analysis-code/statistical-analysis-v2-slim.qmd\*
+5.  products/manuscript/Manuscript.qmd
+6.  products/manuscript/supplement/Supplement-CVD.qmd
 
-*Note: This is a large script and will take some time and processing power to run.
+\*Note: This is a large script and will take some time and processing power to run.
 The script uses loops to load saved data to reduce computation but expect it to take about 15 minutes to run the full script even with those steps.
 
-Note: Because all data, figures, and tables are saved in the relevant subfolders, the manuscript 
-can be produced by simply running the Manuscript.qmd file.
+Note: Because all data, figures, and tables are saved in the relevant subfolders, the manuscript can be produced by simply running the Manuscript.qmd file.
 
-Note: processing-code/processing-code.qmd - This file will not run as the Excel CSV file is not in the repository due to size limitations.
+^^Note: processing-code/processing-code.qmd - This file will not run as the Excel CSV file is not in the repository due to size limitations.
 It was included purely to demonstrate processing steps.
@@
 
-### Software packages
+### Software Packages and Environment
+
+All analyses presented in this project were performed using R within a version-controlled environment managed by the `renv` package.
+This ensures that all package versions and dependencies used to generate figures and tables are locked and reproducible.
+
+To reproduce this analysis:
+
+1.  Clone the repository.
+2.  Open the project in R or RStudio.
+3.  Restore the environment by running:
+
+#renv::restore()
 
@@
 
-**Session packages:**\
-*(includes attached and namespace-loaded packages)*
+**Key R Packages Used (Attached or namespace-loaded)**
 
